---
title: "Oslo Housing Market"
description: "Daily snapshot of for-sale listings in Oslo, AI-classified into 10 housing categories."
date: today
execute:
  freeze: false
  echo: false
  warning: false
  message: false
---

```{r check-data}
# Guard chunk — must be separate from setup so read.csv is never reached when
# the file is missing. knitr::opts_chunk$set(eval=FALSE) disables all
# subsequent chunks reliably in Quarto (knit_exit() is not reliable here).
CSV_PATH <- here::here("finn-housing", "data", "listings_export.csv")
if (!file.exists(CSV_PATH)) {
  cat("> **No data yet.** Run `finn-housing/scraper.R` locally (or wait for the daily workflow) to populate the data.\n")
  knitr::opts_chunk$set(eval = FALSE)
}
```

```{r setup}
library(dplyr)
library(ggplot2)
library(scales)
library(forcats)
library(leaflet)
library(DT)
library(stringr)

# ── Colour palette (shared across all plots) ──────────────────────────────────
CAT_COLORS <- c(
  "Luxury"                        = "#8B0000",
  "Family home"                   = "#2E86AB",
  "Starter / budget"              = "#A8DADC",
  "Investment"                    = "#457B9D",
  "Student / young professional"  = "#F4A261",
  "New development"               = "#2A9D8F",
  "Renovation"                    = "#E9C46A",
  "Central urban"                 = "#264653",
  "Suburban"                      = "#6D6875",
  "Waterfront / premium location" = "#1D3557",
  "Unclassified"                  = "#BBBBBB"
)

# ── Load & clean data ─────────────────────────────────────────────────────────
raw <- read.csv(CSV_PATH, stringsAsFactors = FALSE, fileEncoding = "UTF-8")

df <- raw |>
  mutate(
    # Cast all optional text columns to character — prevents logical-NA vs character
    # type conflicts in coalesce() when columns are entirely NA in the CSV
    across(c(title, address, neighborhood, property_type,
             category, category_confidence, category_reasoning,
             classified_at, url),
           as.character),
    scraped_date = as.Date(substr(scraped_at, 1, 10)),
    price_m      = price / 1e6,
    price_sqm    = ifelse(!is.na(price) & !is.na(size_sqm) & size_sqm > 0,
                          price / size_sqm, NA_real_),
    category     = ifelse(is.na(category) | category == "", "Unclassified", category),
    lat          = suppressWarnings(as.numeric(lat)),
    lon          = suppressWarnings(as.numeric(lon))
  )

classified <- df |> filter(category != "Unclassified")

# ── Value-for-money score ─────────────────────────────────────────────────────
# Definition: how much cheaper is this listing per m² vs the median for its
# category?  score = category_median_price_sqm / listing_price_sqm
# score > 1  →  cheaper than category median  →  good value
# score < 1  →  more expensive than category median  →  premium pricing
#
# A score of 1.40 means you pay 29 % less per m² than the typical listing in
# the same category.
cat_medians <- classified |>
  filter(!is.na(price_sqm), price_sqm > 10000, price_sqm < 250000) |>
  group_by(category) |>
  summarise(cat_med_psqm = median(price_sqm, na.rm = TRUE), .groups = "drop")

classified <- classified |>
  left_join(cat_medians, by = "category") |>
  mutate(
    value_score = ifelse(!is.na(price_sqm) & price_sqm > 0,
                         cat_med_psqm / price_sqm, NA_real_),
    value_pct   = round((value_score - 1) * 100, 1)   # % below(+) / above(-) median
  )

last_scrape  <- max(df$scraped_date, na.rm = TRUE)
total_n      <- nrow(df)
classified_n <- nrow(classified)
med_price    <- median(classified$price, na.rm = TRUE)
med_size     <- median(classified$size_sqm, na.rm = TRUE)
med_psqm     <- median(classified$price_sqm, na.rm = TRUE)
```

Explore **`r format(total_n, big.mark = " ")`** for-sale listings in Oslo, scraped from finn.no and classified by AI into 10 housing segments. Use the charts and map below to find patterns — or scroll to the bottom for this week's standout pick.

---

::: {.callout-note appearance="minimal"}
**`r format(total_n, big.mark = " ")`** listings &nbsp;·&nbsp;
**`r format(classified_n, big.mark = " ")`** classified &nbsp;·&nbsp;
Median price **`r paste0(round(med_price / 1e6, 1), " M NOK")`** &nbsp;·&nbsp;
Median size **`r paste0(round(med_size), " m²")`** &nbsp;·&nbsp;
Median price/m² **`r paste0(format(round(med_psqm / 1000, 0), big.mark = " "), " k NOK")`** &nbsp;·&nbsp;
Last updated **`r format(last_scrape, "%d %b %Y")`**
:::

---

## Oslo on the Map

*Bubble size = asking price. Colour = category. Click a bubble to open the listing.*

```{r map}
map_df <- df |>
  filter(!is.na(lat), !is.na(lon),
         lat > 59.5, lat < 60.2,
         lon > 10.3, lon < 11.2,
         !is.na(price))

if (nrow(map_df) == 0) {
  cat("> Map will appear once geocoding has run (happens automatically after each classify step).\n")
} else {
  # Radius: log-scale of price, capped to avoid giant bubbles
  map_df <- map_df |>
    mutate(
      radius   = pmin(pmax(log(price / 1e6) * 5, 4), 22),
      col_hex  = CAT_COLORS[category],
      col_hex  = ifelse(is.na(col_hex), "#BBBBBB", col_hex),
      popup_txt = paste0(
        "<b><a href='", url, "' target='_blank'>", title, "</a></b><br>",
        "<b>", ifelse(!is.na(price), paste0(format(price, big.mark = " "), " NOK"), "—"), "</b>",
        ifelse(!is.na(size_sqm), paste0(" · ", size_sqm, " m²"), ""),
        ifelse(!is.na(rooms),    paste0(" · ", rooms, " rom"), ""), "<br>",
        ifelse(!is.na(address),  paste0(address, "<br>"), ""),
        "<span style='background:", col_hex, ";color:#fff;padding:2px 6px;border-radius:3px;font-size:0.8em'>",
        category, "</span>"
      )
    )

  leaflet(map_df) |>
    addProviderTiles(providers$CartoDB.Positron) |>
    setView(lng = 10.757, lat = 59.913, zoom = 11) |>
    addCircleMarkers(
      lng        = ~lon,
      lat        = ~lat,
      radius     = ~radius,
      color      = ~col_hex,
      fillColor  = ~col_hex,
      fillOpacity = 0.65,
      weight     = 1,
      opacity    = 0.9,
      popup      = ~popup_txt
    ) |>
    addLegend(
      position = "bottomright",
      colors   = CAT_COLORS[names(CAT_COLORS) %in% unique(map_df$category)],
      labels   = names(CAT_COLORS)[names(CAT_COLORS) %in% unique(map_df$category)],
      title    = "Category",
      opacity  = 0.8
    )
}
```

---

## Price vs Size

*Each point is one listing. The grey trend line shows the general price–size relationship across all categories.*

```{r price-size, fig.height=5.5, fig.width=9}
scatter_df <- classified |>
  filter(!is.na(price_m), !is.na(size_sqm),
         price_m > 0.3, price_m < 35,
         size_sqm > 10, size_sqm < 450)

# Annotations: median markers per category
cat_med <- scatter_df |>
  group_by(category) |>
  summarise(med_sqm = median(size_sqm), med_price = median(price_m), .groups = "drop")

ggplot(scatter_df, aes(x = size_sqm, y = price_m)) +
  geom_point(aes(colour = category), alpha = 0.45, size = 1.6) +
  geom_smooth(method = "lm", formula = y ~ x,
              colour = "grey50", linewidth = 0.8, se = TRUE, alpha = 0.12) +
  scale_colour_manual(values = CAT_COLORS, name = "Category") +
  scale_y_continuous(labels = label_number(suffix = " M", accuracy = 0.5)) +
  scale_x_continuous(labels = label_number(suffix = " m²")) +
  labs(x = "Size (m²)", y = "Asking price",
       title = "Price vs size",
       subtitle = "Linear trend across all listings shown in grey") +
  theme_minimal(base_size = 13) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1),
                               ncol = 2))
```

---

## Market Segments

```{r segments, fig.height=4.5, fig.width=10}
cat_counts <- classified |>
  count(category, name = "n") |>
  arrange(desc(n)) |>
  mutate(
    category = factor(category, levels = category),
    pct      = n / sum(n) * 100,
    label    = paste0(n, "  (", round(pct), "%)")
  )

ggplot(cat_counts, aes(x = n, y = fct_rev(category), fill = category)) +
  geom_col(show.legend = FALSE, width = 0.72) +
  geom_text(aes(label = label), hjust = -0.1, size = 3.4, colour = "grey25") +
  scale_fill_manual(values = CAT_COLORS) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = "Listings", y = NULL,
       title = "Distribution of listings by AI-assigned category") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 11))
```

---

## Price Distribution by Category

```{r price-violin, fig.height=5.5, fig.width=10}
violin_df <- classified |>
  filter(!is.na(price_m), price_m > 0.3, price_m < 30) |>
  mutate(category = factor(category, levels = rev(levels(cat_counts$category))))

# Only draw violins for categories with enough data
enough <- violin_df |> count(category) |> filter(n >= 5) |> pull(category)

ggplot(violin_df |> filter(category %in% enough),
       aes(x = price_m, y = category, fill = category)) +
  geom_violin(trim = TRUE, show.legend = FALSE, alpha = 0.72, linewidth = 0.25) +
  geom_boxplot(width = 0.10, outlier.shape = NA, show.legend = FALSE,
               colour = "grey30", fill = "white", alpha = 0.55) +
  stat_summary(fun = median, geom = "point",
               shape = 21, size = 2, fill = "white", colour = "grey30",
               show.legend = FALSE) +
  scale_fill_manual(values = CAT_COLORS) +
  scale_x_continuous(labels = label_number(suffix = " M NOK", accuracy = 1),
                     limits = c(0, 30)) +
  labs(x = "Asking price", y = NULL,
       title = "Price distribution within each category",
       subtitle = "White dot = median; box = IQR") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```

---

## Value for Money

::: {.callout-tip appearance="simple"}
**How is "value for money" defined here?**

For each listing we compute a **value score**:

$$\text{Value Score} = \frac{\text{Category median price/m}^2}{\text{Listing price/m}^2}$$

A score of **1.0** means the listing is priced exactly at the median per m² for its category.
A score of **1.30** means you pay **30 % less per m²** than the typical listing in that segment — genuine value.
Scores **below 1.0** indicate premium pricing.
:::

```{r value-scatter, fig.height=5.5, fig.width=9}
val_df <- classified |>
  filter(!is.na(value_score), !is.na(size_sqm),
         size_sqm > 10, size_sqm < 450,
         value_score > 0.3, value_score < 3.5,
         price_sqm > 10000)

# Top-5 best-value picks to label
top5 <- val_df |>
  filter(!is.na(title)) |>
  arrange(desc(value_score)) |>
  slice_head(n = 5) |>
  mutate(short_label = paste0(
    str_trunc(coalesce(neighborhood, address, ""), 18, ellipsis = "…"),
    "\n", round(value_score, 2), "×"
  ))

ggplot(val_df, aes(x = size_sqm, y = value_score)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey60", linewidth = 0.7) +
  annotate("text", x = max(val_df$size_sqm, na.rm = TRUE) * 0.98, y = 1.02,
           label = "← at category median", hjust = 1, size = 3.2, colour = "grey55") +
  geom_point(aes(colour = category, size = price_m), alpha = 0.5) +
  ggrepel::geom_label_repel(
    data = top5,
    aes(label = short_label, colour = category),
    size = 3, show.legend = FALSE, fill = "white",
    label.size = 0.25, max.overlaps = 8
  ) +
  scale_colour_manual(values = CAT_COLORS, name = "Category") +
  scale_size_continuous(name = "Price (M NOK)", range = c(1.5, 6),
                        labels = label_number(suffix = " M")) +
  scale_y_continuous(labels = label_number(accuracy = 0.1, suffix = "×")) +
  scale_x_continuous(labels = label_number(suffix = " m²")) +
  labs(x = "Size (m²)", y = "Value score",
       title = "Value for money vs size",
       subtitle = "Above dashed line = cheaper per m² than category median. Labelled = top-5 picks.") +
  theme_minimal(base_size = 13) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1), ncol = 2),
         size   = guide_legend(override.aes = list(alpha = 0.7)))
```

### Top 10 Best-Value Listings

```{r value-table}
top10 <- classified |>
  filter(!is.na(value_score), !is.na(price), !is.na(size_sqm),
         size_sqm >= 25, price_sqm > 10000) |>
  arrange(desc(value_score)) |>
  slice_head(n = 10) |>
  transmute(
    Listing = ifelse(!is.na(url),
      paste0('<a href="', url, '" target="_blank">', str_trunc(coalesce(title, "—"), 40), '</a>'),
      coalesce(title, "—")),
    Price   = paste0(format(price, big.mark = " "), " NOK"),
    `Size`  = paste0(size_sqm, " m²"),
    `kr/m²` = paste0(format(round(price_sqm), big.mark = " "), " NOK"),
    Category = category,
    `Value score` = paste0(round(value_score, 2), "×  (+", round(value_pct), "%)")
  )

datatable(top10, escape = FALSE, rownames = FALSE,
          options = list(dom = "t", pageLength = 10, scrollX = TRUE))
```

---

## Price per m² by Neighbourhood

```{r psqm-hood, fig.height=5, fig.width=9}
hood_df <- classified |>
  filter(!is.na(neighborhood), neighborhood != "",
         !is.na(price_sqm), price_sqm > 10000, price_sqm < 250000)

hood_counts <- hood_df |> count(neighborhood) |> filter(n >= 3)

if (nrow(hood_counts) < 3) {
  cat("> Not enough neighbourhood data yet — will appear as more listings are scraped and geocoded.\n")
} else {
  hood_summary <- hood_df |>
    filter(neighborhood %in% hood_counts$neighborhood) |>
    group_by(neighborhood) |>
    summarise(
      med_psqm = median(price_sqm),
      n        = n(),
      .groups  = "drop"
    ) |>
    arrange(desc(med_psqm)) |>
    slice_head(n = 20) |>
    mutate(neighborhood = factor(neighborhood, levels = rev(neighborhood)))

  ggplot(hood_summary, aes(x = med_psqm / 1000, y = neighborhood)) +
    geom_col(fill = "#264653", alpha = 0.82, width = 0.7) +
    geom_text(aes(label = paste0(round(med_psqm / 1000), "k  (n=", n, ")")),
              hjust = -0.1, size = 3.2, colour = "grey30") +
    scale_x_continuous(
      labels  = label_number(suffix = "k NOK"),
      expand  = expansion(mult = c(0, 0.18))
    ) +
    labs(x = "Median price/m² (1 000 NOK)", y = NULL,
         title = "Median price per m² by neighbourhood",
         subtitle = "Top 20 neighbourhoods with ≥ 3 listings") +
    theme_minimal(base_size = 13) +
    theme(panel.grid.major.y = element_blank())
}
```

---

## Size & Rooms

```{r size-rooms, fig.height=4, fig.width=10}
sr_df <- classified |>
  filter(!is.na(rooms), rooms >= 1, rooms <= 8,
         !is.na(size_sqm), size_sqm > 10, size_sqm < 450)

if (nrow(sr_df) < 10) {
  cat("> Not enough rooms data yet.\n")
} else {
  ggplot(sr_df, aes(x = factor(rooms), y = size_sqm, fill = category)) +
    geom_boxplot(outlier.alpha = 0.3, linewidth = 0.35, alpha = 0.75) +
    scale_fill_manual(values = CAT_COLORS, name = "Category") +
    scale_y_continuous(labels = label_number(suffix = " m²")) +
    labs(x = "Number of rooms", y = "Size (m²)",
         title = "Size distribution by number of rooms",
         subtitle = "Coloured by category") +
    theme_minimal(base_size = 13) +
    guides(fill = guide_legend(ncol = 2))
}
```

---

## Daily Listing Activity

```{r daily-activity, fig.height=3.5, fig.width=9}
daily <- df |>
  count(scraped_date, name = "n") |>
  filter(!is.na(scraped_date)) |>
  arrange(scraped_date)

if (nrow(daily) < 2) {
  cat("> Daily trend will appear after at least 2 days of scraping.\n")
} else {
  ggplot(daily, aes(x = scraped_date, y = n)) +
    geom_col(fill = "#2A9D8F", alpha = 0.85, width = 0.7) +
    geom_text(aes(label = n), vjust = -0.4, size = 3, colour = "grey35") +
    scale_x_date(date_labels = "%b %d") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(x = NULL, y = "New listings",
         title = "New listings scraped per day",
         subtitle = "Cumulative unique listings added to the database each run") +
    theme_minimal(base_size = 13) +
    theme(panel.grid.major.x = element_blank())
}
```

---

## Listing of the Week :trophy:

```{r pick-of-week}
# Algorithm:
#  1. Filter to classified listings with full data (price, size, address, url)
#  2. Require size >= 30 m², valid value_score, scraped within 14 days
#     (fall back to all-time if too few recent listings)
#  3. Rank by value_score (higher = cheaper per m² vs category), break ties by size
#  4. Pick #1

pick_pool <- classified |>
  filter(!is.na(price), !is.na(size_sqm), !is.na(url),
         !is.na(value_score), price_sqm > 10000,
         size_sqm >= 30, category != "Unclassified") |>
  arrange(desc(value_score), desc(size_sqm))

# Prefer recent listings
recent_pool <- pick_pool |> filter(scraped_date >= Sys.Date() - 14)
if (nrow(recent_pool) >= 1) pick_pool <- recent_pool

pick <- pick_pool |> slice_head(n = 1)

if (nrow(pick) == 0) {
  cat("> Not enough classified listings yet to make a recommendation.\n")
} else {
  cat(htmltools::doRenderTags(
    htmltools::tags$div(
      class = "card border-success mb-4",
      style = "max-width:680px; border-width:2px;",

      htmltools::tags$div(
        class = "card-header bg-success text-white d-flex align-items-center gap-2",
        htmltools::tags$span("\U0001F3C6", style = "font-size:1.3em"),
        htmltools::tags$strong("Listing of the Week")
      ),

      htmltools::tags$div(
        class = "card-body",

        htmltools::tags$h5(
          class = "card-title mb-1",
          htmltools::tags$a(
            href = pick$url, target = "_blank", rel = "noopener",
            coalesce(pick$title, "View listing")
          )
        ),

        htmltools::tags$p(
          class = "text-muted mb-2", style = "font-size:0.9em",
          coalesce(pick$address, "Oslo")
        ),

        htmltools::tags$div(
          class = "d-flex flex-wrap gap-3 mb-3",

          htmltools::tags$div(
            htmltools::tags$div(class = "text-muted", style = "font-size:0.75em", "PRICE"),
            htmltools::tags$strong(
              if (!is.na(pick$price)) paste0(format(pick$price, big.mark = " "), " NOK") else "—"
            )
          ),
          htmltools::tags$div(
            htmltools::tags$div(class = "text-muted", style = "font-size:0.75em", "SIZE"),
            htmltools::tags$strong(
              if (!is.na(pick$size_sqm)) paste0(pick$size_sqm, " m²") else "—"
            )
          ),
          htmltools::tags$div(
            htmltools::tags$div(class = "text-muted", style = "font-size:0.75em", "ROOMS"),
            htmltools::tags$strong(
              if (!is.na(pick$rooms)) as.character(pick$rooms) else "—"
            )
          ),
          htmltools::tags$div(
            htmltools::tags$div(class = "text-muted", style = "font-size:0.75em", "PRICE / m²"),
            htmltools::tags$strong(
              if (!is.na(pick$price_sqm))
                paste0(format(round(pick$price_sqm), big.mark = " "), " NOK")
              else "—"
            )
          )
        ),

        # Category badge
        htmltools::tags$p(
          htmltools::tags$span(
            class = "badge rounded-pill text-white me-2",
            style = paste0("background-color:", CAT_COLORS[pick$category]),
            pick$category
          ),
          if (!is.na(pick$value_score))
            htmltools::tags$span(
              class = "badge bg-success-subtle text-success-emphasis",
              paste0("+", round(pick$value_pct), "% value vs category median")
            )
        ),

        # AI reasoning
        if (!is.na(pick$category_reasoning) && nchar(pick$category_reasoning) > 5)
          htmltools::tags$p(
            class = "card-text fst-italic text-muted",
            style = "font-size:0.88em; border-left:3px solid #ccc; padding-left:8px",
            paste0("\u201C", pick$category_reasoning, "\u201D")
          ),

        htmltools::tags$a(
          href = pick$url, target = "_blank", rel = "noopener",
          class = "btn btn-outline-success btn-sm",
          "View on finn.no \u2192"
        )
      ),

      htmltools::tags$div(
        class = "card-footer text-muted",
        style = "font-size:0.8em",
        paste0("Selected by value score algorithm · Scraped ", format(pick$scraped_date, "%d %b %Y"))
      )
    )
  ))
}
```


*Data scraped daily from [finn.no](https://www.finn.no/realestate/homes/search.html?location=0.20061){target="_blank"} for personal/educational use only.
Categories and value scores computed by Claude AI. This is not financial or investment advice.*
