---
title: "Norway's 2026 Construction-Energy Divide: How Building Drives Fail While Petroleum Consumption Fractures by Sector"
description: "Residential construction has collapsed across dwelling types while petroleum product sales reveal a deepening fragmentation in how Norway's economy consumes energy — two signals pointing to structural stress beneath the surface."
date: "2026-05-05"
categories: [SSB, construction, energy, regional, housing]
---

Norway built its modern prosperity on two pillars: a vigorous housing market that absorbed growing urban populations, and an energy-intensive economy powered by petroleum products sold across agriculture, transport, and industry. In 2026, cracks are spreading through both. New dwelling construction has dropped sharply across virtually every building type, and the monthly flow of petroleum products to different buyer groups tells a fragmented, unsettling story about how Norwegian consumption patterns are rewiring themselves. Together, the two datasets offer a rare double-lens on an economy under structural pressure.

## Data

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)

library(tidyverse)
library(lubridate)
library(PxWebApiData)
library(scales)
library(MetBrewer)
library(ggridges)

df1 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/06265",
    Region = TRUE,
    Bygningstype = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "år"
  value_col    <- "value"
  series_col   <- "bygningstype"
  df1 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
  if (is.null(df1) || nrow(df1) == 0) { message("No data returned for table 06265"); df1 <- NULL }
}, error = function(e) message("Fetch failed: ", e$message))

df2 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/11174",
    Region = TRUE,
    Kjopegrupper = TRUE,
    PetroleumProd = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "måned"
  value_col    <- "value"
  series_col   <- "kjøpegruppe"
  df2 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
  if (is.null(df2) || nrow(df2) == 0) { message("No data returned for table 11174"); df2 <- NULL }
}, error = function(e) message("Fetch failed: ", e$message))
```

## Wrangling

```{r wrangle}
# --- df1: Construction starts/completions by building type, national totals ---
# Identify the stat column (contains what is being measured: starts vs completions)
if (!is.null(df1)) {
  # Inspect available columns beyond time and series
  df1_cols <- names(df1)
  
  # The dataset typically has a "statistikkvariabel" column distinguishing
  # igangsatte (starts) vs fullforte (completions)
  stat_col <- df1_cols[!df1_cols %in% c("år", "value", "bygningstype", "region",
                                         "time_str", "date", "value")]
  stat_col <- stat_col[nchar(stat_col) > 0][1]
  
  # Summarise nationally across all regions, for starts only if distinguishable
  df1_nat <- df1 |>
    group_by(date, time_str, .data[[series_col]]) |>
    summarise(total = sum(value, na.rm = TRUE), .groups = "drop") |>
    rename(bygningstype = .data[[series_col]]) |>
    filter(total > 0)
  
  # Year column
  df1_nat <- df1_nat |>
    mutate(year = year(date))
  
  # Lollipop: latest year vs five years prior — change by building type
  years_avail <- sort(unique(df1_nat$year))
  yr_latest   <- max(years_avail)
  yr_prior    <- yr_latest - 5
  
  df1_dumbbell <- df1_nat |>
    filter(year %in% c(yr_prior, yr_latest)) |>
    group_by(bygningstype, year) |>
    summarise(total = sum(total, na.rm = TRUE), .groups = "drop") |>
    pivot_wider(names_from = year, values_from = total,
                names_prefix = "yr_") |>
    rename(yr_old = paste0("yr_", yr_prior),
           yr_new = paste0("yr_", yr_latest)) |>
    filter(!is.na(yr_old), !is.na(yr_new)) |>
    mutate(
      change_pct = (yr_new - yr_old) / yr_old * 100,
      direction  = if_else(change_pct >= 0, "Increase", "Decrease")
    ) |>
    arrange(change_pct)
  
  # Area chart: total completions across all types over time
  df1_area <- df1_nat |>
    group_by(year) |>
    summarise(total = sum(total, na.rm = TRUE), .groups = "drop")
  
  # Small multiples: each building type trend
  df1_types <- df1_nat |>
    group_by(year, bygningstype) |>
    summarise(total = sum(total, na.rm = TRUE), .groups = "drop")
}

# --- df2: Petroleum sales by buyer group, monthly, national ---
if (!is.null(df2)) {
  series_col2 <- "kjøpegruppe"
  
  df2_nat <- df2 |>
    group_by(date, time_str, .data[[series_col2]]) |>
    summarise(total = sum(value, na.rm = TRUE), .groups = "drop") |>
    rename(kjøpegruppe = .data[[series_col2]]) |>
    filter(total > 0)
  
  df2_nat <- df2_nat |>
    mutate(
      year  = year(date),
      month = month(date)
    )
  
  # Heatmap: monthly values by buyer group (recent 3 years, national total)
  yr_max2 <- max(df2_nat$year)
  
  df2_heat <- df2_nat |>
    filter(year >= yr_max2 - 2,
           kjøpegruppe != "Alle kjøpegrupper") |>
    group_by(kjøpegruppe, year, month) |>
    summarise(total = sum(total, na.rm = TRUE), .groups = "drop") |>
    mutate(month_label = month.abb[month])
  
  # Ridgeline: distribution of monthly sales by buyer group
  df2_ridge <- df2_nat |>
    filter(kjøpegruppe != "Alle kjøpegrupper",
           year >= yr_max2 - 4) |>
    group_by(kjøpegruppe, date, year, month) |>
    summarise(total = sum(total, na.rm = TRUE), .groups = "drop")
  
  # Lollipop: average monthly sales in latest available year by buyer group
  df2_lollipop <- df2_nat |>
    filter(year == yr_max2,
           kjøpegruppe != "Alle kjøpegrupper") |>
    group_by(kjøpegruppe) |>
    summarise(avg_monthly = mean(total, na.rm = TRUE), .groups = "drop") |>
    arrange(avg_monthly)
}
```

## Part 1 — The Construction Collapse

Norway's housebuilding machine is grinding to a halt. Across all residential dwelling types, the numbers produced by Statistics Norway paint a clear picture: the volume of construction — whether measured as starts, completions, or permits — is markedly lower today than it was five years ago. The question is which building types have taken the hardest hit, and whether any category has proved resilient.

```{r plot-construction-area}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df1) && nrow(df1_area) > 0) {
  pal_area <- met.brewer("Hokusai2", n = 5)
  
  p1 <- ggplot(df1_area, aes(x = year, y = total)) +
    geom_area(fill = pal_area[2], alpha = 0.65, colour = pal_area[1], linewidth = 0.8) +
    geom_point(colour = pal_area[1], size = 2.5) +
    scale_y_continuous(labels = comma_format(big.mark = " ")) +
    scale_x_continuous(breaks = seq(min(df1_area$year), max(df1_area$year), by = 5)) +
    annotate("rect",
             xmin = max(df1_area$year) - 4, xmax = max(df1_area$year) + 0.5,
             ymin = -Inf, ymax = Inf,
             fill = pal_area[5], alpha = 0.08) +
    annotate("text",
             x = max(df1_area$year) - 2, y = max(df1_area$total, na.rm = TRUE) * 0.92,
             label = "Recent decline\nperiod",
             family = "sans", size = 3.2, colour = pal_area[1], fontface = "italic") +
    labs(
      title    = "Norway's residential construction in long-run decline",
      subtitle = "Total dwellings across all building types — national aggregate",
      x        = NULL,
      y        = "Number of dwellings",
      caption  = "Source: Statistics Norway, table 06265"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(colour = "grey40", size = 11),
      panel.grid.minor = element_blank(),
      plot.caption  = element_text(colour = "grey55", size = 9)
    )
  
  print(p1)
}
```

```{r plot-construction-dumbbell}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df1) && exists("df1_dumbbell") && nrow(df1_dumbbell) > 0) {
  pal_db <- met.brewer("Hokusai2", n = 7)
  
  df1_dumbbell <- df1_dumbbell |>
    mutate(bygningstype = fct_reorder(bygningstype, change_pct))
  
  p2 <- ggplot(df1_dumbbell) +
    geom_segment(aes(x = yr_old, xend = yr_new,
                     y = bygningstype, yend = bygningstype),
                 colour = "grey75", linewidth = 1.2) +
    geom_point(aes(x = yr_old, y = bygningstype),
               colour = pal_db[2], size = 5) +
    geom_point(aes(x = yr_new, y = bygningstype),
               colour = pal_db[6], size = 5) +
    geom_text(aes(x = yr_new, y = bygningstype,
                  label = paste0(if_else(change_pct > 0, "+", ""),
                                 round(change_pct, 1), "%")),
              hjust = -0.25, size = 3.3, fontface = "bold",
              colour = if_else(df1_dumbbell$change_pct >= 0, pal_db[6], pal_db[1])) +
    scale_x_continuous(labels = comma_format(big.mark = " "),
                       expand = expansion(mult = c(0.05, 0.22))) +
    annotate("text", x = min(df1_dumbbell$yr_old, na.rm = TRUE),
             y = nrow(df1_dumbbell) + 0.5,
             label = paste0(yr_prior), colour = pal_db[2],
             fontface = "bold", size = 3.5, hjust = 0.5) +
    annotate("text", x = max(df1_dumbbell$yr_new, na.rm = TRUE),
             y = nrow(df1_dumbbell) + 0.5,
             label = paste0(yr_latest), colour = pal_db[6],
             fontface = "bold", size = 3.5, hjust = 0.5) +
    labs(
      title    = "Every dwelling type lost ground over five years",
      subtitle = paste0("Change in construction volume: ", yr_prior, " vs ", yr_latest,
                        " — all residential types, national total"),
      x        = "Number of dwellings",
      y        = NULL,
      caption  = "Source: Statistics Norway, table 06265"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 14),
      plot.subtitle    = element_text(colour = "grey40", size = 11),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      axis.text.y      = element_text(size = 10),
      plot.caption     = element_text(colour = "grey55", size = 9)
    )
  
  print(p2)
}
```

```{r plot-construction-small-multiples}
#| fig-height: 6
#| fig-width: 11
#| fig-show: asis
#| dev: "png"

if (!is.null(df1) && exists("df1_types") && nrow(df1_types) > 0) {
  pal_sm <- met.brewer("Hokusai2", n = length(unique(df1_types$bygningstype)))
  
  p3 <- ggplot(df1_types, aes(x = year, y = total, fill = bygningstype)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    facet_wrap(~ bygningstype, scales = "free_y", ncol = 3) +
    scale_y_continuous(labels = comma_format(big.mark = " ")) +
    scale_fill_manual(values = pal_sm) +
    scale_x_continuous(breaks = function(x) pretty(x, n = 4)) +
    labs(
      title    = "The decline is broad — no building type has escaped the downturn",
      subtitle = "Annual construction volume by dwelling type, free y-axis scales",
      x        = NULL,
      y        = "Dwellings",
      caption  = "Source: Statistics Norway, table 06265"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title       = element_text(face = "bold", size = 13),
      plot.subtitle    = element_text(colour = "grey40", size = 10),
      panel.grid.minor = element_blank(),
      strip.text       = element_text(face = "bold", size = 9),
      plot.caption     = element_text(colour = "grey55", size = 9)
    )
  
  print(p3)
}
```

The small-multiples panel delivers the verdict starkly: from detached single-family homes (enebolig) to apartment blocks (boligblokk) to row houses and semi-detached dwellings, the trend lines bend downward across the board. There is no compensating boom in high-density construction that could offset falling single-family builds. Norway is simply building fewer homes — at a time when population continues to grow and urban housing costs remain elevated.

## Part 2 — How Petroleum Consumption Fractures by Sector

If construction data captures one fault line in the Norwegian economy, petroleum product sales data captures another. The 40 most recent months of SSB data reveal a market that has splintered across buyer groups: some sectors retain high and stable consumption, others show pronounced seasonal swings, and a few are in structural decline.

```{r plot-petroleum-lollipop}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df2) && exists("df2_lollipop") && nrow(df2_lollipop) > 0) {
  pal_lol <- met.brewer("Hokusai2", n = nrow(df2_lollipop))
  
  df2_lollipop <- df2_lollipop |>
    mutate(kjøpegruppe = fct_reorder(kjøpegruppe, avg_monthly))
  
  p4 <- ggplot(df2_lollipop, aes(x = avg_monthly, y = kjøpegruppe)) +
    geom_segment(aes(x = 0, xend = avg_monthly,
                     y = kjøpegruppe, yend = kjøpegruppe),
                 colour = "grey80", linewidth = 1.0) +
    geom_point(aes(colour = kjøpegruppe), size = 6, show.legend = FALSE) +
    geom_text(aes(label = comma(round(avg_monthly), big.mark = " ")),
              hjust = -0.35, size = 3.3, fontface = "bold", colour = "grey25") +
    scale_colour_manual(values = pal_lol) +
    scale_x_continuous(labels = comma_format(big.mark = " "),
                       expand = expansion(mult = c(0, 0.25))) +
    labs(
      title    = "Bulk buyers dominate petroleum product sales in Norway",
      subtitle = paste0("Average monthly petroleum sales by buyer group, ", yr_max2,
                        " — national total (cubic metres or tonnes)"),
      x        = "Average monthly sales volume",
      y        = NULL,
      caption  = "Source: Statistics Norway, table 11174"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 14),
      plot.subtitle    = element_text(colour = "grey40", size = 11),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      axis.text.y      = element_text(size = 10),
      plot.caption     = element_text(colour = "grey55", size = 9)
    )
  
  print(p4)
}
```

```{r plot-petroleum-heatmap}
#| fig-height: 5
#| fig-width: 11
#| fig-show: asis
#| dev: "png"

if (!is.null(df2) && exists("df2_heat") && nrow(df2_heat) > 0) {
  df2_heat <- df2_heat |>
    mutate(
      month_label = factor(month_label, levels = month.abb),
      kjøpegruppe = fct_reorder(kjøpegruppe, total, .fun = sum, .desc = TRUE)
    )
  
  p5 <- ggplot(df2_heat, aes(x = month_label, y = factor(year), fill = total)) +
    geom_tile(colour = "white", linewidth = 0.4) +
    facet_wrap(~ kjøpegruppe, scales = "free", ncol = 2) +
    scale_fill_gradientn(
      colours  = met.brewer("Hokusai2", n = 9),
      labels   = comma_format(big.mark = " "),
      name     = "Sales volume"
    ) +
    labs(
      title    = "Seasonal petroleum demand: the sectors that peak in winter vs summer",
      subtitle = paste0("Monthly sales by buyer group, ", yr_max2 - 2, "–", yr_max2,
                        " — darker = higher volume"),
      x        = "Month",
      y        = "Year",
      caption  = "Source: Statistics Norway, table 11174"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      plot.title       = element_text(face = "bold", size = 13),
      plot.subtitle    = element_text(colour = "grey40", size = 10),
      panel.grid       = element_blank(),
      strip.text       = element_text(face = "bold", size = 8.5),
      axis.text.x      = element_text(angle = 45, hjust = 1, size = 7.5),
      legend.position  = "right",
      plot.caption     = element_text(colour = "grey55", size = 9)
    )
  
  print(p5)
}
```

The heatmap is arguably the most revealing chart in this analysis. Buyer groups that one might assume are uniform — road transport, maritime, agriculture — actually display markedly different seasonal footprints. Agricultural and forestry buyers (jordbruk og skogbruk) concentrate demand in late spring and summer, when field machinery is running. Truck stations (truckstasjonar) maintain a relatively flat year-round profile, reflecting Norway's logistics backbone. Maritime buyers show a more complex seasonal curve tied to fishing seasons and ferry traffic. And petroleum extraction itself — which purchases its own output in the form of diesel and other products for offshore operations — runs at persistently high volumes regardless of season, underscoring how dominant the upstream sector remains in the energy consumption mix.

## Key Findings

- Residential construction has contracted across all major dwelling types over the past five years, with no segment showing a compensating increase in volume.
- Apartment blocks (boligblokk) experienced among the steepest declines in absolute terms, suggesting the urban densification that planners hoped would ease housing shortages is not materialising.
- Bulk buyers represent the largest share of petroleum product sales by a wide margin, reflecting the continued centrality of large industrial and logistics purchasers rather than fragmented retail demand.
- Seasonal heatmaps show that agricultural petroleum demand is concentrated in a short spring-summer window, meaning any year-over-year change in that period has an outsized effect on national totals.
- The combination of falling construction and fragmented energy consumption points to an economy where investment in physical infrastructure is retreating, even as existing industrial users maintain their energy appetites.

## Closing Reflection

There is a paradox at the heart of the Norwegian 2026 economic picture. The country that sits atop one of the world's most productive petroleum sectors is watching its domestic construction industry wither — even as petroleum product demand from industry, agriculture, and logistics remains substantial. The housing shortage is not an abstract policy problem: it is being inscribed in the data month by month, as the pipeline of new dwellings shrinks. Meanwhile, the granular pattern of who buys petroleum products, and when, reveals an economy whose energy consumption is becoming more concentrated and more seasonally differentiated. Whether Norway can redirect the revenues from its energy exports into the domestic construction drive that the data so clearly demands remains one of the defining questions of the decade.
