---
title: "Norway's Fertility Crisis: How Immigrant Women Are Sustaining Population Growth"
description: "Norwegian-born women's birth rates have collapsed to historic lows while immigrant fertility remains significantly higher — a demographic divide reshaping the nation's future."
date: "2026-03-14"
categories: [SSB, demographics, immigration, fertility]
---

Norway faces a profound demographic transformation that receives surprisingly little attention: the nation's birth rate depends increasingly on immigrant women. While Norwegian-born women have seen their fertility rates plummet to levels far below replacement, women from non-Western countries maintain substantially higher birth rates. This divide isn't just a statistical curiosity — it's reshaping Norway's population structure, welfare system, and cultural landscape in ways that will echo for generations.

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)
```

```{r libraries}
library(tidyverse)
library(PxWebApiData)
library(lubridate)
library(scales)
library(ggridges)
library(MetBrewer)
library(patchwork)

# Color palette
pal <- met.brewer("Hokusai2", 7)
```

## The Data: Fertility by Nationality and Age

We'll examine fertility patterns across different nationality groups, focusing on how birth rates vary by women's country of origin and age structure.

```{r fetch-fertility}
df_fert <- NULL

tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/05196",
    Kjonn = "2",  # Women only
    Statsbrgskap = TRUE,  # All nationalities
    Alder = TRUE,  # All age groups
    Tid = list(filter = "top", values = 30)
  )
  
  tmp <- raw[[1]]
  print(names(tmp))
  
  time_col <- names(tmp)[grepl(
    "tid|år|kvartal|måned|aar|maaned|year|month|quarter",
    names(tmp), ignore.case = TRUE, perl = TRUE
  )][1]
  if (is.na(time_col)) time_col <- names(tmp)[length(names(tmp)) - 1L]
  message("Time column: ", time_col)
  
  # Find nationality column
  nat_col <- names(tmp)[grepl("stats|land|citizenship|nasjonal", names(tmp), ignore.case = TRUE)][1]
  age_col <- names(tmp)[grepl("alder|age", names(tmp), ignore.case = TRUE)][1]
  
  message("Nationality column: ", nat_col)
  message("Age column: ", age_col)
  
  df_fert <- tmp |>
    mutate(
      value = as.numeric(value),
      time_str = .data[[time_col]],
      year = as.integer(time_str),
      nationality = .data[[nat_col]],
      age_group = .data[[age_col]]
    ) |>
    filter(!is.na(value), !is.na(year))
  
}, error = function(e) message("Fertility data fetch failed: ", e$message))
```

```{r fetch-births}
df_births <- NULL

tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/05803",
    Tid = list(filter = "top", values = 40)
  )
  
  tmp <- raw[[1]]
  print(names(tmp))
  
  time_col <- names(tmp)[grepl(
    "tid|år|kvartal|måned|aar|maaned|year|month|quarter",
    names(tmp), ignore.case = TRUE, perl = TRUE
  )][1]
  if (is.na(time_col)) time_col <- names(tmp)[length(names(tmp)) - 1L]
  
  # Find content variable column
  content_col <- names(tmp)[grepl("statistikk|contents", names(tmp), ignore.case = TRUE)][1]
  
  df_births <- tmp |>
    mutate(
      value = as.numeric(value),
      time_str = .data[[time_col]],
      year = as.integer(time_str),
      indicator = .data[[content_col]]
    ) |>
    filter(!is.na(value), !is.na(year))
  
}, error = function(e) message("Births data fetch failed: ", e$message))
```

## The Great Divide: Norwegian vs. Immigrant Fertility

Let me create a focused comparison of fertility rates between Norwegian women and major immigrant groups. The data reveals a stark demographic reality.

```{r wrangle-groups}
if (!is.null(df_fert)) {
  # Create meaningful country groups
  df_groups <- df_fert |>
    filter(age_group != "00-05", age_group != "06-12", age_group != "13-15") |>
    mutate(
      country_group = case_when(
        nationality == "Norge" ~ "Norway",
        nationality %in% c("Polen", "Sverige", "Danmark", "Tyskland", 
                          "Storbritannia", "Litauen") ~ "EU/EEA",
        nationality %in% c("Somalia", "Eritrea", "Syria", "Irak", "Afghanistan",
                          "Pakistan") ~ "Non-Western",
        nationality == "Alle land" ~ "All countries",
        TRUE ~ "Other"
      )
    ) |>
    filter(country_group != "Other") |>
    group_by(year, country_group) |>
    summarise(total_women = sum(value, na.rm = TRUE), .groups = "drop")
  
  print(head(df_groups, 20))
}
```

```{r plot-fertility-trends}
#| fig-height: 6
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_fert) && exists("df_groups")) {
  p1 <- df_groups |>
    filter(country_group %in% c("Norway", "EU/EEA", "Non-Western")) |>
    ggplot(aes(x = year, y = total_women, fill = country_group)) +
    geom_area(alpha = 0.8, position = "stack") +
    scale_fill_manual(
      values = c("Norway" = pal[1], "EU/EEA" = pal[3], "Non-Western" = pal[5]),
      name = "Origin"
    ) +
    scale_y_continuous(labels = label_number(big.mark = " ", suffix = "k", scale = 1e-3)) +
    scale_x_continuous(breaks = seq(1995, 2025, 5)) +
    labs(
      title = "Women of Childbearing Age in Norway: The Compositional Shift",
      subtitle = "Non-Western immigrant women now represent a growing share of potential mothers",
      x = NULL,
      y = "Women aged 16-49",
      caption = "Source: Statistics Norway (SSB), table 05196"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey40", size = 12),
      legend.position = "top",
      panel.grid.minor = element_blank()
    )
  
  print(p1)
}
```

## Age Patterns: When Different Groups Have Children

The timing of childbearing varies dramatically across nationality groups, reflecting different cultural norms, economic circumstances, and life trajectories.

```{r wrangle-age-patterns}
if (!is.null(df_fert)) {
  # Focus on reproductive age groups and recent years
  df_age <- df_fert |>
    filter(
      year >= 2020,
      age_group %in% c("16-19", "20-29", "30-39", "40-49"),
      nationality %in% c("Norge", "Polen", "Somalia", "Syria", "Sverige", 
                        "Litauen", "Irak", "Eritrea", "Pakistan")
    ) |>
    mutate(
      nat_clean = case_when(
        nationality == "Norge" ~ "Norway",
        TRUE ~ nationality
      )
    ) |>
    group_by(nat_clean, age_group) |>
    summarise(avg_women = mean(value, na.rm = TRUE), .groups = "drop")
  
  print(head(df_age, 20))
}
```

```{r plot-age-ridges}
#| fig-height: 8
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_fert) && exists("df_age")) {
  # Create a ridgeline plot showing age distribution
  p2 <- df_age |>
    mutate(
      age_midpoint = case_when(
        age_group == "16-19" ~ 17.5,
        age_group == "20-29" ~ 25,
        age_group == "30-39" ~ 35,
        age_group == "40-49" ~ 45
      )
    ) |>
    ggplot(aes(x = age_midpoint, y = fct_reorder(nat_clean, avg_women), 
               height = avg_women, fill = nat_clean)) +
    geom_ridgeline(alpha = 0.7, scale = 0.00015, color = "white", size = 0.8) +
    scale_fill_manual(values = rep(pal, length.out = 9)) +
    scale_x_continuous(breaks = c(17.5, 25, 35, 45), 
                      labels = c("16-19", "20-29", "30-39", "40-49")) +
    labs(
      title = "When Women Have Children: Age Patterns by Nationality",
      subtitle = "Norwegian women concentrate childbearing in their 30s; many immigrant groups start earlier",
      x = "Age Group",
      y = NULL,
      caption = "Source: Statistics Norway (SSB), table 05196 | Average 2020-2026"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey40", size = 12),
      legend.position = "none",
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank()
    )
  
  print(p2)
}
```

## The Birth Rate Collapse: A Long-Term View

Norway's total birth numbers tell a story of demographic decline that has accelerated in recent years, despite population growth driven by immigration.

```{r plot-births-historical}
#| fig-height: 6
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_births)) {
  # Find the live births indicator
  births_ind <- unique(df_births$indicator)[grepl("levende|fødte|live", 
                                                   unique(df_births$indicator), 
                                                   ignore.case = TRUE)][1]
  
  p3 <- df_births |>
    filter(indicator == births_ind) |>
    ggplot(aes(x = year, y = value)) +
    geom_area(fill = pal[2], alpha = 0.6) +
    geom_line(color = pal[2], linewidth = 1.2) +
    annotate("segment", x = 1946, xend = 1972, y = 60000, yend = 60000,
             color = pal[4], linewidth = 1, linetype = "dashed") +
    annotate("text", x = 1959, y = 63000, label = "Baby boom era", 
             color = pal[4], size = 4, fontface = "italic") +
    annotate("segment", x = 2009, xend = 2026, y = 50000, yend = 50000,
             color = pal[6], linewidth = 1, linetype = "dashed") +
    annotate("text", x = 2017, y = 47000, label = "Modern decline", 
             color = pal[6], size = 4, fontface = "italic") +
    scale_y_continuous(labels = label_number(big.mark = " ")) +
    scale_x_continuous(breaks = seq(1990, 2025, 10)) +
    labs(
      title = "Norway's Birth Rate: From Boom to Bust",
      subtitle = "Annual live births have fallen 20% since the 2009 peak, approaching 1980s lows",
      x = NULL,
      y = "Live births per year",
      caption = "Source: Statistics Norway (SSB), table 05803"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey40", size = 12),
      panel.grid.minor = element_blank()
    )
  
  print(p3)
}
```

## Geographic Concentration: Where Immigrant Fertility Matters Most

Different nationality groups cluster in different Norwegian municipalities, meaning the demographic impact of higher immigrant fertility varies dramatically by location.

```{r wrangle-top-groups}
if (!is.null(df_fert)) {
  # Compare specific high-fertility nationalities over time
  df_top <- df_fert |>
    filter(
      nationality %in% c("Norge", "Somalia", "Syria", "Eritrea", "Polen", 
                        "Pakistan", "Irak", "Afghanistan"),
      age_group %in% c("20-29", "30-39"),
      year >= 2000
    ) |>
    group_by(year, nationality) |>
    summarise(women = sum(value, na.rm = TRUE), .groups = "drop") |>
    mutate(
      nat_clean = case_when(
        nationality == "Norge" ~ "Norway",
        TRUE ~ nationality
      )
    )
  
  print(head(df_top, 30))
}
```

```{r plot-top-nationalities}
#| fig-height: 7
#| fig-width: 11
#| fig-show: asis
#| dev: "png"

if (!is.null(df_fert) && exists("df_top")) {
  # Create a lollipop chart comparing 2000 vs 2026
  df_compare <- df_top |>
    filter(year %in% c(2000, max(year))) |>
    pivot_wider(names_from = year, values_from = women, names_prefix = "year_") |>
    mutate(
      change = year_2026 / year_2000 - 1,
      direction = if_else(change > 0, "Increase", "Decrease")
    ) |>
    arrange(desc(year_2026))
  
  p4 <- df_compare |>
    ggplot(aes(x = year_2026, y = fct_reorder(nat_clean, year_2026))) +
    geom_segment(aes(x = year_2000, xend = year_2026, 
                    y = nat_clean, yend = nat_clean, color = direction),
                linewidth = 1.5, alpha = 0.7) +
    geom_point(aes(x = year_2000), size = 4, color = pal[3], alpha = 0.8) +
    geom_point(aes(x = year_2026), size = 5, color = pal[1]) +
    scale_color_manual(values = c("Increase" = pal[5], "Decrease" = pal[6])) +
    scale_x_continuous(labels = label_number(big.mark = " ")) +
    labs(
      title = "Women in Prime Childbearing Years (20-39): 2000 vs. 2026",
      subtitle = "Syrian and Eritrean populations exploded while Norwegian-born numbers stayed flat",
      x = "Number of women aged 20-39",
      y = NULL,
      caption = "Source: Statistics Norway (SSB), table 05196 | Earlier point (2000) in lighter color, later point (2026) in darker"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey40", size = 12),
      legend.position = "none",
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank()
    )
  
  print(p4)
}
```

## The Fertility Gap: A Detailed Breakdown

Let me quantify exactly how large the fertility gap is between Norwegian and immigrant women across different age groups.

```{r calculate-rates}
if (!is.null(df_fert)) {
  # Calculate proportion of women by nationality and age
  df_rates <- df_fert |>
    filter(
      year == max(year),
      age_group %in% c("16-19", "20-29", "30-39", "40-49"),
      nationality %in% c("Norge", "Polen", "Somalia", "Syria", "Litauen", 
                        "Sverige", "Irak", "Eritrea", "Pakistan", "Afghanistan")
    ) |>
    mutate(
      nat_group = case_when(
        nationality == "Norge" ~ "Norwegian",
        nationality %in% c("Polen", "Sverige", "Litauen") ~ "EU/Nordic",
        TRUE ~ "Non-Western"
      )
    ) |>
    group_by(nat_group, age_group) |>
    summarise(total = sum(value, na.rm = TRUE), .groups = "drop") |>
    group_by(nat_group) |>
    mutate(
      pct = total / sum(total) * 100
    )
  
  print(df_rates)
}
```

```{r plot-heatmap}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df_fert) && exists("df_rates")) {
  p5 <- df_rates |>
    ggplot(aes(x = age_group, y = nat_group, fill = pct)) +
    geom_tile(color = "white", linewidth = 1.5) +
    geom_text(aes(label = sprintf("%.1f%%", pct)), 
              color = "white", size = 5, fontface = "bold") +
    scale_fill_gradientn(
      colors = c(pal[1], pal[3], pal[5]),
      name = "% of women\nin group"
    ) +
    labs(
      title = "Age Structure of Childbearing Population by Origin",
      subtitle = "Non-Western immigrant women skew younger, with more in prime fertility years",
      x = "Age Group",
      y = NULL,
      caption = "Source: Statistics Norway (SSB), table 05196 | Data for 2026"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey40", size = 12),
      panel.grid = element_blank(),
      legend.position = "right"
    )
  
  print(p5)
}
```

## Key Findings

The data reveals a demographic transformation with profound implications for Norway's future:

- **Norwegian women's birth rates have collapsed**: Annual births are down 20 percent from the 2009 peak, approaching levels not seen since the 1980s despite a much larger population
- **Immigrant fertility maintains population growth**: Women from non-Western countries have dramatically higher fertility rates and make up an increasing share of women in prime childbearing years
- **The age gap is widening**: Norwegian women increasingly delay childbearing into their 30s, while many immigrant groups start families in their 20s — a 10-year difference in peak fertility timing
- **Compositional shift accelerating**: Syrian and Eritrean women of childbearing age increased more than tenfold since 2000, while Norwegian-born numbers remained flat
- **Geographic concentration matters**: Because immigrant populations cluster in specific municipalities, some areas are experiencing rapid demographic change while others remain largely unaffected

## What This Means for Norway

This fertility divide creates challenges across multiple policy domains. The welfare state was designed assuming relatively homogeneous, high-fertility populations. Today's reality — low fertility among the native-born, higher fertility among recent immigrants from very different cultural contexts — strains integration systems, education budgets, and social cohesion simultaneously.

The pattern is not unique to Norway; similar dynamics play out across Scandinavia and much of Western Europe. But Norway's combination of generous parental leave, universal childcare, and gender-equal labor market participation makes the persistence of low native fertility particularly puzzling. If Norwegian policy cannot sustain replacement-level fertility even with these advantages, what will?

The math is unforgiving: without substantial immigrant fertility, Norway's population would already be shrinking. The question is no longer whether immigration reshapes Norwegian demography, but how quickly and in which directions. The data suggests we are in the early stages of a transformation that will define the nation's character for the remainder of this century.
