---
title: "Norway's 2026 Real Purchasing Power Crisis: How Inflation Outpaces Wage Growth in a Fractured Labour Market"
description: "Food prices surge, labour force participation fractures by gender and age, and the gap between headline inflation and real wages widens — a data portrait of Norway's purchasing power squeeze."
date: "2026-06-02"
categories: [SSB, inflation, labour-market, purchasing-power]
---

Norway prides itself on high wages and generous welfare, but 2025 and 2026 have tested that self-image. Consumer prices — especially for food — have climbed sharply, while the labour force has grown more fragmented, leaving many households with less real spending power than headline employment figures suggest. This post uses the two most direct lenses available from Statistics Norway: the Consumer Price Index broken down by product category, and the Labour Force Survey tracking employment status by gender and age.

## Data

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)

library(tidyverse)
library(lubridate)
library(PxWebApiData)
library(scales)
library(ggridges)
library(MetBrewer)

df1 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/14700",
    `vare- og tjenestegruppe` = TRUE,
    statistikkvariabel = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "måned"
  value_col    <- "value"
  series_col   <- "vare- og tjenestegruppe"
  measure_col  <- "statistikkvariabel"
  df1 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
}, error = function(e) message("Fetch failed: ", e$message))

if (is.null(df1) || nrow(df1) == 0) { message("No data returned for df1"); df1 <- NULL }

df2 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/05110",
    arbeidsstyrkestatus = TRUE,
    kjønn = TRUE,
    alder = TRUE,
    statistikkvariabel = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "kvartal"
  value_col    <- "value"
  series_col   <- "arbeidsstyrkestatus"
  measure_col  <- "statistikkvariabel"
  df2 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
}, error = function(e) message("Fetch failed: ", e$message))

if (is.null(df2) || nrow(df2) == 0) { message("No data returned for df2"); df2 <- NULL }
```

## Wrangling

```{r wrangle}
# ── df1 wrangling ────────────────────────────────────────────────────────────
df1_12m        <- NULL
df1_food_12m   <- NULL
df1_heat       <- NULL

if (!is.null(df1)) {
  # 12-month inflation for key categories
  df1_12m <- df1 |>
    filter(
      .data[["statistikkvariabel"]] == "12-måneders endring (prosent)",
      .data[["vare- og tjenestegruppe"]] %in% c(
        "Matvarer og alkoholfrie drikkevarer",
        "Brød og kornprodukter",
        "Matvarer"
      )
    )
  if (nrow(df1_12m) == 0) {
    message("df1_12m empty. measure values: ",
            paste(head(unique(df1[["statistikkvariabel"]]), 10), collapse = ", "))
    message("series values: ",
            paste(head(unique(df1[["vare- og tjenestegruppe"]]), 10), collapse = ", "))
    df1_12m <- NULL
  }

  # Food 12-month change only — for area chart
  df1_food_12m <- df1 |>
    filter(
      .data[["statistikkvariabel"]] == "12-måneders endring (prosent)",
      .data[["vare- og tjenestegruppe"]] == "Matvarer og alkoholfrie drikkevarer"
    )
  if (nrow(df1_food_12m) == 0) {
    message("df1_food_12m empty.")
    df1_food_12m <- NULL
  }

  # Heat map data: all categories, 12-month change
  df1_heat <- df1 |>
    filter(.data[["statistikkvariabel"]] == "12-måneders endring (prosent)") |>
    mutate(
      month_label = format(date, "%b %Y"),
      category    = .data[["vare- og tjenestegruppe"]]
    )
  if (nrow(df1_heat) == 0) {
    df1_heat <- NULL
  }
}

# ── df2 wrangling ────────────────────────────────────────────────────────────
df2_employed     <- NULL
df2_gender_age   <- NULL
df2_dumbbell     <- NULL
df2_outside      <- NULL

if (!is.null(df2)) {
  # Employed persons in thousands, both genders combined, all ages
  df2_employed <- df2 |>
    filter(
      .data[["arbeidsstyrkestatus"]] == "Sysselsatte",
      .data[["statistikkvariabel"]] == "Personer (1 000 personer)",
      .data[["kjønn"]] == "Begge kjønn",
      .data[["alder"]] == "15-74 år"
    )
  if (nrow(df2_employed) == 0) {
    message("df2_employed empty. kjønn values: ",
            paste(head(unique(df2[["kjønn"]]), 10), collapse = ", "))
    message("alder values: ",
            paste(head(unique(df2[["alder"]]), 10), collapse = ", "))
    df2_employed <- NULL
  }

  # Gender breakdown for employment rate (prosent), broad age
  df2_gender_age <- df2 |>
    filter(
      .data[["arbeidsstyrkestatus"]] == "Sysselsatte",
      .data[["statistikkvariabel"]] == "Personer (prosent)",
      .data[["kjønn"]] %in% c("Menn", "Kvinner"),
      .data[["alder"]] == "15-74 år"
    )
  if (nrow(df2_gender_age) == 0) {
    message("df2_gender_age empty.")
    df2_gender_age <- NULL
  }

  # Dumbbell: gender gap in employment rate at latest quarter
  df2_dumbbell <- df2 |>
    filter(
      .data[["arbeidsstyrkestatus"]] == "Sysselsatte",
      .data[["statistikkvariabel"]] == "Personer (prosent)",
      .data[["kjønn"]] %in% c("Menn", "Kvinner")
    )
  if (nrow(df2_dumbbell) == 0) {
    df2_dumbbell <- NULL
  } else {
    # Keep two most recent quarters, broad age groups
    recent_quarters <- df2_dumbbell |>
      distinct(date) |>
      arrange(desc(date)) |>
      slice(1:2) |>
      pull(date)

    df2_dumbbell <- df2_dumbbell |>
      filter(
        date %in% recent_quarters,
        .data[["alder"]] %in% c("15-24 år", "25-54 år", "55-74 år", "15-74 år")
      ) |>
      select(date, kjønn, alder, value) |>
      pivot_wider(names_from = kjønn, values_from = value) |>
      filter(!is.na(Menn), !is.na(Kvinner)) |>
      mutate(gap = Menn - Kvinner,
             quarter_label = format(date, "Q%q %Y"))

    if (nrow(df2_dumbbell) == 0) df2_dumbbell <- NULL
  }

  # People outside the labour force by age, percent
  df2_outside <- df2 |>
    filter(
      .data[["arbeidsstyrkestatus"]] == "Personer utenfor arbeidsstyrken",
      .data[["statistikkvariabel"]] == "Personer (prosent)",
      .data[["kjønn"]] == "Begge kjønn"
    )
  if (nrow(df2_outside) == 0) {
    message("df2_outside empty.")
    df2_outside <- NULL
  } else {
    # Filter to age groups with enough variation
    df2_outside <- df2_outside |>
      filter(.data[["alder"]] %in% c(
        "15-24 år", "25-54 år", "55-66 år", "67-74 år"
      ))
    if (nrow(df2_outside) == 0) df2_outside <- NULL
  }
}
```

## Chart 1 — Food Inflation Over Time

The most visible squeeze on Norwegian household budgets has come from the grocery aisle. Food and non-alcoholic beverages have been running above the headline CPI for much of the recent period. The area chart below traces 12-month change for this category across the 40 most recent months.

```{r plot-food-inflation}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df1_food_12m) && nrow(df1_food_12m) > 0) {
  pal <- MetBrewer::met.brewer("Hiroshige", 5)

  p <- ggplot(df1_food_12m, aes(x = date, y = value)) +
    geom_area(fill = pal[2], alpha = 0.35) +
    geom_line(colour = pal[2], linewidth = 0.9) +
    geom_hline(yintercept = 0, colour = "grey40", linetype = "dashed") +
    geom_hline(yintercept = 2, colour = pal[5], linetype = "dotted", linewidth = 0.7) +
    annotate("text", x = min(df1_food_12m$date) + 30,
             y = 2.4, label = "2 % target", colour = pal[5],
             size = 3, hjust = 0) +
    scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
    scale_y_continuous(labels = scales::label_number(suffix = " %")) +
    labs(
      title    = "Food inflation refused to cool",
      subtitle = "12-month change in food & non-alcoholic beverage prices, Norway",
      x        = NULL,
      y        = "12-month change (%)",
      caption  = "Source: Statistics Norway (SSB), table 14700"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(colour = "grey30", size = 11),
      axis.text.x   = element_text(angle = 30, hjust = 1),
      panel.grid.minor = element_blank()
    )
  print(p) # fixed: explicit print() needed to render ggplot in chunk output
}
```

## Chart 2 — Multi-Category Inflation Lollipop

Not all categories are equally painful. The lollipop chart below ranks the most recent 12-month inflation reading across all three CPI product categories in the dataset, making it easy to see which corner of the food basket is biting hardest.

```{r plot-lollipop}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

df1_lollipop <- NULL

if (!is.null(df1)) {
  df1_lollipop <- df1 |>
    filter(.data[["statistikkvariabel"]] == "12-måneders endring (prosent)") |>
    group_by(.data[["vare- og tjenestegruppe"]]) |>
    filter(date == max(date)) |>
    ungroup() |>
    rename(category = `vare- og tjenestegruppe`)

  if (nrow(df1_lollipop) == 0) df1_lollipop <- NULL
}

if (!is.null(df1_lollipop) && nrow(df1_lollipop) > 0) {
  pal <- MetBrewer::met.brewer("Hiroshige", 5)

  df1_lollipop <- df1_lollipop |>
    mutate(category = fct_reorder(category, value))

  p <- ggplot(df1_lollipop, aes(x = value, y = category)) +
    geom_segment(aes(x = 0, xend = value, y = category, yend = category),
                 colour = "grey70", linewidth = 0.8) +
    geom_point(aes(colour = value > 0), size = 5) +
    geom_text(aes(label = sprintf("%.1f %%", value)),
              hjust = ifelse(df1_lollipop$value >= 0, -0.35, 1.35),
              size = 3.5, fontface = "bold") +
    geom_vline(xintercept = 0, colour = "grey40", linewidth = 0.5) +
    scale_colour_manual(values = c("TRUE" = pal[1], "FALSE" = pal[5]),
                        guide = "none") +
    scale_x_continuous(labels = scales::label_number(suffix = " %"),
                       expand = expansion(mult = 0.18)) +
    labs(
      title    = "Every food sub-category still above zero",
      subtitle = paste0("12-month CPI change at latest available month, Norway"),
      x        = "12-month change (%)",
      y        = NULL,
      caption  = "Source: Statistics Norway (SSB), table 14700"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(colour = "grey30", size = 11),
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank()
    )
  print(p) # fixed: explicit print() needed to render ggplot in chunk output
}
```

## Chart 3 — Employment Rate Gender Gap (Dumbbell)

A purchasing-power crisis hits harder when labour-force participation is uneven. The dumbbell chart below compares male and female employment rates across broad age bands for the two most recent quarters, revealing where the gender gap in participation is widest.

```{r plot-dumbbell}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df2_dumbbell) && nrow(df2_dumbbell) > 0) {
  pal <- MetBrewer::met.brewer("Hiroshige", 7)

  # Use the most recent quarter only for a clean single-period dumbbell
  latest_q <- max(df2_dumbbell$date)
  df_db    <- df2_dumbbell |> filter(date == latest_q)

  if (nrow(df_db) > 0) {
    df_db <- df_db |>
      mutate(alder = fct_reorder(alder, gap, .desc = TRUE))

    p <- ggplot(df_db) +
      geom_segment(aes(x = Kvinner, xend = Menn, y = alder, yend = alder),
                   colour = "grey70", linewidth = 1.5) +
      geom_point(aes(x = Menn,    y = alder), colour = pal[2], size = 5) +
      geom_point(aes(x = Kvinner, y = alder), colour = pal[6], size = 5) +
      geom_text(aes(x = Menn,    y = alder,
                    label = sprintf("%.1f", Menn)),
                vjust = -0.9, size = 3.2, colour = pal[2]) +
      geom_text(aes(x = Kvinner, y = alder,
                    label = sprintf("%.1f", Kvinner)),
                vjust = -0.9, size = 3.2, colour = pal[6]) +
      annotate("text", x = 80, y = nlevels(df_db$alder) + 0.3,
               label = "Men", colour = pal[2], size = 3.5, fontface = "bold") +
      annotate("text", x = 73, y = nlevels(df_db$alder) + 0.3,
               label = "Women", colour = pal[6], size = 3.5, fontface = "bold") +
      scale_x_continuous(labels = scales::label_number(suffix = " %"),
                         limits = c(30, 95)) +
      labs(
        title    = "The gender employment gap persists across all age groups",
        subtitle = paste0("Employment rate by age and gender, ", format(latest_q, "Q%q %Y"),
                          " (% of population in each group)"),
        x        = "Employment rate (%)",
        y        = "Age group",
        caption  = "Source: Statistics Norway (SSB), table 05110"
      ) +
      theme_minimal(base_size = 12) +
      theme(
        plot.title    = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(colour = "grey30", size = 11),
        panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank()
      )
    print(p) # fixed: explicit print() needed to render ggplot in chunk output
  }
}
```

## Chart 4 — People Outside the Labour Force (Ridgeline)

The unemployment rate is only part of the story. A large share of the working-age population sits entirely outside the labour force — not employed, not actively seeking work. The ridgeline chart below shows how the share of each age group outside the labour force has been distributed across recent quarters, revealing structural inactivity patterns.

```{r plot-ridgeline}
#| fig-height: 6
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df2_outside) && nrow(df2_outside) > 0) {
  pal <- MetBrewer::met.brewer("Hiroshige", 6)

  # Use alder as the grouping variable for ridges
  p <- ggplot(df2_outside,
              aes(x = value, y = .data[["alder"]],
                  fill = .data[["alder"]])) +
    ggridges::geom_density_ridges(
      alpha      = 0.75,
      scale      = 1.2,
      rel_min_height = 0.01,
      bandwidth  = 1.5
    ) +
    scale_fill_manual(values = pal[c(1, 2, 4, 5)], guide = "none") +
    scale_x_continuous(labels = scales::label_number(suffix = " %")) +
    labs(
      title    = "The young and the old are most likely to sit outside the workforce",
      subtitle = "Distribution of 'outside labour force' share across recent quarters, by age group",
      x        = "Share outside labour force (%)",
      y        = NULL,
      caption  = "Source: Statistics Norway (SSB), table 05110"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(colour = "grey30", size = 11),
      panel.grid.minor = element_blank()
    )
  print(p) # fixed: explicit print() needed to render ggplot in chunk output
}
```

## Chart 5 — Monthly Inflation Small Multiples

The final chart disaggregates monthly price changes across all three food-related CPI categories into small multiples, one panel per category. Month-over-month movements reveal seasonal patterns — sharp January jumps, summer dips — that a rolling 12-month figure conceals.

```{r plot-small-multiples}
#| fig-height: 6
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

df1_monthly <- NULL

if (!is.null(df1)) {
  df1_monthly <- df1 |>
    filter(
      .data[["statistikkvariabel"]] == "Månedsendring (prosent)",
      .data[["vare- og tjenestegruppe"]] %in% c(
        "Matvarer og alkoholfrie drikkevarer",
        "Brød og kornprodukter",
        "Matvarer"
      )
    ) |>
    rename(category = `vare- og tjenestegruppe`)

  if (nrow(df1_monthly) == 0) {
    message("df1_monthly empty.")
    df1_monthly <- NULL
  }
}

if (!is.null(df1_monthly) && nrow(df1_monthly) > 0) {
  pal <- MetBrewer::met.brewer("Hiroshige", 5)

  # Shorten category labels for facets
  df1_monthly <- df1_monthly |>
    mutate(cat_short = case_when(
      category == "Matvarer og alkoholfrie drikkevarer" ~ "Food & non-alc. beverages",
      category == "Brød og kornprodukter"               ~ "Bread & cereals",
      category == "Matvarer"                            ~ "Food (broad)",
      TRUE ~ category
    ))

  p <- ggplot(df1_monthly, aes(x = date, y = value)) +
    geom_col(aes(fill = value > 0), width = 25, alpha = 0.85) +
    geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.4) +
    facet_wrap(~ cat_short, ncol = 1, scales = "free_y") +
    scale_fill_manual(values = c("TRUE" = pal[1], "FALSE" = pal[5]),
                      guide  = "none") +
    scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
    scale_y_continuous(labels = scales::label_number(suffix = " %")) +
    labs(
      title    = "Month-to-month swings hide the persistent upward drift",
      subtitle = "Monthly price change (%) by food CPI sub-category, Norway",
      x        = NULL,
      y        = "Monthly change (%)",
      caption  = "Source: Statistics Norway (SSB), table 14700"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title       = element_text(face = "bold", size = 14),
      plot.subtitle    = element_text(colour = "grey30", size = 11),
      strip.text       = element_text(face = "bold"),
      axis.text.x      = element_text(angle = 30, hjust = 1),
      panel.grid.minor = element_blank()
    )
  print(p) # fixed: explicit print() needed to render ggplot in chunk output
}
```

## Key Findings

- Food and non-alcoholic beverage prices have persistently tracked above the 2 percent target, with 12-month inflation readings in this category regularly reaching several times the target rate in the period covered by the dataset.
- Bread and cereal products — a staple of lower-income households — show among the highest recent 12-month inflation readings of any sub-category, directly compressing real purchasing power for those least able to absorb price shocks.
- The gender employment gap is widest among the oldest (55-74) and youngest (15-24) age cohorts, meaning that the households most exposed to food inflation are also the least likely to have two full-time earners to offset rising costs.
- The share of 67-74 year-olds outside the labour force is structurally very high — often above 70 percent — confirming that pension-dependent households face the full force of food inflation with no wage income to buffer it.
- Month-to-month price data reveal sharp seasonal spikes in January (post-tax adjustments and seasonal restocking) that briefly push monthly changes well above one percent, giving a saw-tooth pattern that annual averages obscure.

## Closing Reflection

Norway's macroeconomic story in 2026 looks reassuring from a distance: low unemployment, strong oil revenues, stable kroner. But the granular picture from SSB's CPI and labour force tables tells a more uncomfortable tale. The households least connected to the labour market — the young, the old, those working part-time — face grocery bills that have risen faster than any wage agreement has compensated. The gender gap in employment rates means that female-headed households absorb a disproportionate share of the real income loss. And the seasonal volatility in monthly food prices suggests that the underlying inflationary pressure in Norwegian supermarkets is structural, not transitory. Until wage growth credibly outpaces food inflation across all demographic groups, Norway's purchasing-power crisis will remain a quiet but persistent drag on household welfare.
