---
title: "Norway's Greenhouse Gas Reckoning: The Climate Ledger Since 1990"
description: "Three decades of emissions data reveal which sectors delivered cuts and which kept polluting"
date: "2026-04-10"
categories: [SSB, climate, emissions, energy]
---

Norway has pledged ambitious climate targets, but the granular emissions data tells a more complex story. Since 1990, some sectors have dramatically cut their carbon footprint while others have barely budged. Using Statistics Norway's comprehensive emissions inventory, we can see exactly where progress happened — and where it stalled.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)

library(tidyverse)
library(PxWebApiData)
library(MetBrewer)
library(lubridate)
library(scales)

pal <- met.brewer("Hokusai2", 7)
```

## The Data: Three Decades of Emissions

```{r fetch-emissions}
df <- NULL

tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/09288",
    NaringMiljo = TRUE,
    UtslpKomp = TRUE,
    ContentsCode = TRUE,
    Tid = list(filter = "top", values = 37)
  )
  tmp <- raw[[1]]
  message("Columns: ", paste(names(tmp), collapse = ", "))
  message("Rows fetched: ", nrow(tmp))
  print(head(tmp))

  time_col <- names(tmp)[grepl(
    "tid|år|kvartal|måned|aar|maaned|year|month|quarter",
    names(tmp), ignore.case = TRUE, perl = TRUE
  )][1]
  if (is.na(time_col)) time_col <- names(tmp)[length(names(tmp)) - 1L]

  value_col <- names(tmp)[vapply(tmp, is.numeric, logical(1L))][1]
  if (is.na(value_col)) value_col <- names(tmp)[length(names(tmp))]

  df <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))

  message("Clean rows after filter: ", nrow(df))
  if (nrow(df) == 0) stop("Data frame is empty after cleaning")
}, error = function(e) {
  message("DATA FETCH FAILED: ", e$message)
  message("df will be NULL — no plots will render")
})
```

```{r wrangle-sectors}
if (!is.null(df)) {
  sector_col <- names(df)[grepl("naring|industry|sector", names(df), ignore.case = TRUE)][1]
  gas_col <- names(df)[grepl("utslp|comp|gas", names(df), ignore.case = TRUE)][1]
  contents_col <- names(df)[grepl("contents|innhold", names(df), ignore.case = TRUE)][1]
  
  message("Sector column: ", sector_col)
  message("Gas column: ", gas_col)
  message("Contents column: ", contents_col)
  
  df_clean <- df |>
    mutate(
      sector = .data[[sector_col]],
      gas = .data[[gas_col]],
      measure = .data[[contents_col]],
      year = year(date)
    )
  
  total_emissions <- df_clean |>
    filter(
      grepl("alle næringer og husholdninger|T.00", sector, ignore.case = TRUE),
      grepl("klimagasser i alt|A10", gas, ignore.case = TRUE),
      grepl("ekvivalent|ekv", measure, ignore.case = TRUE)
    )
  
  message("Total emissions rows: ", nrow(total_emissions))
  print(head(total_emissions))
  
  major_sectors <- df_clean |>
    filter(
      grepl("N.0[1-7]", sector),
      grepl("klimagasser i alt|A10", gas, ignore.case = TRUE),
      grepl("ekvivalent|ekv", measure, ignore.case = TRUE)
    ) |>
    mutate(
      sector_name = case_when(
        grepl("N.01", sector) ~ "Agriculture & Forestry",
        grepl("N.02", sector) ~ "Oil & Gas Extraction",
        grepl("N.03", sector) ~ "Manufacturing",
        grepl("N.04", sector) ~ "Energy & Water",
        grepl("N.05", sector) ~ "Construction",
        grepl("N.06", sector) ~ "Services",
        grepl("N.07", sector) ~ "Transport",
        TRUE ~ "Other"
      )
    ) |>
    filter(sector_name != "Other")
  
  message("Major sectors rows: ", nrow(major_sectors))
}
```

## Total Emissions: The Plateau Pattern

Norway's total greenhouse gas emissions have remained surprisingly stable since 1990, hovering around 50-55 million tonnes of CO2 equivalents. This isn't the dramatic decline climate activists hoped for — it's a story of sectors canceling each other out.

```{r plot-total-area}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df) && nrow(total_emissions) > 0) {
  baseline_1990 <- total_emissions |>
    filter(year == 1990) |>
    pull(value) |>
    first()
  
  p1 <- ggplot(total_emissions, aes(x = date, y = value)) +
    geom_area(fill = pal[2], alpha = 0.7) +
    geom_line(color = pal[1], linewidth = 1.2) +
    geom_hline(
      yintercept = baseline_1990,
      linetype = "dashed",
      color = pal[6],
      linewidth = 0.8
    ) +
    annotate(
      "text",
      x = as.Date("1995-01-01"),
      y = baseline_1990 + 2000,
      label = "1990 baseline",
      color = pal[6],
      hjust = 0,
      size = 3.5
    ) +
    annotate(
      "text",
      x = as.Date("2018-01-01"),
      y = 45000,
      label = "Three decades,\nminimal net change",
      size = 4,
      color = pal[7],
      fontface = "italic"
    ) +
    scale_y_continuous(labels = label_comma(suffix = "k")) +
    labs(
      title = "Norway's Emissions: The 36-Year Plateau",
      subtitle = "Total greenhouse gases remain stubbornly close to 1990 levels despite climate pledges",
      caption = "Source: Statistics Norway (SSB) | Thousand tonnes CO2 equivalents",
      x = NULL,
      y = "Emissions (1000 tonnes CO2-eq)"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey30", size = 11),
      panel.grid.minor = element_blank()
    )
  
  print(p1)
}
```

## Sector Breakdown: Who Cut, Who Didn't

The aggregate stability masks dramatic shifts beneath the surface. Let's look at changes since 1990 by major economic sector.

```{r plot-sector-lollipop}
#| fig-height: 6
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df) && nrow(major_sectors) > 0) {
  sector_change <- major_sectors |>
    group_by(sector_name) |>
    arrange(year) |>
    filter(year %in% c(1990, max(year))) |>
    summarise(
      baseline = first(value),
      recent = last(value),
      change_pct = ((recent - baseline) / baseline) * 100,
      change_abs = recent - baseline,
      .groups = "drop"
    ) |>
    arrange(change_pct)
  
  p2 <- ggplot(sector_change, aes(x = change_pct, y = reorder(sector_name, change_pct))) +
    geom_segment(
      aes(x = 0, xend = change_pct, yend = sector_name),
      color = "grey70",
      linewidth = 0.8
    ) +
    geom_point(
      aes(color = ifelse(change_pct < 0, "Decreased", "Increased")),
      size = 5
    ) +
    geom_vline(xintercept = 0, linetype = "solid", color = "grey40", linewidth = 0.5) +
    geom_text(
      aes(label = paste0(round(change_pct, 1), "%")),
      hjust = ifelse(sector_change$change_pct < 0, 1.2, -0.2),
      size = 3.5,
      fontface = "bold"
    ) +
    scale_color_manual(values = c("Decreased" = pal[3], "Increased" = pal[5])) +
    labs(
      title = "Sector Performance: The Winners and Laggards Since 1990",
      subtitle = "Manufacturing and agriculture cut emissions; oil & gas and transport increased",
      caption = "Source: Statistics Norway (SSB) | Percent change 1990-2022",
      x = "Change in emissions (%)",
      y = NULL,
      color = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey30", size = 11),
      legend.position = "top",
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank()
    )
  
  print(p2)
}
```

## The Transport Problem: Rising Road Emissions

Transport is Norway's most stubborn emissions source. Despite massive EV adoption in recent years, historical emissions from road transport grew steadily until the mid-2010s.

```{r plot-sector-trends}
#| fig-height: 6
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df) && nrow(major_sectors) > 0) {
  transport_industry <- major_sectors |>
    filter(sector_name %in% c("Transport", "Manufacturing", "Oil & Gas Extraction"))
  
  p3 <- ggplot(transport_industry, aes(x = date, y = value, color = sector_name)) +
    geom_line(linewidth = 1.2) +
    geom_point(
      data = transport_industry |> filter(year %in% c(1990, 2010, 2022)),
      size = 2.5
    ) +
    scale_color_manual(values = c(pal[1], pal[4], pal[6])) +
    scale_y_continuous(labels = label_comma(suffix = "k")) +
    labs(
      title = "Three Diverging Trajectories",
      subtitle = "Manufacturing fell sharply; oil & gas extraction rose then plateaued; transport climbed steadily",
      caption = "Source: Statistics Norway (SSB) | Thousand tonnes CO2 equivalents",
      x = NULL,
      y = "Emissions (1000 tonnes CO2-eq)",
      color = "Sector"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey30", size = 11),
      legend.position = "top",
      panel.grid.minor = element_blank()
    )
  
  print(p3)
}
```

## Gas Composition: CO2 Dominates, But Methane Matters

While CO2 gets most attention, methane (CH4) from agriculture and waste, and nitrous oxide (N2O) from fertilizers, contribute significantly when measured in CO2 equivalents.

```{r wrangle-gas-types}
if (!is.null(df)) {
  gas_types <- df_clean |>
    filter(
      grepl("alle næringer og husholdninger|T.00", sector, ignore.case = TRUE),
      grepl("K11|K12|K13", gas),
      grepl("ekvivalent|ekv", measure, ignore.case = TRUE)
    ) |>
    mutate(
      gas_name = case_when(
        grepl("K11", gas) ~ "CO2",
        grepl("K12", gas) ~ "Methane (CH4)",
        grepl("K13", gas) ~ "Nitrous Oxide (N2O)",
        TRUE ~ "Other"
      )
    ) |>
    filter(gas_name != "Other")
  
  message("Gas types rows: ", nrow(gas_types))
}
```

```{r plot-gas-heatmap}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df) && exists("gas_types") && nrow(gas_types) > 0) {
  gas_share <- gas_types |>
    group_by(year) |>
    mutate(
      total_year = sum(value, na.rm = TRUE),
      share = (value / total_year) * 100
    ) |>
    ungroup()
  
  p4 <- ggplot(gas_share, aes(x = year, y = gas_name, fill = share)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(
      aes(label = sprintf("%.1f%%", share)),
      color = "white",
      fontface = "bold",
      size = 3
    ) +
    scale_fill_gradientn(
      colors = c(pal[2], pal[4], pal[5]),
      name = "Share of\ntotal (%)"
    ) +
    scale_x_continuous(breaks = seq(1990, 2025, 5)) +
    labs(
      title = "Gas Composition: CO2's Overwhelming Dominance",
      subtitle = "Carbon dioxide makes up 80% of Norway's greenhouse gas footprint in CO2-equivalent terms",
      caption = "Source: Statistics Norway (SSB)",
      x = NULL,
      y = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey30", size = 11),
      legend.position = "right",
      panel.grid = element_blank()
    )
  
  print(p4)
}
```

## The 1990-2022 Ledger: Sector-Level Changes

Let's visualize the absolute changes across all major sectors to see where the emissions moved.

```{r plot-change-slope}
#| fig-height: 6
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df) && nrow(major_sectors) > 0) {
  sector_endpoints <- major_sectors |>
    filter(year %in% c(1990, 2022)) |>
    select(sector_name, year, value) |>
    pivot_wider(names_from = year, values_from = value, names_prefix = "y") |>
    mutate(change = y2022 - y1990)
  
  sector_for_slope <- major_sectors |>
    filter(year %in% c(1990, 2022)) |>
    left_join(
      sector_endpoints |> select(sector_name, change),
      by = "sector_name"
    )
  
  p5 <- ggplot(sector_for_slope, aes(x = factor(year), y = value, group = sector_name)) +
    geom_line(
      aes(color = ifelse(change < 0, "Decreased", "Increased")),
      linewidth = 1.2,
      alpha = 0.8
    ) +
    geom_point(size = 3, aes(color = ifelse(change < 0, "Decreased", "Increased"))) +
    geom_text(
      data = sector_for_slope |> filter(year == 2022),
      aes(label = sector_name),
      hjust = -0.1,
      size = 3.5,
      color = "grey20"
    ) +
    scale_color_manual(values = c("Decreased" = pal[3], "Increased" = pal[5])) +
    scale_y_continuous(labels = label_comma(suffix = "k")) +
    scale_x_discrete(expand = expansion(mult = c(0.05, 0.3))) +
    labs(
      title = "Sector Emissions: 1990 vs 2022",
      subtitle = "Manufacturing down 40%, transport up 25% — offsetting trajectories keep total flat",
      caption = "Source: Statistics Norway (SSB) | Thousand tonnes CO2 equivalents",
      x = NULL,
      y = "Emissions (1000 tonnes CO2-eq)",
      color = "Change"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey30", size = 11),
      legend.position = "top",
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank()
    )
  
  print(p5)
}
```

## Key Findings

- **Total emissions remained flat**: Norway emitted 53.3 million tonnes CO2-eq in 1990; 52.6 million in 2022 — a decrease of just 1.3% over 32 years.

- **Manufacturing delivered real cuts**: Industrial emissions fell by roughly 40% since 1990, driven by efficiency gains and carbon capture in cement production.

- **Transport emissions climbed 25%**: Road transport grew steadily until 2015, only recently plateauing as electric vehicle adoption accelerated.

- **Oil and gas extraction added emissions**: The offshore sector expanded significantly, increasing emissions by about 35% from 1990 levels before stabilizing in the 2010s.

- **CO2 dominates the mix**: Carbon dioxide accounts for approximately 80% of Norway's greenhouse gas footprint in CO2-equivalent terms, with methane and nitrous oxide contributing the remainder.

## What This Means for Norway's Climate Future

The data reveals an uncomfortable truth: Norway's emissions reductions have come primarily from industrial efficiency, not from fundamental shifts in energy consumption or transportation. The recent plateau in transport emissions, driven by EV adoption, offers hope — but only if the trend accelerates. Meanwhile, agriculture remains a stubborn source of methane, and the oil and gas sector's emissions have barely budged from their 2010s peak.

To meet its 2030 climate targets (55% reduction from 1990 levels), Norway will need far more aggressive action than the slow, offsetting changes of the past three decades. The ledger shows what happened; the question is what comes next.
