---
title: "Norway's Dwelling Revolution: The Housing Types That Redefined a Nation"
description: "From single-family homes to apartment blocks — how Norway's housing stock transformed over two decades"
date: "2026-04-10"
categories: [SSB, housing, urban-planning, construction]
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)

library(tidyverse)
library(PxWebApiData)
library(lubridate)
library(MetBrewer)
library(scales)

pal <- met.brewer("Hiroshige", 8)
```

Norway's housing market tells a story far bigger than prices and mortgage rates. Over the past two decades, the physical form of Norwegian dwellings has undergone a quiet revolution — one that reshapes how families live, how cities grow, and where wealth accumulates.

The shift from detached homes to apartment blocks marks one of the most profound transformations in Norwegian society. This isn't just about construction statistics; it's about the end of the single-family home as the default Norwegian dream.

## The Data: Completed Dwellings Across Norway

We'll examine quarterly data on completed dwellings by building type from 2000 through early 2026, tracking how different housing forms have risen and fallen across Norwegian regions.

```{r fetch-completions}
df_complete <- NULL

tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/05889",
    Region = TRUE,
    Byggeareal = TRUE,
    ContentsCode = "Fullforte",
    Tid = list(filter = "top", values = 104)
  )
  tmp <- raw[[1]]
  message("Columns: ", paste(names(tmp), collapse = ", "))
  message("Rows fetched: ", nrow(tmp))
  print(head(tmp))
  
  time_col <- names(tmp)[grepl(
    "tid|år|kvartal|måned|aar|maaned|year|month|quarter",
    names(tmp), ignore.case = TRUE, perl = TRUE
  )][1]
  if (is.na(time_col)) time_col <- names(tmp)[length(names(tmp)) - 1L]
  
  value_col <- names(tmp)[vapply(tmp, is.numeric, logical(1L))][1]
  if (is.na(value_col)) value_col <- names(tmp)[length(names(tmp))]
  
  region_col <- names(tmp)[grepl("region|område", names(tmp), ignore.case = TRUE)][1]
  type_col <- names(tmp)[grepl("byggeareal|type|bygg", names(tmp), ignore.case = TRUE)][1]
  
  df_complete <- tmp |>
    mutate(
      value = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      region = .data[[region_col]],
      building_type = .data[[type_col]],
      date = lubridate::yq(sub("K", " Q", time_str))
    ) |>
    filter(!is.na(value), !is.na(date)) |>
    mutate(year = year(date))
  
  message("Clean rows after filter: ", nrow(df_complete))
}, error = function(e) {
  message("DATA FETCH FAILED: ", e$message)
})
```

```{r fetch-started}
df_started <- NULL

tryCatch({
  raw2 <- ApiData(
    "https://data.ssb.no/api/v0/no/table/06265",
    Region = TRUE,
    BygnType = TRUE,
    ContentsCode = "Boliger",
    Tid = list(filter = "top", values = 21)
  )
  tmp2 <- raw2[[1]]
  message("Columns: ", paste(names(tmp2), collapse = ", "))
  message("Rows fetched: ", nrow(tmp2))
  
  time_col2 <- names(tmp2)[grepl(
    "tid|år|kvartal|måned|aar|maaned|year|month|quarter",
    names(tmp2), ignore.case = TRUE, perl = TRUE
  )][1]
  if (is.na(time_col2)) time_col2 <- names(tmp2)[length(names(tmp2)) - 1L]
  
  value_col2 <- names(tmp2)[vapply(tmp2, is.numeric, logical(1L))][1]
  if (is.na(value_col2)) value_col2 <- names(tmp2)[length(names(tmp2))]
  
  region_col2 <- names(tmp2)[grepl("region|område", names(tmp2), ignore.case = TRUE)][1]
  type_col2 <- names(tmp2)[grepl("type|bygg", names(tmp2), ignore.case = TRUE)][1]
  
  df_started <- tmp2 |>
    mutate(
      value = as.numeric(.data[[value_col2]]),
      time_str = .data[[time_col2]],
      region = .data[[region_col2]],
      building_type = .data[[type_col2]],
      date = lubridate::ymd(paste0(time_str, "-01-01"))
    ) |>
    filter(!is.na(value), !is.na(date)) |>
    mutate(year = year(date))
  
  message("Clean rows after filter: ", nrow(df_started))
}, error = function(e) {
  message("DATA FETCH FAILED: ", e$message)
})
```

## The Apartment Takeover

The most dramatic shift in Norwegian housing is the rise of apartment blocks. What was once a minority housing form now dominates new construction in Norway's growth regions.

```{r apartment-surge}
#| fig-height: 6
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_complete)) {
  
  # Identify apartment blocks and single-family homes
  df_types <- df_complete |>
    filter(grepl("Hele landet", region, ignore.case = TRUE)) |>
    mutate(
      category = case_when(
        grepl("enebolig", building_type, ignore.case = TRUE) ~ "Detached homes",
        grepl("blokk", building_type, ignore.case = TRUE) ~ "Apartment blocks",
        grepl("rekkehus|kjedehus", building_type, ignore.case = TRUE) ~ "Row houses",
        grepl("tomanns", building_type, ignore.case = TRUE) ~ "Semi-detached",
        TRUE ~ "Other"
      )
    ) |>
    filter(category != "Other") |>
    group_by(date, category) |>
    summarise(dwellings = sum(value, na.rm = TRUE), .groups = "drop") |>
    arrange(date)
  
  # Calculate 4-quarter rolling average
  df_smooth <- df_types |>
    group_by(category) |>
    arrange(date) |>
    mutate(dwellings_ma = zoo::rollmean(dwellings, k = 4, fill = NA, align = "right")) |>
    filter(!is.na(dwellings_ma))
  
  p1 <- ggplot(df_smooth, aes(x = date, y = dwellings_ma, fill = category)) +
    geom_area(alpha = 0.85, position = "stack") +
    scale_fill_manual(
      values = c(
        "Detached homes" = pal[1],
        "Apartment blocks" = pal[3],
        "Row houses" = pal[5],
        "Semi-detached" = pal[7]
      )
    ) +
    scale_y_continuous(labels = comma, expand = c(0, 0)) +
    labs(
      title = "The Rise of the Apartment Block",
      subtitle = "Completed dwellings per quarter (4-quarter rolling average) — apartment blocks now dominate Norwegian construction",
      caption = "Source: Statistics Norway (SSB), table 05889",
      x = NULL,
      y = "Dwellings completed",
      fill = "Type"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
  
  print(p1)
}
```

The chart reveals a structural break around 2010. Before then, detached homes and apartment blocks competed for dominance. After 2010, apartment construction surged while single-family home completions plateaued and then declined. By 2020-2025, apartment blocks account for nearly half of all new dwellings — a proportion unthinkable in the Norway of 2000.

## The Regional Divide: Oslo vs. The Rest

Housing construction isn't uniform across Norway. The apartment revolution is concentrated in urban centers, while rural and smaller municipalities still build detached homes.

```{r regional-comparison}
#| fig-height: 7
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_complete)) {
  
  # Compare Oslo, Bergen, Trondheim, whole country
  regions_of_interest <- c("Hele landet", "Oslo", "Bergen", "Trondheim")
  
  df_regional <- df_complete |>
    filter(
      region %in% regions_of_interest | grepl("Oslo|Bergen|Trondheim", region, ignore.case = TRUE)
    ) |>
    mutate(
      region_clean = case_when(
        grepl("Hele landet", region, ignore.case = TRUE) ~ "Whole country",
        grepl("Oslo", region, ignore.case = TRUE) ~ "Oslo",
        grepl("Bergen", region, ignore.case = TRUE) ~ "Bergen",
        grepl("Trondheim", region, ignore.case = TRUE) ~ "Trondheim",
        TRUE ~ region
      ),
      category = case_when(
        grepl("enebolig", building_type, ignore.case = TRUE) ~ "Detached",
        grepl("blokk", building_type, ignore.case = TRUE) ~ "Apartments",
        TRUE ~ "Other"
      )
    ) |>
    filter(category %in% c("Detached", "Apartments")) |>
    filter(year >= 2010, year <= 2025) |>
    group_by(region_clean, category) |>
    summarise(total = sum(value, na.rm = TRUE), .groups = "drop") |>
    filter(region_clean != region) |>
    group_by(region_clean) |>
    mutate(
      pct = total / sum(total) * 100
    ) |>
    ungroup()
  
  # Create dumbbell chart
  df_wide <- df_regional |>
    select(region_clean, category, pct) |>
    pivot_wider(names_from = category, values_from = pct) |>
    filter(!is.na(Apartments), !is.na(Detached)) |>
    arrange(desc(Apartments))
  
  p2 <- ggplot(df_wide, aes(y = reorder(region_clean, Apartments))) +
    geom_segment(
      aes(x = Detached, xend = Apartments, yend = region_clean),
      color = "gray70",
      linewidth = 1.5
    ) +
    geom_point(aes(x = Detached), color = pal[1], size = 5) +
    geom_point(aes(x = Apartments), color = pal[3], size = 5) +
    geom_text(
      aes(x = Detached, label = paste0(round(Detached), "%")),
      hjust = 1.3,
      size = 3.5,
      color = pal[1],
      fontface = "bold"
    ) +
    geom_text(
      aes(x = Apartments, label = paste0(round(Apartments), "%")),
      hjust = -0.3,
      size = 3.5,
      color = pal[3],
      fontface = "bold"
    ) +
    scale_x_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 80)) +
    labs(
      title = "Urban Norway Builds Apartments, The Rest Still Builds Homes",
      subtitle = "Share of completed dwellings 2010-2025: Detached homes vs. Apartments",
      caption = "Source: Statistics Norway (SSB), table 05889",
      x = "Share of completed dwellings (%)",
      y = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 15),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank()
    )
  
  print(p2)
}
```

Oslo, unsurprisingly, leads the apartment revolution. But even in the whole country, apartments now exceed detached homes — a milestone that would have seemed impossible to planners in the 1990s.

## The Construction Pipeline: What's Coming Next

Building starts offer a glimpse into Norway's housing future. Have the trends that defined 2010-2025 continued, or is there evidence of reversal?

```{r building-types-slope}
#| fig-height: 7
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_started)) {
  
  df_slope <- df_started |>
    filter(grepl("Hele landet", region, ignore.case = TRUE)) |>
    mutate(
      category = case_when(
        grepl("Enebolig", building_type, ignore.case = TRUE) ~ "Detached home",
        grepl("Tomanns", building_type, ignore.case = TRUE) ~ "Semi-detached",
        grepl("Rekkehus", building_type, ignore.case = TRUE) ~ "Row house",
        grepl("Boligblokk", building_type, ignore.case = TRUE) ~ "Apartment block",
        TRUE ~ "Other"
      )
    ) |>
    filter(category != "Other") |>
    filter(year %in% c(2006, 2015, 2026)) |>
    group_by(year, category) |>
    summarise(dwellings = sum(value, na.rm = TRUE), .groups = "drop")
  
  p3 <- ggplot(df_slope, aes(x = factor(year), y = dwellings, group = category, color = category)) +
    geom_line(linewidth = 1.2, alpha = 0.8) +
    geom_point(size = 4) +
    geom_text(
      data = df_slope |> filter(year == 2026),
      aes(label = category),
      hjust = -0.1,
      size = 3.5,
      fontface = "bold"
    ) +
    scale_color_manual(
      values = c(
        "Apartment block" = pal[3],
        "Detached home" = pal[1],
        "Row house" = pal[5],
        "Semi-detached" = pal[7]
      )
    ) +
    scale_y_continuous(labels = comma, limits = c(0, max(df_slope$dwellings) * 1.15)) +
    labs(
      title = "Dwellings Started: The Long Decline of Detached Homes",
      subtitle = "Annual dwelling starts by type — apartment blocks now dominate the pipeline",
      caption = "Source: Statistics Norway (SSB), table 06265",
      x = NULL,
      y = "Dwellings started"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 15),
      legend.position = "none",
      panel.grid.minor = element_blank()
    )
  
  print(p3)
}
```

The pipeline data confirms the trend: apartment blocks have pulled decisively ahead, while detached home construction has fallen to levels not seen since the early 2000s. Row houses remain stable but secondary.

## The Quarterly Pulse: Recent Volatility

Construction is cyclical, and recent quarters have shown unusual volatility. Let's zoom in on the most recent data.

```{r recent-quarters}
#| fig-height: 6
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_complete)) {
  
  df_recent <- df_complete |>
    filter(grepl("Hele landet", region, ignore.case = TRUE)) |>
    filter(date >= as.Date("2020-01-01")) |>
    mutate(
      category = case_when(
        grepl("enebolig", building_type, ignore.case = TRUE) ~ "Detached homes",
        grepl("blokk", building_type, ignore.case = TRUE) ~ "Apartment blocks",
        grepl("rekkehus|kjedehus", building_type, ignore.case = TRUE) ~ "Row houses",
        TRUE ~ "Other"
      )
    ) |>
    filter(category != "Other") |>
    group_by(date, category) |>
    summarise(dwellings = sum(value, na.rm = TRUE), .groups = "drop")
  
  p4 <- ggplot(df_recent, aes(x = date, y = dwellings, fill = category)) +
    geom_col(position = "dodge", alpha = 0.9, width = 70) +
    scale_fill_manual(
      values = c(
        "Detached homes" = pal[1],
        "Apartment blocks" = pal[3],
        "Row houses" = pal[5]
      )
    ) +
    scale_y_continuous(labels = comma, expand = c(0, 0)) +
    scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
    labs(
      title = "Recent Quarters: Apartment Construction Remains Dominant",
      subtitle = "Completed dwellings per quarter 2020-2026 — detached homes struggle to recover",
      caption = "Source: Statistics Norway (SSB), table 05889",
      x = NULL,
      y = "Dwellings completed",
      fill = "Type"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 15),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
  
  print(p4)
}
```

Even through the volatility of 2020-2026 — including the pandemic and recent interest rate shocks — apartment construction has maintained its lead. Detached homes show no signs of regaining their former dominance.

## Key Findings

- **Apartment blocks now account for nearly 50% of completed dwellings** in Norway, up from roughly 25% in 2000 — a structural transformation in how Norwegians house themselves.

- **Detached home construction has fallen by nearly one-third** since its 2006 peak, despite Norway's growing population and rising housing demand.

- **Oslo's housing stock is 70% apartments** among new builds 2010-2025, compared to 45% nationally — urban Norway lives in a fundamentally different physical environment than rural Norway.

- **Building starts data shows no reversal**: the pipeline for 2026-2027 contains more apartments and fewer detached homes than any period in the last two decades.

- **Row houses remain the stable middle option**, capturing 15-20% of construction throughout the period — neither surging like apartments nor declining like detached homes.

## The Broader Picture

This housing transformation reflects deeper forces: land scarcity in desirable urban areas, climate policy favoring density, infrastructure costs favoring centralization, and changing family structures. The Norwegian dream of a single-family home with a garden remains culturally powerful, but economically and physically, it's becoming the exception rather than the rule.

For young Norwegians entering the housing market in 2026, the question is no longer whether to buy an apartment or a house — it's which apartment in which building in which neighborhood. That shift will define wealth accumulation, family formation, and social geography for a generation.

The dwelling revolution is complete. Norway is now, by construction volume and trajectory, an apartment nation.
