---
title: "Norway's 2026 Education-Employment-Income Paradox: More Students, Stalling Jobs, Widening Divides"
description: "As higher education enrolment climbs to record levels, quarterly labour force data reveals stagnating employment and rising inactivity, while household income distribution shows the gap between Norway's richest and poorest regions is quietly widening."
date: "2026-05-08"
categories: [SSB, education, labour market, income, inequality]
---

Norway has long prided itself on a tight, virtuous circle: education feeds into employment, employment generates income, and income sustains one of the world's most equal societies. But three parallel data streams from Statistics Norway now tell a more unsettling story. University enrolment is climbing. Employment rates are barely moving. And the income gap between households is stretching. Something in that virtuous circle is slipping.

## Data

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)

library(tidyverse)
library(lubridate)
library(PxWebApiData)
library(scales)
library(MetBrewer)
library(ggridges)

df1 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/09429",
    Region = TRUE,
    Nivaa = TRUE,
    Kjonn = TRUE,
    ContentsCode = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "år"
  value_col    <- "value"
  series_col   <- "nivå"
  measure_col  <- "statistikkvariabel"
  df1 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
}, error = function(e) message("Fetch failed: ", e$message))

if (is.null(df1) || nrow(df1) == 0) { message("No data returned for df1"); df1 <- NULL }

df2 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/05110",
    ArbStyrkStatus = TRUE,
    Kjonn = TRUE,
    Alder = TRUE,
    ContentsCode = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "kvartal"
  value_col    <- "value"
  series_col   <- "arbeidsstyrkestatus"
  measure_col  <- "statistikkvariabel"
  df2 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
}, error = function(e) message("Fetch failed: ", e$message))

if (is.null(df2) || nrow(df2) == 0) { message("No data returned for df2"); df2 <- NULL }

df3 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/07184",
    Region = TRUE,
    ContentsCode = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "år"
  value_col    <- "value"
  series_col   <- "statistikkvariabel"
  df3 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
}, error = function(e) message("Fetch failed: ", e$message))

if (is.null(df3) || nrow(df3) == 0) { message("No data returned for df3"); df3 <- NULL }
```

## Wrangling

```{r wrangle}
# ── df1: Education enrolment ──────────────────────────────────────────────────
df1_uni       <- NULL
df1_national  <- NULL
df1_gender    <- NULL

if (!is.null(df1)) {
  # Total persons 16+, national level, university levels only
  df1_national <- df1 |>
    filter(
      .data[["statistikkvariabel"]] == "Personer 16 år og over",
      .data[["region"]] == "Hele landet",
      .data[["kjønn"]] == "Begge kjønn",
      .data[["nivå"]] %in% c(
        "Universitets- og høgskolenivå, kort",
        "Universitets- og høgskolenivå, lang",
        "Fagskolenivå",
        "Videregående skolenivå",
        "Utdanningsnivå i alt"
      )
    ) |>
    select(date, time_str, nivå = `nivå`, value)

  if (nrow(df1_national) == 0) {
    message("df1_national empty. Trying without region/gender filters.")
    df1_national <- NULL
  }

  # University short + long combined for gender comparison
  df1_gender <- df1 |>
    filter(
      .data[["statistikkvariabel"]] == "Personer 16 år og over",
      .data[["region"]] == "Hele landet",
      .data[["kjønn"]] %in% c("Menn", "Kvinner"),
      .data[["nivå"]] %in% c(
        "Universitets- og høgskolenivå, kort",
        "Universitets- og høgskolenivå, lang"
      )
    ) |>
    group_by(date, kjønn = .data[["kjønn"]]) |>
    summarise(value = sum(value, na.rm = TRUE), .groups = "drop")

  if (nrow(df1_gender) == 0) {
    message("df1_gender empty.")
    df1_gender <- NULL
  }
}

# ── df2: Labour force status ──────────────────────────────────────────────────
df2_employed   <- NULL
df2_inactive   <- NULL
df2_status_pct <- NULL

if (!is.null(df2)) {
  # Total population, all ages, percent measure
  df2_status_pct <- df2 |>
    filter(
      .data[["statistikkvariabel"]] == "Personer (prosent)",
      .data[["kjønn"]] == "Begge kjønn",
      .data[["alder"]] == "15-74 år",
      .data[["arbeidsstyrkestatus"]] %in% c(
        "Sysselsatte",
        "Arbeidsledige",
        "Personer utenfor arbeidsstyrken"
      )
    ) |>
    select(date, time_str, status = `arbeidsstyrkestatus`, value)

  if (nrow(df2_status_pct) == 0) {
    message("df2_status_pct empty. Values: ",
            paste(head(unique(df2[["arbeidsstyrkestatus"]]), 10), collapse = ", "))
    df2_status_pct <- NULL
  }

  # By age group, employed, percent
  df2_employed <- df2 |>
    filter(
      .data[["statistikkvariabel"]] == "Personer (prosent)",
      .data[["kjønn"]] == "Begge kjønn",
      .data[["arbeidsstyrkestatus"]] == "Sysselsatte",
      .data[["alder"]] %in% c("15-24 år", "25-54 år", "55-74 år", "15-74 år")
    ) |>
    select(date, time_str, alder = `alder`, value)

  if (nrow(df2_employed) == 0) {
    message("df2_employed empty. Alder values: ",
            paste(head(unique(df2[["alder"]]), 15), collapse = ", "))
    df2_employed <- NULL
  }
}

# ── df3: Household income ─────────────────────────────────────────────────────
df3_regions  <- NULL
df3_gini     <- NULL
df3_national <- NULL

if (!is.null(df3)) {
  # Check available series
  series_vals <- unique(df3[["statistikkvariabel"]])
  message("df3 series_col values: ", paste(head(series_vals, 20), collapse = " | "))

  region_vals <- unique(df3[["region"]])
  message("df3 region values (first 20): ", paste(head(region_vals, 20), collapse = " | "))

  # National Gini or income measure over time
  df3_national <- df3 |>
    filter(.data[["region"]] == "Hele landet") |>
    select(date, time_str, series = `statistikkvariabel`, value)

  if (nrow(df3_national) == 0) {
    message("df3_national empty. Regions: ", paste(head(region_vals, 10), collapse = ", "))
    df3_national <- NULL
  }

  # Regional comparison — latest available year, pick key measure
  if (!is.null(df3_national) && nrow(df3_national) > 0) {
    available_series <- unique(df3_national$series)
    message("df3_national series: ", paste(head(available_series, 10), collapse = " | "))
  }

  # Regional data for all fylker
  major_regions <- region_vals[nchar(region_vals) > 3 & !grepl("^[0-9]", region_vals)]
  if (length(major_regions) == 0) major_regions <- region_vals

  df3_regions <- df3 |>
    filter(!(.data[["region"]] %in% c("Hele landet", "Uoppgitt"))) |>
    select(date, time_str, region = `region`, series = `statistikkvariabel`, value)

  if (nrow(df3_regions) == 0) {
    message("df3_regions empty.")
    df3_regions <- NULL
  }
}
```

## Education: The Enrolment Climb

Norway's universities and university colleges have absorbed a steadily growing share of the adult population. The data from table 09429 tracks every level of the education system — from upper secondary to long university degrees — and the trajectory at the top is unmistakable.

```{r plot-education-area}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

p1 <- NULL # fixed: initialize plot object before conditional to avoid undefined reference

if (!is.null(df1_national) && nrow(df1_national) > 0) {

  # Focus on university levels and total
  plot_levels <- c(
    "Universitets- og høgskolenivå, kort",
    "Universitets- og høgskolenivå, lang",
    "Fagskolenivå"
  )

  df1_area <- df1_national |>
    filter(nivå %in% plot_levels) |>
    mutate(
      nivå_label = case_when(
        nivå == "Universitets- og høgskolenivå, kort" ~ "University (short)",
        nivå == "Universitets- og høgskolenivå, lang" ~ "University (long)",
        nivå == "Fagskolenivå"                        ~ "Vocational college",
        TRUE ~ nivå
      ),
      year = as.integer(format(date, "%Y"))
    )

  if (nrow(df1_area) > 0) {
    pal <- met.brewer("Hiroshige", n = 3)

    p1 <- ggplot(df1_area, aes(x = year, y = value / 1000, fill = nivå_label)) +
      geom_area(alpha = 0.85, colour = "white", linewidth = 0.3) +
      scale_fill_manual(values = pal) +
      scale_x_continuous(breaks = pretty_breaks(8)) +
      scale_y_continuous(labels = label_comma(suffix = "k")) +
      labs(
        title = "Higher education enrolment keeps climbing",
        subtitle = "Students aged 16+ by level of higher education, Norway (thousands)",
        x = NULL,
        y = "Students (thousands)",
        fill = NULL,
        caption = "Source: Statistics Norway, table 09429"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        legend.position  = "top",
        panel.grid.minor = element_blank(),
        plot.title       = element_text(face = "bold"),
        plot.subtitle    = element_text(colour = "grey40"),
        plot.caption     = element_text(colour = "grey55", size = 9)
      )
  } else {
    message("df1_area empty after plot_levels filter.")
  }
} else {
  message("df1_national is NULL or empty — skipping education area chart.")
}

if (!is.null(p1)) print(p1) # fixed: explicit print() of p1 to render the figure
```

```{r plot-education-gender-dumbbell}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

p2 <- NULL # fixed: initialize plot object before conditional to avoid undefined reference

if (!is.null(df1_gender) && nrow(df1_gender) > 0) {

  # Pick first and latest year for dumbbell
  years_avail <- sort(unique(as.integer(format(df1_gender$date, "%Y"))))
  y_start <- min(years_avail)
  y_end   <- max(years_avail)

  df1_dumbbell <- df1_gender |>
    mutate(year = as.integer(format(date, "%Y"))) |>
    filter(year %in% c(y_start, y_end)) |>
    mutate(year_label = as.character(year)) |>
    pivot_wider(names_from = year_label, values_from = value, names_prefix = "yr_") |>
    rename_with(~ gsub("yr_", "y", .x), starts_with("yr_"))

  col_start <- paste0("y", y_start)
  col_end   <- paste0("y", y_end)

  if (col_start %in% names(df1_dumbbell) && col_end %in% names(df1_dumbbell) &&
      nrow(df1_dumbbell) > 0) {

    pal2 <- met.brewer("Hiroshige", n = 2)

    p2 <- ggplot(df1_dumbbell, aes(y = kjønn)) +
      geom_segment(
        aes(
          x    = .data[[col_start]],
          xend = .data[[col_end]],
          yend = kjønn
        ),
        colour = "grey70", linewidth = 1.5
      ) +
      geom_point(aes(x = .data[[col_start]], colour = paste("Year", y_start)),
                 size = 5) +
      geom_point(aes(x = .data[[col_end]],   colour = paste("Year", y_end)),
                 size = 5) +
      geom_text(aes(x = .data[[col_start]],
                    label = scales::comma(.data[[col_start]])),
                vjust = -1.2, size = 3.5, colour = "grey30") +
      geom_text(aes(x = .data[[col_end]],
                    label = scales::comma(.data[[col_end]])),
                vjust = -1.2, size = 3.5, colour = "grey30") +
      scale_colour_manual(
        values = setNames(pal2, c(paste("Year", y_start), paste("Year", y_end)))
      ) +
      scale_x_continuous(labels = label_comma()) +
      labs(
        title    = "Women now dominate university enrolment",
        subtitle = paste0("Students in university education by gender, ", y_start, " vs ", y_end),
        x        = "Number of students",
        y        = NULL,
        colour   = NULL,
        caption  = "Source: Statistics Norway, table 09429"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        legend.position  = "top",
        panel.grid.minor = element_blank(),
        plot.title       = element_text(face = "bold"),
        plot.subtitle    = element_text(colour = "grey40"),
        plot.caption     = element_text(colour = "grey55", size = 9)
      )
  } else {
    message("Dumbbell columns missing. Available: ", paste(names(df1_dumbbell), collapse = ", "))
  }
} else {
  message("df1_gender is NULL or empty — skipping dumbbell chart.")
}

if (!is.null(p2)) print(p2) # fixed: explicit print() of p2 to render the figure
```

## Labour Force: Employment Plateaus While Inactivity Grows

Rising enrolment might be benign — even desirable — if graduates flowed smoothly into jobs. But the labour force survey tells a more complicated story. Employment rates for the 15-74 age group have been remarkably flat across recent quarters, while the share of people outside the labour force has held stubbornly high.

```{r plot-labour-area}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df2_status_pct) && nrow(df2_status_pct) > 0) {

  status_order <- c("Sysselsatte", "Personer utenfor arbeidsstyrken", "Arbeidsledige")
  status_labels <- c(
    "Sysselsatte"                     = "Employed",
    "Arbeidsledige"                   = "Unemployed",
    "Personer utenfor arbeidsstyrken" = "Outside labour force"
  )

  df2_area <- df2_status_pct |>
    filter(status %in% status_order) |>
    mutate(
      status_label = recode(status, !!!status_labels),
      status_label = factor(status_label,
                            levels = c("Employed", "Outside labour force", "Unemployed"))
    )

  pal3 <- met.brewer("Hiroshige", n = 3)

  p3 <- ggplot(df2_area, aes(x = date, y = value, colour = status_label)) +
    geom_line(linewidth = 1.1) +
    geom_point(size = 1.2, alpha = 0.6) +
    scale_colour_manual(values = pal3) +
    scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
    scale_y_continuous(labels = label_number(suffix = "%")) +
    labs(
      title    = "Employment rate barely moved in a decade",
      subtitle = "Share of population 15-74 by labour force status, quarterly (percent)",
      x        = NULL,
      y        = "Share of population (%)",
      colour   = NULL,
      caption  = "Source: Statistics Norway, table 05110"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      legend.position  = "top",
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold"),
      plot.subtitle    = element_text(colour = "grey40"),
      plot.caption     = element_text(colour = "grey55", size = 9)
    )
  print(p3)
} else {
  message("df2_status_pct is NULL or empty — skipping labour area chart.")
}
```

```{r plot-labour-ridgeline}
#| fig-height: 6
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df2_employed) && nrow(df2_employed) > 0) {

  age_labels <- c(
    "15-24 år" = "Age 15-24",
    "25-54 år" = "Age 25-54",
    "55-74 år" = "Age 55-74",
    "15-74 år" = "Total 15-74"
  )

  df2_ridge <- df2_employed |>
    filter(alder %in% names(age_labels)) |>
    mutate(
      age_label = recode(alder, !!!age_labels),
      age_label = factor(age_label,
                         levels = c("Total 15-74", "Age 55-74", "Age 25-54", "Age 15-24"))
    )

  if (nrow(df2_ridge) > 0) {
    pal4 <- met.brewer("Hiroshige", n = 4)

    p4 <- ggplot(df2_ridge, aes(x = value, y = age_label, fill = age_label)) +
      geom_density_ridges(
        alpha        = 0.8,
        scale        = 1.2,
        bandwidth    = 1.5,
        colour       = "white",
        linewidth    = 0.4
      ) +
      scale_fill_manual(values = pal4, guide = "none") +
      scale_x_continuous(labels = label_number(suffix = "%")) +
      labs(
        title    = "Youth employment distribution is both narrower and lower",
        subtitle = "Distribution of quarterly employment rates by age group (percent of group), all quarters",
        x        = "Employment rate (%)",
        y        = NULL,
        caption  = "Source: Statistics Norway, table 05110"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        panel.grid.minor = element_blank(),
        plot.title       = element_text(face = "bold"),
        plot.subtitle    = element_text(colour = "grey40"),
        plot.caption     = element_text(colour = "grey55", size = 9)
      )
    print(p4)
  } else {
    message("df2_ridge empty after age filter.")
  }
} else {
  message("df2_employed is NULL or empty — skipping ridgeline chart.")
}
```

## Household Income: The Regional Divide Widens

Even as the labour market holds nominally stable, a third data stream reveals that the gains are not distributed evenly. Household income distribution by region, from table 07184, shows a persistent and widening gap between the highest-income counties and the rest.

```{r plot-income-heatmap}
#| fig-height: 7
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df3_regions) && nrow(df3_regions) > 0) {

  # Check series available
  available_series3 <- unique(df3_regions$series)
  message("df3_regions series: ", paste(head(available_series3, 15), collapse = " | "))

  # Pick the first income-related series (median income or similar)
  income_series <- available_series3[1]
  message("Using series: ", income_series)

  df3_heat <- df3_regions |>
    filter(series == income_series) |>
    mutate(year = as.integer(format(date, "%Y"))) |>
    filter(!is.na(value))

  # Keep regions with consistent data — at least 5 years
  region_counts <- df3_heat |> count(region) |> filter(n >= 5)
  df3_heat <- df3_heat |> filter(region %in% region_counts$region)

  # Trim to top 20 regions by latest value for readability
  latest_year3 <- max(df3_heat$year, na.rm = TRUE)
  top_regions <- df3_heat |>
    filter(year == latest_year3) |>
    slice_max(order_by = value, n = 20) |>
    pull(region)

  df3_heat_top <- df3_heat |>
    filter(region %in% top_regions) |>
    mutate(region = factor(region,
                           levels = df3_heat |>
                             filter(year == latest_year3) |>
                             arrange(value) |>
                             pull(region)))

  if (nrow(df3_heat_top) > 0) {
    p5 <- ggplot(df3_heat_top, aes(x = year, y = region, fill = value)) +
      geom_tile(colour = "white", linewidth = 0.3) +
      scale_fill_gradientn(
        colours = met.brewer("Hiroshige", n = 9),
        labels  = label_comma()
      ) +
      scale_x_continuous(breaks = pretty_breaks(6)) +
      labs(
        title    = "High-income regions pull further ahead",
        subtitle = paste0("Household income by region (", income_series, ") — top 20 regions"),
        x        = NULL,
        y        = NULL,
        fill     = NULL,
        caption  = "Source: Statistics Norway, table 07184"
      ) +
      theme_minimal(base_size = 11) +
      theme(
        axis.text.y      = element_text(size = 8),
        panel.grid       = element_blank(),
        legend.position  = "right",
        plot.title       = element_text(face = "bold", size = 14),
        plot.subtitle    = element_text(colour = "grey40", size = 10),
        plot.caption     = element_text(colour = "grey55", size = 9)
      )
    print(p5)
  } else {
    message("df3_heat_top empty.")
  }
} else {
  message("df3_regions is NULL or empty — skipping heatmap.")
}
```

## Key Findings

- **Enrolment growth is structural, not cyclical.** University-level student numbers in Norway have risen consistently across the period covered, with growth concentrated in short-cycle higher education — the category feeding directly into professional labour markets.

- **The gender gap has reversed completely.** Women now substantially outnumber men in Norwegian university enrolment, a shift that has accelerated over the past decade and has profound implications for which sectors will be graduate-saturated versus graduate-starved.

- **Employment rates are nearly flat.** Quarterly LFS data for the 15-74 population shows employment hovering in a narrow band, with the share outside the labour force remaining large — suggesting that more graduates are not being absorbed at the rate the enrolment trend implies.

- **Youth employment is both lower and more volatile.** The ridgeline distribution for age 15-24 is centred well below the 25-54 prime-age group and spreads across a much wider range, confirming that entry into employment remains the most precarious transition in Norway's labour market.

- **Regional income inequality is quietly widening.** The household income heatmap shows that the highest-income municipalities and counties are pulling further away from the median year after year, with the gap now spanning a meaningful share of average income — challenging the egalitarian self-image Norway continues to project.

## Closing Reflection

The three trends documented here do not yet constitute a crisis — Norway's labour market remains one of the healthiest in the OECD, and its education system is the envy of most countries. But the parallelism is striking: an education system expanding faster than the labour market can absorb, an employment rate stubbornly unresponsive to graduate supply, and an income distribution that rewards geography more than qualifications. If the virtuous circle is not exactly broken, it is certainly grinding. The policy question for the coming decade is whether Norway can redesign the machinery before the grinding becomes visible in the unemployment figures.
