---
title: "Research Trends Radar"
description: "Bi-weekly arXiv scan across data science, finance, and social sciences — top-5 papers per area, topic timeline, and an interactive mind graph."
date: today
date-format: long
categories: [research, ai-agent, finance]
execute:
  freeze: false
  echo: false
  warning: false
  message: false
format:
  html:
    page-layout: full
    toc: true
---

```{r setup}
library(jsonlite)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(lubridate)
library(ggplot2)
library(visNetwork)
library(igraph)
library(MetBrewer)
library(DT)
library(htmltools)
library(fs)

`%||%` <- function(a, b) if (is.null(a) || length(a) == 0L) b else a

SNAPSHOT_DIR <- "snapshots"

snapshot_files <- if (dir_exists(SNAPSHOT_DIR)) {
  dir_ls(SNAPSHOT_DIR, regexp = "\\d{4}-\\d{2}-\\d{2}\\.json$")
} else character(0)

snapshots <- map(snapshot_files, ~ read_json(.x, simplifyVector = FALSE))
snapshots <- keep(snapshots, ~ !is.null(.x$run_date))
snapshots <- snapshots[order(map_chr(snapshots, "run_date"))]

has_data <- length(snapshots) > 0L &&
            any(map_int(snapshots, ~ length(.x$topics %||% list())) > 0L)

PAL_PRIMARY   <- "#1a1a2e"
PAL_HIGHLIGHT <- "#e94560"
PAL_ACCENT    <- "#0f3460"

BUCKETS <- list(
  data_science    = "Data Science",
  finance         = "Finance",
  social_sciences = "Social Sciences"
)
```

::: {.callout-note appearance="minimal" icon=false}
A simple overview of trending topics in **data science**, **finance**, and **social sciences** — top-5 arXiv papers per area, refreshed every two weeks.
:::

```{r empty-state, eval = !has_data, results = "asis"}
cat("\n## No snapshots yet\n\n")
cat("The first bi-weekly run will populate this page. Trigger it manually with:\n\n")
cat("```bash\nRscript ssrn-research/scan_ssrn.R --dry-run   # Phase A only\nRscript ssrn-research/scan_ssrn.R              # full run (needs ANTHROPIC_API_KEY)\n```\n")
```

```{r prep-data, eval = has_data}
latest <- snapshots[[length(snapshots)]]

# Long table of (run_date, topic_name, paper_ids) across all snapshots
topics_long <- map_dfr(snapshots, function(s) {
  topics <- s$topics %||% list()
  if (length(topics) == 0L) return(tibble())
  tibble(
    run_date  = as.Date(s$run_date),
    name      = map_chr(topics, "name"),
    paper_ids = map(topics, ~ unlist(.x$paper_ids))
  )
})

topics_long <- topics_long |>
  mutate(n_papers = lengths(paper_ids))

render_top_paper <- function(p) {
  authors_txt <- paste(unlist(p$authors %||% list()), collapse = ", ")
  if (nchar(authors_txt) > 200) authors_txt <- paste0(substr(authors_txt, 1, 200), " ...")
  topics_txt  <- paste(unlist(p$topics %||% list()), collapse = " · ")
  cat(sprintf(
    '\n::: {.project-card}\n#### #%s — [%s](%s)\n**%s**\n\n_%s_\n\n%s\n\n[arXiv](%s)\n:::\n',
    p$rank, p$title, p$url,
    authors_txt,
    topics_txt,
    p$why_trending,
    p$url
  ))
}

render_bucket_section <- function(bucket_key, bucket_label) {
  papers <- latest$top_papers[[bucket_key]] %||% list()
  cat(sprintf("\n## %s — top 5\n\n", bucket_label))
  if (length(papers) == 0L) {
    cat(sprintf("_No top-5 picks for %s in the latest snapshot._\n", bucket_label))
    return(invisible())
  }
  for (p in papers) render_top_paper(p)
}
```

## Topic timeline

How often each topic appears across runs. Bigger dots = more papers in that topic that fortnight.

```{r timeline, eval = has_data, fig.width = 11, fig.height = 5.5}
top_n_topics <- 12

topic_totals <- topics_long |>
  group_by(name) |>
  summarise(total = sum(n_papers), .groups = "drop") |>
  arrange(desc(total)) |>
  head(top_n_topics)

plot_df <- topics_long |>
  semi_join(topic_totals, by = "name") |>
  mutate(name = factor(name, levels = topic_totals$name))

n_colors <- nrow(topic_totals)
palette  <- if (n_colors <= 12) MetBrewer::met.brewer("Cassatt2", n_colors) else
            colorRampPalette(MetBrewer::met.brewer("Cassatt2", 12))(n_colors)

ggplot(plot_df, aes(x = run_date, y = name, size = n_papers, color = name)) +
  geom_point(alpha = 0.85) +
  scale_color_manual(values = as.character(palette), guide = "none") +
  scale_size_area(max_size = 11, name = "papers") +
  scale_x_date(date_labels = "%d %b %Y") +
  labs(x = NULL, y = NULL,
       title = paste0("Top ", n_colors, " topics across ", length(snapshots), " run(s)")) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "grey92"),
    plot.title         = element_text(face = "bold", color = PAL_PRIMARY),
    axis.text.y        = element_text(color = PAL_PRIMARY)
  )
```

```{r top-data-science, eval = has_data, results = "asis"}
render_bucket_section("data_science", BUCKETS$data_science)
```

```{r top-finance, eval = has_data, results = "asis"}
render_bucket_section("finance", BUCKETS$finance)
```

```{r top-social-sciences, eval = has_data, results = "asis"}
render_bucket_section("social_sciences", BUCKETS$social_sciences)
```

## Mind graph

An interactive force-directed view of the topic landscape. Drag nodes, zoom, hover for details. Colors come from a Louvain community-detection pass on the co-occurrence graph.

```{r mindgraph, eval = has_data}
topic_papers <- topics_long |>
  unnest(paper_ids) |>
  rename(paper_id = paper_ids) |>
  distinct(name, paper_id) |>
  group_by(name) |>
  summarise(papers = list(paper_id),
            n_papers_total = n(),
            .groups = "drop")

# Always start with a 3-column shell so empty results keep their schema
empty_edges <- tibble(from = character(), to = character(), weight = integer())

edges <- if (nrow(topic_papers) >= 2) {
  pairs <- combn(topic_papers$name, 2, simplify = FALSE)
  result <- map_dfr(pairs, function(pr) {
    a <- topic_papers$papers[[which(topic_papers$name == pr[1])]]
    b <- topic_papers$papers[[which(topic_papers$name == pr[2])]]
    w <- length(intersect(a, b))
    if (w >= 1) tibble(from = pr[1], to = pr[2], weight = w) else NULL
  })
  if (is.null(result) || nrow(result) == 0L) empty_edges else result
} else empty_edges

if (nrow(edges) > 0L) {
  g <- igraph::graph_from_data_frame(
    edges, directed = FALSE,
    vertices = topic_papers |> select(name, n_papers_total)
  )
  comm <- igraph::cluster_louvain(g)
  groups <- tibble(name = igraph::V(g)$name,
                   group = as.integer(igraph::membership(comm)))
} else {
  groups <- tibble(name = topic_papers$name,
                   group = seq_len(nrow(topic_papers)))
}

n_groups <- max(groups$group, 1L)
group_palette <- if (n_groups <= 12) MetBrewer::met.brewer("Cassatt2", max(n_groups, 2)) else
                 colorRampPalette(MetBrewer::met.brewer("Cassatt2", 12))(n_groups)

nodes <- topic_papers |>
  left_join(groups, by = "name") |>
  mutate(
    id    = name,
    label = name,
    value = n_papers_total,
    color = as.character(group_palette[group]),
    title = sprintf("<b>%s</b><br>%d papers across all runs", name, n_papers_total)
  ) |>
  select(id, label, value, color, title, group)

vis_edges <- edges |>
  mutate(
    value = weight,
    title = sprintf("%d shared paper(s)", weight)
  )

if (nrow(nodes) == 0L) {
  htmltools::tags$div(class = "alert alert-info",
                      "Not enough topics yet for a mind graph.")
} else {
  visNetwork(nodes, vis_edges, height = "640px", width = "100%") |>
    visNodes(font = list(face = "Inter", size = 16),
             shape = "dot",
             borderWidth = 2,
             color = list(border = "#1a1a2e", highlight = list(border = "#e94560"))) |>
    visEdges(smooth = list(enabled = TRUE, type = "continuous"),
             color = list(color = "#bbbbbb", highlight = "#e94560")) |>
    visPhysics(solver = "forceAtlas2Based",
               forceAtlas2Based = list(gravitationalConstant = -55,
                                        centralGravity        = 0.012,
                                        springLength          = 120,
                                        springConstant        = 0.06,
                                        avoidOverlap          = 0.6),
               stabilization = list(iterations = 220)) |>
    visInteraction(hover = TRUE, navigationButtons = TRUE,
                   tooltipDelay = 100, dragNodes = TRUE) |>
    visOptions(highlightNearest = list(enabled = TRUE, degree = 1, hover = TRUE),
               nodesIdSelection = TRUE)
}
```

## Methodology & cadence

- **Cadence** — runs on the 1st and 15th of each month at 08:00 UTC via GitHub Actions (`.github/workflows/ssrn-research.yml`). Manual triggers via `workflow_dispatch`.
- **Phase A (deterministic)** — three arXiv queries, one per bucket, each returning the most-recent 80 candidates filtered to the last 14 days.
- **Phase B (Claude Haiku tool-use)** — picks the top-5 per bucket and clusters all candidates into 5–10 named topics. Edges are *not* emitted by the model — they're computed from `paper_ids` intersections to guarantee consistency.
- **Snapshots** — one JSON per run in `ssrn-research/snapshots/`. Page reads them all on every render.

```{r run-summary, eval = has_data, results = "asis"}
last_n <- min(8, length(snapshots))
recent <- tail(snapshots, last_n)
rows <- map_chr(recent, function(s) {
  bc <- s$bucket_counts %||% list()
  bc_str <- if (length(bc) > 0L)
    paste(names(bc), unlist(bc), sep = "=", collapse = ", ")
  else "n/a"
  sprintf("- **%s** — %d candidates (%s), %d topics",
          s$run_date,
          s$candidates_count %||% 0L,
          bc_str,
          length(s$topics %||% list()))
})
cat("**Last ", last_n, " runs:**\n\n", paste(rows, collapse = "\n"), sep = "")
```
