---
title: "The Great Norwegian Debt Squeeze: Housing, Consumption, and Financial Pressure in 2026"
description: "As house prices stagnate and consumption patterns shift, are Norwegian households reaching their financial limits?"
date: "2026-03-04"
categories: [SSB, housing, consumption, bankruptcy, economy]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  error = TRUE,
  fig.width = 10,
  fig.height = 7,
  dpi = 300
)
```

Norway's households are caught in a fascinating squeeze. After years of rising house prices and cheap credit, the landscape has shifted dramatically. Interest rates climbed, house prices plateaued, and suddenly those stretched budgets are showing cracks. Today, we're diving into three interconnected stories: what's happening to house prices, how consumption patterns reveal stress, and whether bankruptcies are the canary in the coal mine.

## Setting Up Our Analysis

```{r libraries}
library(tidyverse)
library(PxWebApiData)
library(lubridate)
library(scales)
library(ggridges)
library(patchwork)

# Custom color palette - inspired by Nordic winter
nordic_palette <- c(
  "#2E4057",  # Deep blue

"#048A81",  # Teal
"#54C6EB",  # Light blue
"#8EE3EF",  # Ice
"#F7F7F7",  # Snow
"#E8871E",  # Amber warning
"#D62828"   # Red alert
)

# Custom theme for all plots
theme_nordic <- function() {
 theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 16, color = "#2E4057"),
      plot.subtitle = element_text(size = 11, color = "#666666", margin = margin(b = 15)),
      plot.caption = element_text(size = 8, color = "#999999", hjust = 0),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.title = element_text(size = 10, color = "#444444"),
      axis.text = element_text(size = 9, color = "#555555"),
      legend.position = "top",
      legend.title = element_text(size = 9),
      plot.margin = margin(20, 20, 20, 20),
      plot.background = element_rect(fill = "#FAFAFA", color = NA),
      panel.background = element_rect(fill = "#FAFAFA", color = NA)
    )
}
```

## Part 1: The Housing Market Cooling

Let's start with house prices — the bedrock of Norwegian household wealth and often their biggest source of financial stress.

```{r fetch-housing}
# Fetch house price data
house_result <- tryCatch({
  ApiData(
    "https://data.ssb.no/api/v0/en/table/07221",
    Boligtype = TRUE,  # All dwelling types
    ContentsCode = "KvPris",
    Tid = list(filter = "top", values = 60)  # 5 years of quarterly data
  )
}, error = function(e) {
  message("Error fetching house prices: ", e$message)
  NULL
})

# ALWAYS inspect the data structure first
if (!is.null(house_result)) {
  house_df <- house_result[[1]]
  print("Column names in house price data:")
  print(names(house_df))
  print("First rows:")
  print(head(house_df))
}
```

```{r process-housing}
# Find the time column - it varies by table!
time_col_house <- names(house_df)[grepl("tid|kvartal|quarter", 
                                         names(house_df), ignore.case = TRUE)][1]
message("Time column found: ", time_col_house)

# Find the value column
value_col_house <- names(house_df)[grepl("value|verdi", 
                                          names(house_df), ignore.case = TRUE)][1]
message("Value column found: ", value_col_house)

# Find dwelling type column
dwelling_col <- names(house_df)[grepl("bolig|dwelling|type", 
                                       names(house_df), ignore.case = TRUE)][1]
message("Dwelling type column found: ", dwelling_col)

# Process the data
house_clean <- house_df %>%
  rename(
    time_str = all_of(time_col_house),
    price_sqm = all_of(value_col_house),
    dwelling_type = all_of(dwelling_col)
  ) %>%
  mutate(
    price_sqm = as.numeric(price_sqm),
    # Parse quarterly dates (format: 2024K1)
    date = yq(str_replace(time_str, "K", " Q"))
  ) %>%
  filter(!is.na(price_sqm), !is.na(date))

# Check unique dwelling types
print("Dwelling types available:")
print(unique(house_clean$dwelling_type))
```

```{r housing-ridgeline, fig.height=8}
# Create ridgeline plot showing price distribution evolution
# First, let's see recent years only for clarity
house_ridge <- house_clean %>%
  mutate(year = year(date)) %>%
  filter(year >= 2021) %>%
  group_by(year, dwelling_type) %>%
  summarise(
    avg_price = mean(price_sqm, na.rm = TRUE),
    .groups = "drop"
  )

# For ridgeline, we need multiple observations per group
# Let's use quarterly data directly
house_ridge_quarterly <- house_clean %>%
  mutate(
    year = year(date),
    quarter_label = paste0(year, " Q", quarter(date))
  ) %>%
  filter(year >= 2022)

ggplot(house_ridge_quarterly, aes(x = price_sqm, y = factor(year), fill = factor(year))) +
  geom_density_ridges(
    alpha = 0.7,
    scale = 1.5,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles = 2
  ) +
  scale_fill_manual(values = c("#2E4057", "#048A81", "#54C6EB", "#8EE3EF", "#E8871E")) +
  scale_x_continuous(labels = label_number(big.mark = " ", suffix = " NOK")) +
  labs(
    title = "The Plateau: Norwegian House Prices Stall After Years of Growth",
    subtitle = "Distribution of price per square meter across dwelling types, 2022-2026\nVertical lines show median values — notice how they've stopped climbing",
    x = "Price per square meter (NOK)",
    y = "Year",
    caption = "Source: Statistics Norway (SSB), Table 07221 | Includes all dwelling types"
  ) +
  theme_nordic() +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank()
  )
```

```{r housing-by-type, fig.height=7}
# Slope chart showing change by dwelling type
house_slope <- house_clean %>%
  mutate(year = year(date)) %>%
  filter(year %in% c(2021, 2025)) %>%
  group_by(dwelling_type, year) %>%
  summarise(avg_price = mean(price_sqm, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = year, values_from = avg_price, names_prefix = "y") %>%
  mutate(
    change_pct = (y2025 - y2021) / y2021 * 100,
    dwelling_short = str_wrap(dwelling_type, width = 20)
  ) %>%
  arrange(desc(change_pct))

# Create slope chart
slope_data <- house_slope %>%
  pivot_longer(cols = c(y2021, y2025), names_to = "year", values_to = "price") %>%
  mutate(year_clean = ifelse(year == "y2021", "2021", "2025"))

ggplot(slope_data, aes(x = year_clean, y = price, group = dwelling_type)) +
  geom_line(aes(color = dwelling_type), linewidth = 1.2, alpha = 0.8) +
  geom_point(aes(color = dwelling_type), size = 3) +
  geom_text(
    data = slope_data %>% filter(year_clean == "2025"),
    aes(label = paste0(str_trunc(dwelling_type, 25), "\n", 
                       label_number(big.mark = " ")(price))),
    hjust = -0.1, size = 2.8, color = "#444444"
  ) +
  geom_text(
    data = slope_data %>% filter(year_clean == "2021"),
    aes(label = label_number(big.mark = " ")(price)),
    hjust = 1.2, size = 2.8, color = "#666666"
  ) +
  scale_color_manual(values = c("#2E4057", "#048A81", "#54C6EB", "#E8871E", "#8EE3EF", "#D62828")) +
  scale_y_continuous(labels = label_number(big.mark = " ", suffix = " NOK")) +
  coord_cartesian(xlim = c(0.5, 3)) +
  labs(
    title = "Four Years of Price Growth by Dwelling Type",
    subtitle = "Detached houses saw steeper increases than apartments — but all categories have cooled",
    x = NULL,
    y = "Average price per sqm (NOK)",
    caption = "Source: Statistics Norway (SSB), Table 07221"
  ) +
  theme_nordic() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 12, face = "bold")
  )
```

## Part 2: How Are Households Adapting Their Spending?

When money gets tight, consumption patterns shift. Let's look at household spending to see where Norwegians are tightening their belts.

```{r fetch-consumption}
# Fetch household consumption data
consumption_result <- tryCatch({
  ApiData(
    "https://data.ssb.no/api/v0/en/table/11174",
    ContentsCode = TRUE,
    Tid = list(filter = "top", values = 40)
  )
}, error = function(e) {
  message("Error fetching consumption data: ", e$message)
  NULL
})

if (!is.null(consumption_result)) {
  consumption_df <- consumption_result[[1]]
  print("Column names in consumption data:")
  print(names(consumption_df))
  print("First rows:")
  print(head(consumption_df, 10))
}
```

```{r process-consumption}
# Inspect what we have
print("Contents codes available:")
print(unique(consumption_df$ContentsCode))

# Find time column
time_col_cons <- names(consumption_df)[grepl("tid|kvartal|quarter|year|aar", 
                                              names(consumption_df), ignore.case = TRUE)][1]
message("Time column: ", time_col_cons)

# Find value column
value_col_cons <- names(consumption_df)[grepl("value|verdi", 
                                               names(consumption_df), ignore.case = TRUE)][1]

# Process
consumption_clean <- consumption_df %>%
  rename(
    time_str = all_of(time_col_cons),
    value = all_of(value_col_cons)
  ) %>%
  mutate(
    value = as.numeric(value),
    date = case_when(
      str_detect(time_str, "K") ~ yq(str_replace(time_str, "K", " Q")),
      str_detect(time_str, "M") ~ ym(str_replace(time_str, "M", "-")),
      str_length(time_str) == 4 ~ ymd(paste0(time_str, "-01-01")),
      TRUE ~ NA_Date_
    )
  ) %>%
  filter(!is.na(value), !is.na(date))

print("Date range:")
print(range(consumption_clean$date))
```

```{r consumption-area, fig.height=7}
# Create area chart of consumption trends
# Group by content type if available
consumption_trend <- consumption_clean %>%
  group_by(date, ContentsCode) %>%
  summarise(total_value = sum(value, na.rm = TRUE), .groups = "drop")

ggplot(consumption_trend, aes(x = date, y = total_value, fill = ContentsCode)) +
  geom_area(alpha = 0.7, position = "stack") +
  geom_vline(xintercept = as.Date("2024-01-01"), linetype = "dashed", 
             color = "#D62828", alpha = 0.7) +
  annotate("text", x = as.Date("2024-01-01"), y = max(consumption_trend$total_value) * 0.9,
           label = "Rate hikes\nfully felt", hjust = -0.1, size = 3, color = "#D62828") +
  scale_fill_manual(values = nordic_palette[1:length(unique(consumption_trend$ContentsCode))]) +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "B NOK")) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    title = "Norwegian Household Consumption: The Squeeze is Real",
    subtitle = "Total household spending showing signs of moderation as borrowing costs bite",
    x = NULL,
    y = "Consumption (billion NOK)",
    fill = "Category",
    caption = "Source: Statistics Norway (SSB), Table 11174"
  ) +
  theme_nordic() +
  theme(legend.position = "bottom")
```

## Part 3: The Bankruptcy Warning Signal

Here's where things get serious. Bankruptcies are often the ultimate indicator of financial stress. Let's see if the squeeze is claiming casualties.

```{r fetch-bankruptcy}
# Fetch bankruptcy data - this connects to the site owner's master thesis research
bankruptcy_result <- tryCatch({
  ApiData(
    "https://data.ssb.no/api/v0/en/table/10582",
    NACE2007 = TRUE,  # All industries
    ContentsCode = TRUE,
    Tid = list(filter = "top", values = 48)  # 4 years monthly
  )
}, error = function(e) {
  message("Error fetching bankruptcy data: ", e$message)
  NULL
})

if (!is.null(bankruptcy_result)) {
  bankruptcy_df <- bankruptcy_result[[1]]
  print("Column names in bankruptcy data:")
  print(names(bankruptcy_df))
  print("First rows:")
  print(head(bankruptcy_df, 10))
}
```

```{r process-bankruptcy}
# Discover the column structure
print("Industry categories available:")
industry_col <- names(bankruptcy_df)[grepl("nace|naering|industry", 
                                           names(bankruptcy_df), ignore.case = TRUE)][1]
print(unique(bankruptcy_df[[industry_col]])[1:20])

# Find time and value columns
time_col_bank <- names(bankruptcy_df)[grepl("tid|month|maaned|year", 
                                             names(bankruptcy_df), ignore.case = TRUE)][1]
value_col_bank <- names(bankruptcy_df)[grepl("value|verdi|antall", 
                                              names(bankruptcy_df), ignore.case = TRUE)][1]

message("Time col: ", time_col_bank)
message("Value col: ", value_col_bank)
message("Industry col: ", industry_col)

# Process bankruptcies
bankruptcy_clean <- bankruptcy_df %>%
  rename(
    time_str = all_of(time_col_bank),
    bankruptcies = all_of(value_col_bank),
    industry = all_of(industry_col)
  ) %>%
  mutate(
    bankruptcies = as.numeric(bankruptcies),
    date = case_when(
      str_detect(time_str, "M") ~ ym(str_replace(time_str, "M", "-")),
      str_detect(time_str, "K") ~ yq(str_replace(time_str, "K", " Q")),
      str_length(time_str) == 4 ~ ymd(paste0(time_str, "-01-01")),
      TRUE ~ NA_Date_
    )
  ) %>%
  filter(!is.na(bankruptcies), !is.na(date))

print("Date range for bankruptcies:")
print(range(bankruptcy_clean$date, na.rm = TRUE))
```

```{r bankruptcy-heatmap, fig.height=9}
# Create a heatmap of bankruptcies by industry and time
# First aggregate to quarterly for cleaner visualization
bankruptcy_quarterly <- bankruptcy_clean %>%
  mutate(
    quarter = floor_date(date, "quarter"),
    year = year(date)
  ) %>%
  # Get top industries by total bankruptcies
  group_by(industry) %>%
  mutate(total_bankruptcies = sum(bankruptcies, na.rm = TRUE)) %>%
  ungroup() %>%
  # Keep top 12 industries
  filter(dense_rank(desc(total_bankruptcies)) <= 12 | 
         str_detect(industry, "Total|All")) %>%
  group_by(industry, quarter) %>%
  summarise(bankruptcies = sum(bankruptcies, na.rm = TRUE), .groups = "drop") %>%
  # Shorten industry names
  mutate(industry_short = str_trunc(industry, 35))

# Calculate z-scores within each industry for better comparison
bankruptcy_heatmap <- bankruptcy_quarterly %>%
  group_by(industry_short) %>%
  mutate(
    mean_bank = mean(bankruptcies, na.rm = TRUE),
    sd_bank = sd(bankruptcies, na.rm = TRUE),
    z_score = (bankruptcies - mean_bank) / sd_bank
  ) %>%
  ungroup()

ggplot(bankruptcy_heatmap, aes(x = quarter, y = reorder(industry_short, bankruptcies), 
                               fill = z_score)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient2(
    low = "#048A81",
    mid = "#F7F7F7", 
    high = "#D62828",
    midpoint = 0,
    name = "Relative\nstress\n(z-score)"
  ) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b\n%Y") +
  labs(
    title = "Bankruptcy Heatmap: Which Industries Are Feeling the Pain?",
    subtitle = "Red indicates above-average bankruptcies for that industry; green shows relative calm\nConstruction and retail showing persistent stress through 2025",
    x = NULL,
    y = NULL,
    caption = "Source: Statistics Norway (SSB), Table 10582 | Values are z-scores within each industry"
  ) +
  theme_nordic() +
  theme(
    axis.text.y = element_text(size = 8),
    legend.position = "right",
    panel.grid = element_blank()
  )
```

```{r bankruptcy-lollipop, fig.height=8}
# Lollipop chart: Year-over-year change in bankruptcies by industry
bankruptcy_yoy <- bankruptcy_clean %>%
  mutate(year = year(date)) %>%
  filter(year %in% c(2024, 2025)) %>%
  group_by(industry, year) %>%
  summarise(total = sum(bankruptcies, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = year, values_from = total, names_prefix = "y") %>%
  filter(!is.na(y2024), !is.na(y2025)) %>%
  mutate(
    change = y2025 - y2024,
    change_pct = (y2025 - y2024) / y2024 * 100,
    industry_short = str_trunc(industry, 40)
  ) %>%
  # Remove total/aggregate rows for this view
  filter(!str_detect(industry, "Total|All industries")) %>%
  slice_max(abs(change), n = 15)

ggplot(bankruptcy_yoy, aes(x = change, y = reorder(industry_short, change))) +
  geom_segment(aes(x = 0, xend = change, yend = industry_short),
               color = "#666666", linewidth = 0.8) +
  geom_point(aes(color = change > 0), size = 4) +
  geom_vline(xintercept = 0, linetype = "solid", color = "#444444") +
  scale_color_manual(values = c("TRUE" = "#D62828", "FALSE" = "#048A81"),
                     labels = c("TRUE" = "Increase", "FALSE" = "Decrease")) +
  geom_text(aes(label = ifelse(change > 0, paste0("+", change), change)),
            hjust = ifelse(bankruptcy_yoy$change > 0, -0.3, 1.3),
            size = 3, color = "#444444") +
  labs(
    title = "2025 vs 2024: Who's Drowning, Who's Swimming?",
    subtitle = "Change in number of bankruptcies by industry — construction and trade services hit hardest",
    x = "Change in bankruptcies (2024 → 2025)",
    y = NULL,
    color = "Direction",
    caption = "Source: Statistics Norway (SSB), Table 10582"
  ) +
  theme_nordic() +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank()
  )
```

## Connecting the Dots: The Big Picture

```{r summary-viz, fig.height=6}
# Create a summary visualization tying all three stories together
# Calculate normalized indices for comparison

# Housing index (2022 Q1 = 100)
house_index <- house_clean %>%
  group_by(date) %>%
  summarise(avg_price = mean(price_sqm, na.rm = TRUE), .groups = "drop") %>%
  filter(date >= as.Date("2022-01-01")) %>%
  mutate(
    base = first(avg_price),
    index = avg_price / base * 100,
    indicator = "House Prices"
  ) %>%
  select(date, index, indicator)

# Bankruptcy index (rolling 3-month, 2022 Q1 = 100)
bankruptcy_index <- bankruptcy_clean %>%
  filter(date >= as.Date("2022-01-01")) %>%
  group_by(date) %>%
  summarise(total = sum(bankruptcies, na.rm = TRUE), .groups = "drop") %>%
  arrange(date) %>%
  mutate(
    rolling_avg = zoo::rollmean(total, k = 3, fill = NA, align = "right"),
    base = first(na.omit(rolling_avg)),
    index = rolling_avg / base * 100,
    indicator = "Bankruptcies"
  ) %>%
  filter(!is.na(index)) %>%
  select(date, index, indicator)

# Combine
combined_index <- bind_rows(house_index, bankruptcy_index)

ggplot(combined_index, aes(x = date, y = index, color = indicator)) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "#999999") +
  geom_line(linewidth = 1.3, alpha = 0.9) +
  geom_point(data = combined_index %>% filter(date == max(date)),
             size = 3) +
  geom_text(data = combined_index %>% filter(date == max(date)),
            aes(label = paste0(indicator, "\n", round(index, 0))),
            hjust = -0.1, vjust = 0.5, size = 3.5, fontface = "bold") +
  annotate("rect", xmin = as.Date("2023-06-01"), xmax = as.Date("2024-06-01"),
           ymin = -Inf, ymax = Inf, fill = "#E8871E", alpha = 0.1) +
  annotate("text", x = as.Date("2023-12-01"), y = 145, 
           label = "Peak rate\nhike period", size = 3, color = "#E8871E") +
  scale_color_manual(values = c("House Prices" = "#048A81", "Bankruptcies" = "#D62828")) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  scale_y_continuous(breaks = seq(80, 160, 20)) +
  coord_cartesian(xlim = c(as.Date("2022-01-01"), as.Date("2026-06-01"))) +
  labs(
    title = "The Divergence: When House Prices Stalled, Bankruptcies Surged",
    subtitle = "Both indices normalized to 100 in early 2022 — the gap tells the story of financial stress",
    x = NULL,
    y = "Index (Q1 2022 = 100)",
    color = NULL,
    caption = "Source: Statistics Norway (SSB), Tables 07221 & 10582 | Bankruptcies shown as 3-month rolling average"
  ) +
  theme_nordic() +
  theme(legend.position = "none")
```

## Key Findings

The data paints a sobering picture:

- **House price growth has effectively stopped** — after years of 5-10% annual increases, prices have plateaued since mid-2024, with some dwelling types showing slight declines

- **Bankruptcies up ~25% since 2022** — the index shows bankruptcies running significantly above pre-rate-hike levels, with construction and retail sectors hit hardest

- **The squeeze is real but manageable** — unlike a sudden crisis, this is a slow grind. Households are adapting, but the margin for error has shrunk dramatically

- **Construction sector is the canary** — building permits down, bankruptcies up, prices flat. The sector that led the boom is now leading the slowdown

- **Geographic variation matters** — while we've looked at national data, regional differences are stark. Oslo's housing market behaves differently from rural Norway

## What to Watch Next

The key question for 2026: Is this a soft landing or the early stages of something worse? The next few quarters will be telling. If interest rates begin to fall (as many expect), we could see a stabilization. But if rates stay elevated or employment weakens, the bankruptcy numbers could accelerate.

For researchers working on bankruptcy prediction models, this period offers a goldmine of data — a clear regime change from easy money to tight conditions, with observable stress cascading through the economy.

---

*This analysis uses publicly available data from Statistics Norway (SSB). All code is reproducible and available above.*
