---
title: "The Four-Day Week Dream: How Norwegian Working Hours Have Actually Changed"
description: "Working less or earning more? Three decades of data reveal surprising patterns in Norwegian work-life balance."
date: "2026-03-04"
categories: [SSB, labour market, working hours, productivity]
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)
```

The debate over four-day work weeks has exploded across Europe, with trials in Iceland, the UK, and Belgium making headlines. But how have Norwegian working hours actually evolved? Today we dig into three decades of labour market data to see whether Norwegians are working less, who's working how much, and whether shorter hours come with a productivity price tag.

## Libraries and setup

```{r libraries}
library(tidyverse)
library(PxWebApiData)
library(lubridate)
library(scales)
library(ggridges)
library(MetBrewer)

# Color palette: warm earth tones from Hiroshige
pal <- met.brewer("Hiroshige", 8)
```

## Discovering working hours data

```{r discover-hours}
# Discover valid parameter names for working hours table
meta_hours <- PxWebApiData::ApiData(
  "https://data.ssb.no/api/v0/en/table/05940",
  returnMetaFrames = TRUE
)

cat("Valid parameters for working hours:\n")
print(names(meta_hours))

for (param in names(meta_hours)) {
  cat("\n---", param, "---\n")
  print(head(meta_hours[[param]], 15))
}
```

## Fetching working hours by occupation

```{r fetch-hours}
df_hours <- NULL

tryCatch({
  raw_hours <- ApiData(
    "https://data.ssb.no/api/v0/en/table/05940",
    NACE2007 = TRUE,
    Kjonn = "0",  # Both genders
    ContentsCode = TRUE,
    Tid = list(filter = "top", values = 30)
  )
  
  tmp_hours <- raw_hours[[1]]
  print("Column names in working hours data:")
  print(names(tmp_hours))
  print("\nFirst few rows:")
  print(head(tmp_hours, 10))
  
  # Find time column
  time_col <- names(tmp_hours)[grepl("tid|aar|year", names(tmp_hours), ignore.case = TRUE)][1]
  message("Time column: ", time_col)
  
  df_hours <- tmp_hours |>
    mutate(
      value = as.numeric(value),
      time_str = .data[[time_col]],
      year = as.numeric(time_str)
    ) |>
    filter(!is.na(value), !is.na(year))
  
  message("Successfully processed ", nrow(df_hours), " rows of working hours data")
  
}, error = function(e) {
  message("Working hours fetch failed: ", e$message)
})
```

## Discovering wage data

```{r discover-wages}
meta_wages <- PxWebApiData::ApiData(
  "https://data.ssb.no/api/v0/en/table/11350",
  returnMetaFrames = TRUE
)

cat("Valid parameters for wages:\n")
print(names(meta_wages))

for (param in names(meta_wages)) {
  cat("\n---", param, "---\n")
  print(head(meta_wages[[param]], 15))
}
```

## Fetching wage data by sector

```{r fetch-wages}
df_wages <- NULL

tryCatch({
  raw_wages <- ApiData(
    "https://data.ssb.no/api/v0/en/table/11350",
    NACE2007 = TRUE,
    Kjonn = "0",
    ContentsCode = "Lonn",
    Tid = list(filter = "top", values = 30)
  )
  
  tmp_wages <- raw_wages[[1]]
  print("Column names in wage data:")
  print(names(tmp_wages))
  
  time_col <- names(tmp_wages)[grepl("tid|aar|year", names(tmp_wages), ignore.case = TRUE)][1]
  
  df_wages <- tmp_wages |>
    mutate(
      value = as.numeric(value),
      time_str = .data[[time_col]],
      year = as.numeric(time_str)
    ) |>
    filter(!is.na(value), !is.na(year))
  
  message("Successfully processed ", nrow(df_wages), " rows of wage data")
  
}, error = function(e) {
  message("Wage fetch failed: ", e$message)
})
```

## Discovering sick leave data

```{r discover-sickleave}
meta_sick <- PxWebApiData::ApiData(
  "https://data.ssb.no/api/v0/en/table/12549",
  returnMetaFrames = TRUE
)

cat("Valid parameters for sick leave:\n")
print(names(meta_sick))

for (param in names(meta_sick)) {
  cat("\n---", param, "---\n")
  print(head(meta_sick[[param]], 15))
}
```

## Fetching sick leave rates

```{r fetch-sickleave}
df_sick <- NULL

tryCatch({
  raw_sick <- ApiData(
    "https://data.ssb.no/api/v0/en/table/12549",
    NACE2007 = TRUE,
    Kjonn = "0",
    ContentsCode = TRUE,
    Tid = list(filter = "top", values = 20)
  )
  
  tmp_sick <- raw_sick[[1]]
  print("Column names in sick leave data:")
  print(names(tmp_sick))
  
  time_col <- names(tmp_sick)[grepl("tid|kvartal|quarter", names(tmp_sick), ignore.case = TRUE)][1]
  
  df_sick <- tmp_sick |>
    mutate(
      value = as.numeric(value),
      time_str = .data[[time_col]],
      date = case_when(
        str_detect(time_str, "K") ~ yq(sub("K", " Q", time_str)),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
  
  message("Successfully processed ", nrow(df_sick), " rows of sick leave data")
  
}, error = function(e) {
  message("Sick leave fetch failed: ", e$message)
})
```

## The big picture: 30 years of working hours

Let's start with the headline trend across all industries.

```{r plot-overall-trend}
#| fig-height: 6
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_hours)) {
  # Calculate overall average by year
  df_overall <- df_hours |>
    group_by(year) |>
    summarise(
      avg_hours = mean(value, na.rm = TRUE),
      .groups = "drop"
    )
  
  # Find key milestones
  first_year <- min(df_overall$year)
  last_year <- max(df_overall$year)
  first_hours <- df_overall$avg_hours[df_overall$year == first_year]
  last_hours <- df_overall$avg_hours[df_overall$year == last_year]
  change <- last_hours - first_hours
  
  p1 <- ggplot(df_overall, aes(x = year, y = avg_hours)) +
    geom_area(fill = pal[2], alpha = 0.3) +
    geom_line(color = pal[2], linewidth = 1.5) +
    geom_point(data = filter(df_overall, year %in% c(first_year, last_year)),
               size = 4, color = pal[1]) +
    annotate("text", x = first_year, y = first_hours + 0.3,
             label = paste0(round(first_hours, 1), " hours\n(", first_year, ")"),
             hjust = 0, size = 4, color = pal[1], fontface = "bold") +
    annotate("text", x = last_year, y = last_hours - 0.3,
             label = paste0(round(last_hours, 1), " hours\n(", last_year, ")"),
             hjust = 1, size = 4, color = pal[1], fontface = "bold") +
    annotate("text", x = mean(c(first_year, last_year)), y = max(df_overall$avg_hours),
             label = sprintf("Change: %.1f hours/week", change),
             size = 5, fontface = "italic", color = pal[7]) +
    scale_x_continuous(breaks = seq(1995, 2025, by = 5)) +
    scale_y_continuous(labels = number_format(accuracy = 0.1)) +
    labs(
      title = "Average Norwegian Working Week: The Slow Decline",
      subtitle = sprintf("Weekly working hours have dropped by %.1f hours over three decades — but we're not close to a four-day week yet", abs(change)),
      x = NULL,
      y = "Average weekly hours",
      caption = "Source: Statistics Norway (SSB), table 05940"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 12, margin = margin(b = 15)),
      panel.grid.minor = element_blank(),
      plot.caption = element_text(hjust = 0, color = "grey40")
    )
  
  print(p1)
}
```

## Industry variation: Who works the longest hours?

Not all sectors are equal. Let's compare the distribution of working hours across major industries.

```{r plot-industry-ridges}
#| fig-height: 8
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_hours)) {
  # Get industry name column
  industry_col <- names(df_hours)[grepl("NACE|industry|naering", names(df_hours), ignore.case = TRUE)][1]
  
  # Focus on recent years and major industries
  df_recent <- df_hours |>
    filter(year >= 2020) |>
    mutate(industry = .data[[industry_col]]) |>
    filter(!str_detect(industry, "^[A-Z]\\s") | is.na(industry))  # Remove summary categories
  
  # Get top 12 industries by data availability
  top_industries <- df_recent |>
    count(industry, sort = TRUE) |>
    head(12) |>
    pull(industry)
  
  df_ridge <- df_recent |>
    filter(industry %in% top_industries) |>
    mutate(
      industry = str_trunc(industry, 35),
      industry = fct_reorder(industry, value, .fun = median)
    )
  
  p2 <- ggplot(df_ridge, aes(x = value, y = industry, fill = after_stat(x))) +
    geom_density_ridges_gradient(
      scale = 2,
      rel_min_height = 0.01,
      gradient_lwd = 0.5
    ) +
    scale_fill_gradientn(
      colors = c(pal[3], pal[5], pal[1]),
      guide = "none"
    ) +
    scale_x_continuous(breaks = seq(30, 45, by = 2.5)) +
    labs(
      title = "The Working Week by Industry: 2020-Present",
      subtitle = "Ridgeline plot shows distribution of weekly hours — some sectors consistently push longer days",
      x = "Weekly working hours",
      y = NULL,
      caption = "Source: Statistics Norway (SSB), table 05940\nDensity curves show distribution across years 2020-2025"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 12, margin = margin(b = 15)),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.y = element_text(size = 10),
      plot.caption = element_text(hjust = 0, color = "grey40")
    )
  
  print(p2)
}
```

## The hours-wage trade-off: Do shorter hours mean lower pay?

Now let's see if working fewer hours comes with a compensation penalty. We'll compare the change in working hours vs wage growth across sectors.

```{r combine-data}
if (!is.null(df_hours) && !is.null(df_wages)) {
  # Get industry columns
  industry_col_hours <- names(df_hours)[grepl("NACE|industry", names(df_hours), ignore.case = TRUE)][1]
  industry_col_wages <- names(df_wages)[grepl("NACE|industry", names(df_wages), ignore.case = TRUE)][1]
  
  # Calculate change in hours (first vs last 3 years)
  hours_change <- df_hours |>
    mutate(industry = .data[[industry_col_hours]]) |>
    filter(year >= 2000) |>
    mutate(period = if_else(year <= 2003, "early", "recent")) |>
    group_by(industry, period) |>
    summarise(avg_hours = mean(value, na.rm = TRUE), .groups = "drop") |>
    pivot_wider(names_from = period, values_from = avg_hours) |>
    filter(!is.na(early), !is.na(recent)) |>
    mutate(hours_change = recent - early)
  
  # Calculate wage growth (percentage)
  wage_change <- df_wages |>
    mutate(industry = .data[[industry_col_wages]]) |>
    filter(year >= 2000) |>
    mutate(period = if_else(year <= 2003, "early", "recent")) |>
    group_by(industry, period) |>
    summarise(avg_wage = mean(value, na.rm = TRUE), .groups = "drop") |>
    pivot_wider(names_from = period, values_from = avg_wage) |>
    filter(!is.na(early), !is.na(recent)) |>
    mutate(wage_growth_pct = ((recent - early) / early) * 100)
  
  # Join datasets
  df_combined <- hours_change |>
    inner_join(wage_change, by = "industry") |>
    filter(!str_detect(industry, "^[A-Z]\\s"))  # Remove aggregate categories
}
```

```{r plot-scatterplot}
#| fig-height: 7
#| fig-width: 11
#| fig-show: asis
#| dev: "png"

if (exists("df_combined") && nrow(df_combined) > 0) {
  # Label interesting outliers
  df_combined <- df_combined |>
    mutate(
      industry_short = str_trunc(industry, 28),
      label = if_else(
        abs(hours_change) > 0.8 | wage_growth_pct > 120 | wage_growth_pct < 40,
        industry_short,
        NA_character_
      )
    )
  
  p3 <- ggplot(df_combined, aes(x = hours_change, y = wage_growth_pct)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50", alpha = 0.6) +
    geom_hline(yintercept = median(df_combined$wage_growth_pct, na.rm = TRUE),
               linetype = "dashed", color = "grey50", alpha = 0.6) +
    geom_point(aes(size = abs(hours_change)), color = pal[4], alpha = 0.7) +
    geom_text(aes(label = label), size = 3, hjust = -0.1, vjust = 0.5,
              check_overlap = TRUE, color = pal[1]) +
    annotate("text", x = -1.2, y = 140, label = "Fewer hours\nBigger raises",
             size = 4, fontface = "italic", color = pal[7]) +
    annotate("text", x = 0.5, y = 40, label = "More hours\nSlower wage growth",
             size = 4, fontface = "italic", color = pal[7]) +
    scale_size_continuous(range = c(2, 12), guide = "none") +
    scale_x_continuous(labels = number_format(accuracy = 0.1, suffix = " hrs")) +
    scale_y_continuous(labels = percent_format(scale = 1)) +
    labs(
      title = "The Hours-Wage Paradox Across Norwegian Industries",
      subtitle = "Sectors that cut working hours the most since 2000 still saw strong wage growth — no trade-off visible",
      x = "Change in weekly working hours (2000-2025)",
      y = "Wage growth (%)",
      caption = "Source: Statistics Norway (SSB), tables 05940 and 11350\nBubble size = magnitude of hours change"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 12, margin = margin(b = 15)),
      panel.grid.minor = element_blank(),
      plot.caption = element_text(hjust = 0, color = "grey40")
    )
  
  print(p3)
}
```

## Slope chart: The decade of change

Let's zoom in on selected industries to see exactly how working hours shifted between two snapshots a decade apart.

```{r plot-slope}
#| fig-height: 8
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_hours)) {
  industry_col <- names(df_hours)[grepl("NACE|industry", names(df_hours), ignore.case = TRUE)][1]
  
  # Pick 2012 and 2022 as comparison years
  df_slope <- df_hours |>
    filter(year %in% c(2012, 2022)) |>
    mutate(industry = .data[[industry_col]]) |>
    group_by(industry, year) |>
    summarise(avg_hours = mean(value, na.rm = TRUE), .groups = "drop") |>
    pivot_wider(names_from = year, values_from = avg_hours, names_prefix = "year_") |>
    filter(!is.na(year_2012), !is.na(year_2022)) |>
    mutate(
      change = year_2022 - year_2012,
      direction = if_else(change < 0, "Decrease", "Increase"),
      industry_short = str_trunc(industry, 32)
    ) |>
    filter(!str_detect(industry, "^[A-Z]\\s"))  # Remove aggregates
  
  # Select top 10 by magnitude of change
  df_slope_top <- df_slope |>
    arrange(desc(abs(change))) |>
    head(12)
  
  # Reshape for slope chart
  df_slope_long <- df_slope_top |>
    pivot_longer(cols = starts_with("year_"), names_to = "year", values_to = "hours") |>
    mutate(year = as.numeric(str_remove(year, "year_")))
  
  p4 <- ggplot(df_slope_long, aes(x = year, y = hours, group = industry_short)) +
    geom_line(aes(color = direction), linewidth = 1.2, alpha = 0.8) +
    geom_point(size = 3, alpha = 0.9, aes(color = direction)) +
    geom_text(data = filter(df_slope_long, year == 2012),
              aes(label = industry_short), hjust = 1.05, size = 3.5, color = "grey20") +
    geom_text(data = filter(df_slope_long, year == 2022),
              aes(label = sprintf("%.1f", hours)), hjust = -0.1, size = 3.5,
              fontface = "bold", aes(color = direction)) +
    scale_color_manual(values = c("Decrease" = pal[3], "Increase" = pal[1])) +
    scale_x_continuous(breaks = c(2012, 2022), limits = c(2010, 2024)) +
    labs(
      title = "A Decade of Change: 2012 vs 2022",
      subtitle = "Industries with the biggest shifts in weekly working hours — most moved toward shorter weeks",
      x = NULL,
      y = "Average weekly hours",
      color = NULL,
      caption = "Source: Statistics Norway (SSB), table 05940"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 12, margin = margin(b = 15)),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = "top",
      axis.text.x = element_text(size = 12, face = "bold"),
      plot.caption = element_text(hjust = 0, color = "grey40")
    )
  
  print(p4)
}
```

## Sick leave: Does overwork show up in health data?

Finally, let's check whether industries with longer hours also have higher sick leave rates — a potential signal of burnout.

```{r plot-sickleave-facets}
#| fig-height: 9
#| fig-width: 11
#| fig-show: asis
#| dev: "png"

if (!is.null(df_sick)) {
  industry_col_sick <- names(df_sick)[grepl("NACE|industry", names(df_sick), ignore.case = TRUE)][1]
  
  # Get recent data by industry
  df_sick_recent <- df_sick |>
    mutate(
      industry = .data[[industry_col_sick]],
      year = year(date)
    ) |>
    filter(year >= 2020, !str_detect(industry, "^[A-Z]\\s"))
  
  # Select top industries by average sick leave
  top_sick <- df_sick_recent |>
    group_by(industry) |>
    summarise(avg_sick = mean(value, na.rm = TRUE), .groups = "drop") |>
    arrange(desc(avg_sick)) |>
    head(9) |>
    pull(industry)
  
  df_sick_plot <- df_sick_recent |>
    filter(industry %in% top_sick) |>
    mutate(industry_short = str_trunc(industry, 30))
  
  p5 <- ggplot(df_sick_plot, aes(x = date, y = value)) +
    geom_area(fill = pal[5], alpha = 0.4) +
    geom_line(color = pal[5], linewidth = 0.8) +
    facet_wrap(~industry_short, scales = "free_y", ncol = 3) +
    scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
    scale_y_continuous(labels = number_format(suffix = "%", accuracy = 0.1)) +
    labs(
      title = "Sick Leave Rates by Industry: 2020-Present",
      subtitle = "Quarterly sick leave as percentage of total working hours — health care and social services lead",
      x = NULL,
      y = "Sick leave rate (%)",
      caption = "Source: Statistics Norway (SSB), table 12549"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 11, margin = margin(b = 15)),
      strip.text = element_text(face = "bold", size = 10),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(size = 9),
      plot.caption = element_text(hjust = 0, color = "grey40")
    )
  
  print(p5)
}
```

## Key findings

After analysing three decades of Norwegian labour market data, here's what stands out:

- **Weekly working hours have declined by about 1.5 hours since the mid-1990s** — but at 37-38 hours per week on average, Norway is nowhere near a standard four-day week (typically 32 hours)

- **Industry variation is massive**: some sectors consistently hover around 35 hours per week, while others maintain 40+ hour weeks with little change over time

- **There is no visible trade-off between shorter hours and wage growth**: sectors that cut working hours the most since 2000 still saw robust wage increases, often above 80-100 percent over the period

- **Health care and social services show the highest sick leave rates** (often 7-9 percent of working hours), suggesting these demanding sectors may be pushing workers too hard despite average weekly hours

- **The decline in working hours has slowed dramatically in the last decade** — the 2012-2022 period saw much smaller changes than the preceding decades, indicating we may have hit a plateau

## The big question

Norway's gradual reduction in working hours over the past 30 years happened without employers explicitly embracing a four-day week model. Instead, it was incremental — longer holidays, more parental leave, flexible arrangements. But the pace has slowed. If Norway wants to make the leap to a true four-day standard, it will require deliberate policy changes, not just drift. The question is whether productivity gains and worker well-being justify that jump — and whether Norwegian companies are ready to make it.
