---
title: "Grand Slam Point-by-Point Analysis"
description: "Serve patterns, rally length, break point pressure, distance run, and long-point fatigue across all four Grand Slams (2011–2020)"
date: "2026-03-12"
categories: [Tennis, R, Data Analysis]
execute:
  warning: false
  message: false
  freeze: false
---

```{r setup}
#| echo: false
library(tidyverse)
library(scales)
library(gt)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(colour = "grey40", size = 12),
      panel.grid.minor = element_blank(),
      legend.position  = "bottom"
    )
)

slam_cols <- c(
  "ausopen"    = "#0057A8",
  "frenchopen" = "#C8593C",
  "usopen"     = "#3D5A99",
  "wimbledon"  = "#27AE60"
)
slam_labels <- c(
  "ausopen"    = "Australian Open",
  "frenchopen" = "French Open",
  "usopen"     = "US Open",
  "wimbledon"  = "Wimbledon"
)
```

## Data

Point-by-point data from [Jeff Sackmann's Grand Slam repository](https://github.com/JeffSackmann/tennis_slam_pointbypoint). Covers matches on courts with the Hawkeye tracking system — typically from the second week onward plus select early-round showcourts. Australian Open runs 2011–2020; the other three slams run 2011–2019.

```{r load-data}
data_dir <- here::here("data-projects", "data")

slams     <- c("ausopen", "frenchopen", "usopen", "wimbledon")
ao_years  <- 2011:2020
other_years <- 2011:2019

slam_years <- bind_rows(
  tibble(slam = "ausopen",    year = ao_years),
  tibble(slam = "frenchopen", year = other_years),
  tibble(slam = "usopen",     year = other_years),
  tibble(slam = "wimbledon",  year = other_years)
)

base_url <- "https://raw.githubusercontent.com/JeffSackmann/tennis_slam_pointbypoint/master/"

# ── Matches ──────────────────────────────────────────────────────────────────
matches_cache <- file.path(data_dir, "slam_matches.csv")

if (!file.exists(matches_cache)) {
  matches_raw <- pmap_dfr(slam_years, function(slam, year) {
    url <- paste0(base_url, year, "-", slam, "-matches.csv")
    tryCatch(
      read_csv(url, show_col_types = FALSE, col_types = cols(.default = "c")) |>
        mutate(slam = slam, year = year),
      error = \(e) tibble()
    )
  })
  write_csv(matches_raw, matches_cache)
} else {
  matches_raw <- read_csv(matches_cache, show_col_types = FALSE)
}

# ── Points ────────────────────────────────────────────────────────────────────
points_cache <- file.path(data_dir, "slam_points.csv")

keep_cols <- c(
  "match_id", "SetNo", "GameNo", "PointNumber", "PointServer",
  "P1GamesWon", "P2GamesWon", "P1PointsWon", "P2PointsWon",
  "SetWinner", "GameWinner", "PointWinner",
  "P1Score", "P2Score",
  "Speed_KMH", "ServeWidth", "ServeIndicator", "ServeDepth", "ReturnDepth",
  "RallyCount",
  "P1Ace", "P2Ace", "P1Winner", "P2Winner",
  "P1DoubleFault", "P2DoubleFault",
  "P1UnfErr", "P2UnfErr",
  "P1NetPoint", "P2NetPoint",
  "P1BreakPoint", "P2BreakPoint",
  "P1BreakPointWon", "P2BreakPointWon",
  "P1BreakPointMissed", "P2BreakPointMissed",
  "P1DistanceRun", "P2DistanceRun"
)

if (!file.exists(points_cache)) {
  points_raw <- pmap_dfr(slam_years, function(slam, year) {
    url <- paste0(base_url, year, "-", slam, "-points.csv")
    tryCatch({
      df <- read_csv(url, show_col_types = FALSE, col_types = cols(.default = "c"))
      # keep only columns that exist in this file
      df[intersect(keep_cols, names(df))] |>
        mutate(slam = slam, year = as.integer(year))
    }, error = \(e) tibble())
  })
  write_csv(points_raw, points_cache)
} else {
  points_raw <- read_csv(points_cache, show_col_types = FALSE)
}

# ── Clean & type-cast ─────────────────────────────────────────────────────────
pts <- points_raw |>
  mutate(
    across(c(year, SetNo, GameNo, PointNumber, PointServer,
             PointWinner, GameWinner, RallyCount,
             P1Ace, P2Ace, P1Winner, P2Winner,
             P1DoubleFault, P2DoubleFault,
             P1UnfErr, P2UnfErr,
             P1BreakPoint, P2BreakPoint,
             P1BreakPointWon, P2BreakPointWon),
           \(x) suppressWarnings(as.integer(x))),
    across(c(Speed_KMH, P1DistanceRun, P2DistanceRun),
           \(x) suppressWarnings(as.numeric(x))),
    slam_label = slam_labels[slam]
  )

matches <- matches_raw |>
  mutate(year = as.integer(year))
```


## Coverage overview

Not every match is tracked — Hawkeye data is only available for select courts. This heatmap shows how many matches have point data per slam per year.

```{r coverage}
#| fig-height: 4
#| fig-width: 9

pts |>
  distinct(match_id, slam, year) |>
  count(slam, year) |>
  mutate(slam_label = slam_labels[slam]) |>
  ggplot(aes(factor(year), slam_label, fill = n)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = n), size = 3.5, colour = "white", fontface = "bold") +
  scale_fill_gradient(low = "#9ECAE1", high = "#08519C") +
  labs(
    title = "Matches with point-by-point data",
    x = NULL, y = NULL, fill = "Matches"
  ) +
  theme(legend.position = "right", axis.text.x = element_text(angle = 45, hjust = 1))
```


## Rally length by slam

Clay courts at Roland Garros produce the longest rallies; grass at Wimbledon the shortest. The distributions below pool all points by slam.

```{r rally-dist}
#| fig-height: 5
#| fig-width: 11

pts |>
  filter(!is.na(RallyCount), RallyCount >= 1, RallyCount <= 40) |>
  mutate(slam_label = fct_reorder(slam_label, RallyCount, median, .desc = TRUE)) |>
  ggplot(aes(RallyCount, fill = slam)) +
  geom_histogram(binwidth = 1, colour = NA, alpha = 0.85) +
  scale_fill_manual(values = slam_cols, labels = slam_labels) +
  facet_wrap(~slam_label, scales = "free_y") +
  labs(
    title = "Rally length distribution by slam",
    subtitle = "All tracked points, 2011–2020",
    x = "Rally length (shots)", y = "Points", fill = NULL
  ) +
  theme(legend.position = "none")
```

```{r rally-trend}
#| fig-height: 5
#| fig-width: 11

pts |>
  filter(!is.na(RallyCount), RallyCount >= 1) |>
  group_by(slam, year) |>
  summarise(
    median_rally = median(RallyCount),
    mean_rally   = mean(RallyCount),
    .groups = "drop"
  ) |>
  mutate(slam_label = slam_labels[slam]) |>
  ggplot(aes(year, median_rally, colour = slam)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_colour_manual(values = slam_cols, labels = slam_labels) +
  labs(
    title = "Median rally length over time by slam",
    x = NULL, y = "Median rally length (shots)", colour = NULL
  )
```


## Serve speed

First serves are substantially faster than second serves at every slam. Hard-court slams (AO, USO) tend to produce slightly higher serve speeds than clay or grass.

```{r serve-speed-dist}
#| fig-height: 5
#| fig-width: 11

pts |>
  filter(!is.na(Speed_KMH), Speed_KMH > 100, Speed_KMH < 260,
         ServeIndicator %in% c("1", "2")) |>
  mutate(
    serve = if_else(ServeIndicator == "1", "1st serve", "2nd serve"),
    slam_label = slam_labels[slam]
  ) |>
  ggplot(aes(Speed_KMH, fill = slam)) +
  geom_density(alpha = 0.65, colour = NA) +
  scale_fill_manual(values = slam_cols, labels = slam_labels) +
  facet_grid(serve ~ slam_label) +
  labs(
    title = "Serve speed distribution by slam and serve number",
    x = "Speed (km/h)", y = "Density", fill = NULL
  ) +
  theme(legend.position = "none")
```

```{r serve-speed-trend}
#| fig-height: 5
#| fig-width: 11

pts |>
  filter(!is.na(Speed_KMH), Speed_KMH > 100, Speed_KMH < 260,
         ServeIndicator %in% c("1", "2")) |>
  mutate(serve = if_else(ServeIndicator == "1", "1st serve", "2nd serve")) |>
  group_by(slam, year, serve) |>
  summarise(median_speed = median(Speed_KMH), .groups = "drop") |>
  mutate(slam_label = slam_labels[slam]) |>
  ggplot(aes(year, median_speed, colour = slam, linetype = serve)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_colour_manual(values = slam_cols, labels = slam_labels) +
  labs(
    title = "Median serve speed over time",
    x = NULL, y = "Median speed (km/h)", colour = NULL, linetype = NULL
  )
```


## Serve direction under pressure

Does a server change their target when facing a break point? `ServeWidth` encodes: **W** (wide), **B/BC/BW** (body), **C** (centre/T).

```{r serve-direction}
#| fig-height: 6
#| fig-width: 11

pts |>
  filter(!is.na(ServeWidth), ServeWidth != "", ServeIndicator == "1") |>
  mutate(
    direction = case_when(
      ServeWidth %in% c("W", "BW") ~ "Wide",
      ServeWidth %in% c("B", "BC") ~ "Body",
      ServeWidth == "C"            ~ "Centre/T",
      TRUE                         ~ NA_character_
    ),
    pressure = case_when(
      PointServer == 1 & P1BreakPoint == 1 ~ "Break point against server",
      PointServer == 2 & P2BreakPoint == 1 ~ "Break point against server",
      TRUE                                  ~ "Normal point"
    )
  ) |>
  filter(!is.na(direction)) |>
  count(slam, pressure, direction) |>
  group_by(slam, pressure) |>
  mutate(pct = n / sum(n), slam_label = slam_labels[slam]) |>
  ungroup() |>
  ggplot(aes(direction, pct, fill = pressure)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Normal point" = "#2171B5",
                                "Break point against server" = "#CB4335")) +
  scale_y_continuous(labels = percent) +
  facet_wrap(~slam_label) +
  labs(
    title = "1st serve direction: normal vs break point",
    subtitle = "Does pressure shift the target?",
    x = NULL, y = "Share of 1st serves", fill = NULL
  )
```


## Break point conversion

```{r bp-conversion}
#| fig-height: 5
#| fig-width: 11

bp_data <- pts |>
  filter(!is.na(P1BreakPoint), !is.na(P2BreakPoint)) |>
  mutate(
    bp_opp  = P1BreakPoint + P2BreakPoint,
    bp_won  = P1BreakPointWon + P2BreakPointWon
  ) |>
  filter(bp_opp == 1) |>
  group_by(slam, year) |>
  summarise(
    conversion = mean(bp_won == 1, na.rm = TRUE),
    n_bp = n(),
    .groups = "drop"
  ) |>
  filter(n_bp >= 20) |>
  mutate(slam_label = slam_labels[slam])

bp_data |>
  ggplot(aes(year, conversion, colour = slam)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_colour_manual(values = slam_cols, labels = slam_labels) +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Break point conversion rate over time",
    subtitle = "Share of break point opportunities converted",
    x = NULL, y = "Conversion rate", colour = NULL
  )
```

```{r bp-by-slam}
#| fig-height: 4
#| fig-width: 9

bp_data |>
  group_by(slam, slam_label) |>
  summarise(
    avg_conversion = mean(conversion),
    .groups = "drop"
  ) |>
  mutate(slam_label = fct_reorder(slam_label, avg_conversion)) |>
  ggplot(aes(avg_conversion, slam_label, fill = slam)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = slam_cols) +
  scale_x_continuous(labels = percent) +
  labs(
    title = "Average break point conversion by slam",
    x = "Conversion rate", y = NULL
  ) +
  theme(legend.position = "none")
```


## Distance run: winners vs losers

Do match winners consistently run more or less than their opponents? Distance run data reflects how much each player was pushed around the court.

```{r distance-run}
#| fig-height: 5
#| fig-width: 11

# winner column in matches may be a player name or "1"/"2" — try both
matches_winner <- matches_raw |>
  mutate(winner_int = suppressWarnings(as.integer(winner))) |>
  filter(!is.na(winner_int)) |>
  select(match_id, winner_int)

dist_data <- pts |>
  filter(!is.na(P1DistanceRun), !is.na(P2DistanceRun),
         P1DistanceRun > 0, P2DistanceRun > 0) |>
  inner_join(matches_winner, by = "match_id") |>
  mutate(
    winner_dist = if_else(winner_int == 1L, P1DistanceRun, P2DistanceRun),
    loser_dist  = if_else(winner_int == 1L, P2DistanceRun, P1DistanceRun)
  ) |>
  group_by(match_id, slam, year) |>
  summarise(
    winner_total = sum(winner_dist, na.rm = TRUE),
    loser_total  = sum(loser_dist,  na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(winner_total > 0, loser_total > 0) |>
  mutate(
    diff = winner_total - loser_total,
    slam_label = slam_labels[slam]
  )

if (nrow(dist_data) > 0) {
  dist_data |>
    pivot_longer(c(winner_total, loser_total), names_to = "result", values_to = "dist") |>
    mutate(result = if_else(result == "winner_total", "Match winner", "Match loser")) |>
    ggplot(aes(dist / 1000, fill = result)) +
    geom_density(alpha = 0.6, colour = NA) +
    scale_fill_manual(values = c("Match winner" = "#27AE60", "Match loser" = "#CB4335")) +
    scale_x_continuous(labels = \(x) paste0(x, "k")) +
    facet_wrap(~slam_label) +
    labs(
      title = "Total distance run per match: winners vs losers",
      subtitle = "Metres run by winner and loser across tracked matches",
      x = "Distance run (metres, thousands)", y = "Density", fill = NULL
    ) |> print()
} else {
  message("Distance run data not available in this dataset.")
}
```

```{r distance-diff}
#| fig-height: 4
#| fig-width: 9

if (nrow(dist_data) > 0) {
  dist_data |>
    ggplot(aes(diff, fill = slam)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_density(alpha = 0.65, colour = NA) +
    scale_fill_manual(values = slam_cols, labels = slam_labels) +
    labs(
      title = "Winner's distance advantage per match",
      subtitle = "Positive = winner ran more; negative = winner ran less",
      x = "Winner metres − Loser metres", y = "Density", fill = NULL
    )
}
```


## After a long point: does the loser fade?

When a rally goes long, one player wins and one loses. Does the losing player carry any fatigue into the **next** point — or do elite players reset completely?

The test: for every point following a rally of length ≥ threshold, compute the win rate of the player who **lost** the previous long point, split by whether they are serving or returning. Compare against their overall win rate in that role.

```{r fatigue-data}
#| echo: false

# Sort points within each match in rally order
pts_sorted <- pts |>
  filter(!is.na(RallyCount), !is.na(PointWinner), !is.na(PointServer)) |>
  arrange(match_id, SetNo, GameNo, PointNumber)

# For each point, tag the loser of the PREVIOUS point and its rally length
pts_lagged <- pts_sorted |>
  group_by(match_id) |>
  mutate(
    prev_rally  = lag(RallyCount),
    prev_winner = lag(PointWinner),   # who won previous point (1 or 2)
    prev_server = lag(PointServer)    # who served previous point
  ) |>
  ungroup() |>
  filter(!is.na(prev_rally), !is.na(prev_winner)) |>
  mutate(
    # loser of previous point
    prev_loser = if_else(prev_winner == 1L, 2L, 1L),
    # did the loser of previous point win this point?
    loser_won_next = as.integer(PointWinner == prev_loser),
    # role of that loser on THIS point
    prev_loser_role = case_when(
      prev_loser == PointServer ~ "Serving next",
      TRUE                      ~ "Returning next"
    )
  )

# Baselines: overall win rate by role
baseline <- pts_lagged |>
  group_by(prev_loser_role) |>
  summarise(baseline_win = mean(loser_won_next, na.rm = TRUE), .groups = "drop")

# Fatigue effect by rally threshold
thresholds <- c(5, 8, 10, 12, 15, 20)

fatigue_df <- map_dfr(thresholds, function(thr) {
  pts_lagged |>
    filter(prev_rally >= thr) |>
    group_by(prev_loser_role) |>
    summarise(
      win_rate = mean(loser_won_next, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    ) |>
    mutate(threshold = thr)
}) |>
  left_join(baseline, by = "prev_loser_role") |>
  mutate(edge = win_rate - baseline_win)
```

```{r fatigue-plot}
#| fig-height: 5
#| fig-width: 10

fatigue_df |>
  ggplot(aes(threshold, edge, colour = prev_loser_role)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 1.1) +
  geom_point(aes(size = n)) +
  scale_colour_manual(values = c("Serving next" = "#2171B5", "Returning next" = "#CB4335")) +
  scale_size_continuous(range = c(2, 6), labels = comma) +
  scale_x_continuous(breaks = thresholds) +
  scale_y_continuous(labels = \(x) paste0(round(x * 100, 1), " pp")) +
  labs(
    title = "Long-point fatigue: does losing a long rally hurt the next point?",
    subtitle = "Win rate of previous-point loser on next point, relative to their baseline\nNegative = they win next point less often than expected",
    x = "Minimum rally length of previous point (shots)",
    y = "Win rate vs baseline (percentage points)",
    colour = "Role on next point",
    size = "Points (n)"
  )
```

```{r fatigue-by-slam}
#| fig-height: 5
#| fig-width: 11

thr_main <- 10  # focus on rallies of 10+ shots

pts_lagged |>
  filter(prev_rally >= thr_main) |>
  group_by(slam, prev_loser_role) |>
  summarise(
    win_rate = mean(loser_won_next, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  left_join(baseline, by = "prev_loser_role") |>
  mutate(
    edge = win_rate - baseline_win,
    slam_label = slam_labels[slam]
  ) |>
  ggplot(aes(slam_label, edge, fill = slam)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = slam_cols) +
  scale_y_continuous(labels = \(x) paste0(round(x * 100, 1), " pp")) +
  facet_wrap(~prev_loser_role) +
  labs(
    title = "Fatigue effect by slam (rallies ≥ 10 shots)",
    subtitle = "Win rate of long-point loser on next point vs. baseline",
    x = NULL, y = "vs. baseline (pp)", fill = NULL
  ) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r fatigue-table}
fatigue_df |>
  filter(threshold %in% c(5, 10, 15)) |>
  mutate(
    win_pct      = percent(win_rate, accuracy = 0.1),
    baseline_pct = percent(baseline_win, accuracy = 0.1),
    edge_pp      = paste0(round(edge * 100, 2), " pp"),
    n            = comma(n)
  ) |>
  select(
    `Min rally` = threshold,
    Role = prev_loser_role,
    `Win rate (next point)` = win_pct,
    `Baseline win rate` = baseline_pct,
    `Difference` = edge_pp,
    `N` = n
  ) |>
  arrange(`Min rally`, Role) |>
  gt() |>
  tab_header(
    title    = "Long-point fatigue effect",
    subtitle = "Win rate of previous-point loser on the immediately following point"
  ) |>
  tab_style(
    style = cell_text(color = "#CB4335", weight = "bold"),
    locations = cells_body(
      columns = Difference,
      rows    = as.numeric(str_remove(Difference, " pp")) < 0
    )
  )
```
