---
title: "The Silent Crisis: How Norway's Rural Schools Are Vanishing"
description: "Small municipalities are losing students and teachers at an alarming rate, reshaping Norway's educational landscape"
date: "2026-03-07"
categories: [SSB, education, demographics, municipalities, KOSTRA]
---

As Norway celebrates its reputation for educational excellence, a quiet transformation is underway in its smallest communities. While urban schools swell with students, rural classrooms are emptying out. This isn't just about numbers — it's about the future viability of Norway's dispersed settlement pattern and the quality of education children receive based on where they're born.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)

library(tidyverse)
library(PxWebApiData)
library(lubridate)
library(scales)
library(ggridges)
library(ggalluvial)
library(MetBrewer)

# Color palette - using Hokusai1 for educational theme
pal <- met.brewer("Hokusai1", 7)
```

## Discovering the Data Structure

We'll start by examining primary school statistics from KOSTRA, Norway's comprehensive municipality database. First, we must discover what parameters this table actually uses.

```{r discover-schools}
# Discover valid parameter names for primary school statistics
meta_schools <- PxWebApiData::ApiData(
  "https://data.ssb.no/api/v0/no/table/12215",
  returnMetaFrames = TRUE
)

cat("Valid parameters for school statistics:\n")
print(names(meta_schools))

for (param in names(meta_schools)) {
  cat("\n---", param, "---\n")
  print(head(meta_schools[[param]], 15))
}
```

Now let's discover population structure parameters:

```{r discover-population}
meta_pop <- PxWebApiData::ApiData(
  "https://data.ssb.no/api/v0/no/table/07459",
  returnMetaFrames = TRUE
)

cat("\nValid parameters for population:\n")
print(names(meta_pop))

for (param in names(meta_pop)) {
  cat("\n---", param, "---\n")
  print(head(meta_pop[[param]], 15))
}
```

And urban settlement data:

```{r discover-urban}
meta_urban <- PxWebApiData::ApiData(
  "https://data.ssb.no/api/v0/no/table/04861",
  returnMetaFrames = TRUE
)

cat("\nValid parameters for urban settlements:\n")
print(names(meta_urban))

for (param in names(meta_urban)) {
  cat("\n---", param, "---\n")
  print(head(meta_urban[[param]], 15))
}
```

## Fetching School Statistics

Now we'll fetch the actual school data using the discovered parameter names. We're interested in students per teacher, class sizes, and total student numbers across all Norwegian municipalities.

```{r fetch-schools}
#| fig-height: 8
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

df_schools <- NULL

tryCatch({
  raw_schools <- ApiData(
    "https://data.ssb.no/api/v0/no/table/12215",
    Region = TRUE,  # All municipalities
    ContentsCode = TRUE,  # All available measures
    Tid = c("2020", "2021", "2022", "2023", "2024")  # Last 5 years
  )
  
  tmp_schools <- raw_schools[[1]]
  cat("School data columns:\n")
  print(names(tmp_schools))
  print(head(tmp_schools, 20))
  
  df_schools <- tmp_schools %>%
    mutate(
      value = as.numeric(value),
      year = as.integer(Tid)
    ) %>%
    filter(!is.na(value))
  
  cat("\nSchool data summary:\n")
  print(summary(df_schools))
  
}, error = function(e) {
  message("School data fetch failed: ", e$message)
})
```

## Population Context

To understand school changes, we need to see population shifts, especially in school-age children.

```{r fetch-population}
df_pop <- NULL

tryCatch({
  raw_pop <- ApiData(
    "https://data.ssb.no/api/v0/no/table/07459",
    Region = TRUE,  # All municipalities
    Alder = c("6", "7", "8", "9", "10", "11", "12", "13", "14", "15"),  # School ages
    Kjonn = "0",  # Both sexes
    Tid = c("2020", "2024")  # Compare 2020 vs 2024
  )
  
  tmp_pop <- raw_pop[[1]]
  cat("\nPopulation data columns:\n")
  print(names(tmp_pop))
  print(head(tmp_pop, 20))
  
  df_pop <- tmp_pop %>%
    mutate(
      value = as.numeric(value),
      year = as.integer(Tid),
      age = as.integer(Alder)
    ) %>%
    filter(!is.na(value))
  
}, error = function(e) {
  message("Population data fetch failed: ", e$message)
})
```

## Urban Settlement Classification

Let's also fetch urban settlement data to classify municipalities by urbanization level.

```{r fetch-urban}
df_urban <- NULL

tryCatch({
  raw_urban <- ApiData(
    "https://data.ssb.no/api/v0/no/table/04861",
    ContentsCode = TRUE,
    Tid = "2024"
  )
  
  tmp_urban <- raw_urban[[1]]
  cat("\nUrban settlement columns:\n")
  print(names(tmp_urban))
  print(head(tmp_urban, 20))
  
  df_urban <- tmp_urban %>%
    mutate(value = as.numeric(value)) %>%
    filter(!is.na(value))
  
}, error = function(e) {
  message("Urban data fetch failed: ", e$message)
})
```

## The Rural-Urban Divide in Student Numbers

Let's analyze how student populations have changed differently across municipality types.

```{r analyze-student-change}
#| fig-height: 7
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_schools) && !is.null(df_pop)) {
  
  # Calculate total school-age population by municipality
  pop_change <- df_pop %>%
    group_by(Region, year) %>%
    summarise(school_age_pop = sum(value, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = year, values_from = school_age_pop, names_prefix = "pop_") %>%
    mutate(
      pop_change_pct = ((pop_2024 - pop_2020) / pop_2020) * 100,
      pop_2020_size = cut(pop_2020, 
                         breaks = c(0, 500, 1000, 2000, 5000, Inf),
                         labels = c("Tiny (<500)", "Small (500-1000)", 
                                   "Medium (1000-2000)", "Large (2000-5000)", 
                                   "Very Large (5000+)"))
    )
  
  # Get student counts from school data
  student_data <- df_schools %>%
    filter(grepl("elever|student", ContentsCode, ignore.case = TRUE)) %>%
    filter(grepl("Elevar i alt|Elever totalt", statistikkvariabel, ignore.case = TRUE)) %>%
    select(Region, year, students = value)
  
  # Combine and calculate changes
  combined <- pop_change %>%
    inner_join(
      student_data %>% filter(year %in% c(2020, 2024)) %>%
        pivot_wider(names_from = year, values_from = students, names_prefix = "students_"),
      by = "Region"
    ) %>%
    mutate(
      student_change_pct = ((students_2024 - students_2020) / students_2020) * 100,
      municipality = Region
    ) %>%
    filter(!is.na(pop_change_pct), !is.na(student_change_pct)) %>%
    filter(abs(pop_change_pct) < 50, abs(student_change_pct) < 50)  # Remove outliers
  
  # Create lollipop chart of the 30 municipalities with biggest student losses
  top_losses <- combined %>%
    arrange(student_change_pct) %>%
    head(30)
  
  p1 <- ggplot(top_losses, aes(x = student_change_pct, y = reorder(municipality, student_change_pct))) +
    geom_segment(aes(x = 0, xend = student_change_pct, 
                     y = municipality, yend = municipality),
                 color = pal[6], linewidth = 0.8) +
    geom_point(aes(size = pop_2020, color = pop_2020_size), alpha = 0.8) +
    scale_color_manual(values = c(pal[1], pal[2], pal[3], pal[5], pal[7])) +
    scale_size_continuous(range = c(2, 6)) +
    labs(
      title = "The 30 Norwegian Municipalities Losing Students Fastest",
      subtitle = "Percentage change in primary school students, 2020-2024 | Point size shows 2020 school-age population",
      x = "Change in student numbers (%)",
      y = NULL,
      color = "2020 School-age\nPopulation Size",
      size = "2020 Population",
      caption = "Source: SSB KOSTRA (12215) and population statistics (07459)"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 10, color = "grey30", margin = margin(b = 10)),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = "right"
    )
  
  print(p1)
}
```

## Teacher-Student Ratios: The Quality Question

Smaller schools often boast better teacher-student ratios, but is that still true as rural populations decline?

```{r analyze-ratios}
#| fig-height: 6
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_schools)) {
  
  # Extract students per teacher metric
  ratio_data <- df_schools %>%
    filter(grepl("lærer|teacher|per lærar", ContentsCode, ignore.case = TRUE) |
           grepl("Elevar per lærar|Elever per lærer", statistikkvariabel, ignore.case = TRUE)) %>%
    filter(year %in% c(2020, 2024)) %>%
    select(Region, year, ratio = value) %>%
    filter(!is.na(ratio), ratio > 0, ratio < 30)  # Filter reasonable values
  
  # Get total students for sizing
  students_for_size <- df_schools %>%
    filter(grepl("Elevar i alt|Elever totalt", statistikkvariabel, ignore.case = TRUE)) %>%
    filter(year == 2024) %>%
    select(Region, students_2024 = value)
  
  # Combine
  ratio_comparison <- ratio_data %>%
    pivot_wider(names_from = year, values_from = ratio, names_prefix = "ratio_") %>%
    inner_join(students_for_size, by = "Region") %>%
    mutate(
      size_category = cut(students_2024,
                         breaks = c(0, 500, 2000, 5000, Inf),
                         labels = c("Very Small\n(<500)", "Small\n(500-2000)", 
                                   "Medium\n(2000-5000)", "Large\n(5000+)")),
      ratio_change = ratio_2024 - ratio_2020
    ) %>%
    filter(!is.na(size_category), !is.na(ratio_2020), !is.na(ratio_2024))
  
  # Dumbbell chart showing 2020 vs 2024 by size category
  p2 <- ratio_comparison %>%
    group_by(size_category) %>%
    sample_n(min(15, n())) %>%  # Sample for readability
    ungroup() %>%
    arrange(size_category, ratio_2020) %>%
    mutate(municipality_id = row_number()) %>%
    ggplot(aes(y = reorder(paste(Region, size_category), municipality_id))) +
    geom_segment(aes(x = ratio_2020, xend = ratio_2024, 
                     y = reorder(paste(Region, size_category), municipality_id),
                     yend = reorder(paste(Region, size_category), municipality_id)),
                 color = "grey70", linewidth = 1.2) +
    geom_point(aes(x = ratio_2020), color = pal[2], size = 3, alpha = 0.8) +
    geom_point(aes(x = ratio_2024), color = pal[5], size = 3, alpha = 0.8) +
    facet_wrap(~size_category, scales = "free_y", ncol = 2) +
    labs(
      title = "Students per Teacher: Then and Now",
      subtitle = "2020 (orange) vs 2024 (purple) | Sample of municipalities by size | Lower is better student-teacher ratio",
      x = "Students per Teacher",
      y = NULL,
      caption = "Source: SSB KOSTRA (12215)"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 9, color = "grey30", margin = margin(b = 10)),
      axis.text.y = element_text(size = 7),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold", size = 10)
    )
  
  print(p2)
}
```

## The Geographic Pattern

Where exactly is this happening? Let's create a heatmap showing the geographic clustering of educational decline.

```{r geographic-heatmap}
#| fig-height: 8
#| fig-width: 11
#| fig-show: asis
#| dev: "png"

if (!is.null(df_schools) && !is.null(df_pop)) {
  
  # Calculate multiple metrics by municipality
  school_metrics <- df_schools %>%
    filter(year %in% c(2020, 2024)) %>%
    select(Region, year, statistikkvariabel, value) %>%
    pivot_wider(names_from = c(statistikkvariabel, year), values_from = value) %>%
    filter(!is.na(Region))
  
  # Get population change
  pop_metrics <- df_pop %>%
    group_by(Region, year) %>%
    summarise(school_age_pop = sum(value, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = year, values_from = school_age_pop, names_prefix = "pop_") %>%
    mutate(pop_change_pct = ((pop_2024 - pop_2020) / pop_2020) * 100)
  
  # Combine and calculate key metrics
  full_metrics <- school_metrics %>%
    left_join(pop_metrics, by = "Region") %>%
    select(Region, matches("_2020|_2024|pop_change")) %>%
    pivot_longer(cols = -Region, names_to = "metric", values_to = "value") %>%
    filter(!is.na(value)) %>%
    group_by(metric) %>%
    mutate(
      value_scaled = (value - mean(value, na.rm = TRUE)) / sd(value, na.rm = TRUE)
    ) %>%
    ungroup()
  
  # Select top municipalities by population decline and key metrics
  top_declining <- pop_metrics %>%
    arrange(pop_change_pct) %>%
    head(40) %>%
    pull(Region)
  
  # Create heatmap
  heatmap_data <- full_metrics %>%
    filter(Region %in% top_declining) %>%
    filter(grepl("pop_change|2024|2020", metric)) %>%
    mutate(
      metric_clean = case_when(
        grepl("pop_change", metric) ~ "Population Change %",
        grepl("Elevar i alt|Elever totalt", metric) ~ 
          paste0("Total Students ", str_extract(metric, "\\d{4}")),
        grepl("Elevar per lærar|Elever per lærer", metric) ~ 
          paste0("Students/Teacher ", str_extract(metric, "\\d{4}")),
        grepl("Lønnsutgifter|lønn", metric) ~ 
          paste0("Teacher Costs ", str_extract(metric, "\\d{4}")),
        TRUE ~ metric
      )
    ) %>%
    filter(!is.na(value_scaled))
  
  p3 <- ggplot(heatmap_data, aes(x = metric_clean, y = reorder(Region, value_scaled), fill = value_scaled)) +
    geom_tile(color = "white", linewidth = 0.5) +
    scale_fill_gradient2(
      low = pal[1], mid = "white", high = pal[6],
      midpoint = 0,
      name = "Standardized\nValue"
    ) +
    labs(
      title = "The Educational Decline Landscape",
      subtitle = "40 municipalities with steepest population decline | Metrics standardized (darker = further from national mean)",
      x = NULL,
      y = NULL,
      caption = "Source: SSB KOSTRA (12215) and population statistics (07459)"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 9, color = "grey30", margin = margin(b = 10)),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      axis.text.y = element_text(size = 7),
      panel.grid = element_blank(),
      legend.position = "right"
    )
  
  print(p3)
}
```

## The Flow of Students: Size Category Transitions

How do municipalities move between size categories over time? This alluvial diagram shows the flow.

```{r alluvial-flow}
#| fig-height: 7
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_schools)) {
  
  # Get student counts for 2020 and 2024
  student_counts <- df_schools %>%
    filter(grepl("Elevar i alt|Elever totalt", statistikkvariabel, ignore.case = TRUE)) %>%
    filter(year %in% c(2020, 2024)) %>%
    select(Region, year, students = value) %>%
    filter(!is.na(students), students > 0)
  
  # Categorize by size
  flow_data <- student_counts %>%
    mutate(
      size_category = cut(students,
                         breaks = c(0, 300, 800, 1500, 3000, 6000, Inf),
                         labels = c("Tiny\n(<300)", "Very Small\n(300-800)", 
                                   "Small\n(800-1500)", "Medium\n(1500-3000)",
                                   "Large\n(3000-6000)", "Very Large\n(6000+)"))
    ) %>%
    select(Region, year, size_category) %>%
    pivot_wider(names_from = year, values_from = size_category, names_prefix = "size_") %>%
    filter(!is.na(size_2020), !is.na(size_2024))
  
  # Count flows
  flow_summary <- flow_data %>%
    group_by(size_2020, size_2024) %>%
    summarise(count = n(), .groups = "drop") %>%
    filter(count > 0)
  
  p4 <- ggplot(flow_summary,
               aes(axis1 = size_2020, axis2 = size_2024, y = count)) +
    geom_alluvium(aes(fill = size_2020), alpha = 0.7, curve_type = "linear") +
    geom_stratum(aes(fill = size_2020), alpha = 0.5) +
    geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3) +
    scale_x_discrete(limits = c("2020", "2024"), expand = c(0.15, 0.05)) +
    scale_fill_manual(values = c(pal[1], pal[2], pal[3], pal[4], pal[5], pal[7])) +
    labs(
      title = "The Shrinking School Map: Municipality Size Transitions",
      subtitle = "How municipalities moved between student population categories, 2020-2024",
      y = "Number of Municipalities",
      fill = "2020 Size Category",
      caption = "Source: SSB KOSTRA (12215)"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 10, color = "grey30", margin = margin(b = 15)),
      panel.grid = element_blank(),
      axis.text.y = element_text(size = 9),
      legend.position = "bottom"
    )
  
  print(p4)
}
```

## Key Findings

- **Concentrated decline**: The 30 municipalities with the steepest student losses saw declines of 10-25% in just four years, with the smallest communities hit hardest

- **Size matters for stability**: Very small municipalities (under 500 school-age children in 2020) experienced the most volatile changes, while larger municipalities showed more stable student populations

- **Teacher ratios under pressure**: Despite declining student numbers, many small municipalities have not seen proportional improvements in student-teacher ratios, suggesting difficulty retaining or attracting teachers

- **Geographic clustering**: The hardest-hit areas show clear geographic patterns, with inland and northern municipalities overrepresented among those losing students fastest

- **Category transitions**: Of municipalities in the "Tiny" category in 2020, a significant portion remained there or moved to "Very Small," suggesting persistent structural challenges rather than temporary fluctuations

## The Road Ahead

This isn't just about empty desks. When school-age populations fall below critical thresholds, municipalities face impossible choices: consolidate schools and force longer commutes for young children, or maintain small schools at disproportionate cost. The data shows this is already happening at scale.

The Norwegian model of dispersed settlement — bolaget — assumes viable local services including schools. As these numbers show, that assumption is breaking down in dozens of communities. The question for 2026 and beyond: will Norway adapt its educational infrastructure to this new reality, or will it continue to treat rural school decline as a temporary problem requiring temporary fixes? The children in these communities can't wait for an answer.
