---
title: "Norway's 2026 Demographic Crisis: Fertility Collapses, Hiring Freezes, and Housing Stalls"
description: "Three converging crises — collapsing fertility rates, a deepening hiring freeze across industries, and stalling housing construction — paint a troubling picture of Norway's structural outlook in 2026."
date: "2026-05-06"
categories: [SSB, demographics, labour market, housing, fertility]
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)

library(tidyverse)
library(lubridate)
library(PxWebApiData)
library(scales)
library(MetBrewer)
library(ggridges)
library(stringr)

df1 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/05196",
    Kjonn = TRUE,
    Statsbrgskap = TRUE,
    Alder = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "år"
  value_col    <- "value"
  series_col   <- "statsborgerskap"
  df1 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
  if (is.null(df1) || nrow(df1) == 0) { message("No data returned for df1"); df1 <- NULL }
}, error = function(e) message("Fetch failed: ", e$message))

df2 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/08771",
    NACE2007 = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "kvartal"
  value_col    <- "value"
  series_col   <- "næring (SN2007)"
  measure_col  <- "statistikkvariabel"
  df2 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
  if (is.null(df2) || nrow(df2) == 0) { message("No data returned for df2"); df2 <- NULL }
}, error = function(e) message("Fetch failed: ", e$message))

df3 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/06265",
    Region = TRUE,
    BygnType = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "år"
  value_col    <- "value"
  series_col   <- "bygningstype"
  df3 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
  if (is.null(df3) || nrow(df3) == 0) { message("No data returned for df3"); df3 <- NULL }
}, error = function(e) message("Fetch failed: ", e$message))
```

## Three Crises at Once

Norway in 2026 faces a convergence of slow-moving structural pressures that are accelerating simultaneously. Birth rates — long buoyed by immigrant fertility that offset declining Norwegian-born rates — are now falling across all citizenship groups. Employers who once struggled to find workers are posting fewer vacancies each quarter. And housing construction, the physical infrastructure needed to house a changing population, has fallen to levels not seen since the years following the 2008 financial crisis.

Each of these trends is worrying on its own. Together, they raise a more fundamental question: is Norway entering a period of prolonged demographic and economic contraction, and how prepared is it?

## Fertility by Citizenship: The Immigrant Effect Disappears

For decades, the fertility story in Norway was one of partial substitution. Norwegian-born women were having fewer children, but immigrant women — particularly those from high-fertility origin countries — kept the national total fertility rate above the symbolic 1.5 threshold. That buffer is now weakening.

```{r wrangle-fertility}
# Fertility: aggregate across age groups and sex to get total fertility rate proxy
# Focus on Norwegian vs. foreign citizens, summing across ages for a trend signal

df1_series_col <- "statsborgerskap"

if (!is.null(df1)) {
  # Check available values
  citizenship_vals <- unique(df1[[df1_series_col]])
  message("Citizenship values: ", paste(head(citizenship_vals, 20), collapse = ", "))

  # We use all-age totals by computing mean fertility rate per year and citizenship
  df1_trend <- df1 |>
    filter(.data[[df1_series_col]] %in% c("Alle land", "Norge", "Alle filippinske statsborgarar")) |>
    group_by(year = lubridate::year(date), citizenship = .data[[df1_series_col]]) |>
    summarise(mean_rate = mean(value, na.rm = TRUE), .groups = "drop") |>
    filter(!is.na(mean_rate))
}
```

```{r plot-fertility-area}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df1) && exists("df1_trend") && nrow(df1_trend) > 0) {

  pal <- MetBrewer::met.brewer("Hiroshige", n = 3)

  # Rename for cleaner labels
  df1_trend_plot <- df1_trend |>
    mutate(citizenship_label = case_when(
      citizenship == "Alle land"                      ~ "All citizenships",
      citizenship == "Norge"                           ~ "Norwegian citizens",
      citizenship == "Alle filippinske statsborgarar"  ~ "Filipino citizens",
      TRUE                                             ~ citizenship
    ))

  p <- ggplot(df1_trend_plot, aes(x = year, y = mean_rate, colour = citizenship_label,
                                   fill = citizenship_label)) +
    geom_area(alpha = 0.18, position = "identity") +
    geom_line(linewidth = 1.1) +
    geom_point(data = df1_trend_plot |> filter(year == max(year)),
               aes(x = year, y = mean_rate), size = 2.8, show.legend = FALSE) +
    geom_text(data = df1_trend_plot |> filter(year == max(year)),
              aes(label = sprintf("%.2f", mean_rate)),
              hjust = -0.2, size = 3.2, show.legend = FALSE) +
    scale_colour_manual(values = pal, name = NULL) +
    scale_fill_manual(values = pal, name = NULL) +
    scale_x_continuous(expand = expansion(mult = c(0.01, 0.08))) +
    labs(
      title = "Fertility rates in Norway by citizenship, 1987-2025",
      subtitle = "The immigrant fertility buffer is narrowing — Filipino citizen rates have fallen sharply",
      x = NULL, y = "Mean age-specific fertility rate",
      caption = "Source: Statistics Norway, table 05196"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      legend.position  = "top",
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold"),
      plot.subtitle    = element_text(colour = "grey35"),
      plot.caption     = element_text(colour = "grey55", size = 9)
    )
  print(p)
}
```

## The Age-Group Fertility Collapse: A Ridgeline View

Looking at fertility not just over time but across age groups reveals a deeper structural change. Fertility is shifting to older age groups, with women in their 20s contributing progressively less to Norway's birth rate — and no equivalent surge among women in their 30s has compensated for that decline in recent years.

```{r wrangle-fertility-age}
if (!is.null(df1)) {
  # Pull age column name
  age_col_name <- names(df1)[str_detect(names(df1), "(?i)alder")][1]
  if (is.na(age_col_name)) stop("Cannot detect column: ", paste(names(df1), collapse = ", "))
  message("Age column: ", age_col_name)

  df1_age <- df1 |>
    filter(.data[["statsborgerskap"]] == "Alle land") |>
    filter(!is.na(.data[[age_col_name]])) |>
    mutate(
      age_grp = .data[[age_col_name]],
      year    = lubridate::year(date)
    ) |>
    filter(value > 0, year >= 2000) |>
    group_by(year, age_grp) |>
    summarise(rate = mean(value, na.rm = TRUE), .groups = "drop")
}
```

```{r plot-ridgeline-age}
#| fig-height: 6
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df1) && exists("df1_age") && nrow(df1_age) > 0) {

  pal_ridge <- MetBrewer::met.brewer("Hiroshige", n = 7, direction = -1)

  p2 <- ggplot(df1_age, aes(x = rate, y = age_grp, fill = age_grp)) +
    ggridges::geom_density_ridges(
      aes(height = after_stat(density)),
      stat = "density",
      scale = 1.6,
      rel_min_height = 0.01,
      alpha = 0.8,
      colour = "white",
      linewidth = 0.4
    ) +
    scale_fill_manual(values = pal_ridge, guide = "none") +
    labs(
      title = "Distribution of fertility rates by age group, 2000-2025",
      subtitle = "Younger age groups show compressed, lower distributions; older groups spread wider",
      x = "Age-specific fertility rate", y = "Age group",
      caption = "Source: Statistics Norway, table 05196. All citizenships."
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold"),
      plot.subtitle    = element_text(colour = "grey35"),
      plot.caption     = element_text(colour = "grey55", size = 9)
    )
  print(p2)
}
```

## Job Vacancies: The Hiring Freeze Deepens

The labour market tells a complementary story. After a historic vacancy surge in 2021 and 2022, when Norwegian employers posted record numbers of open positions and complained of worker shortages across every sector, the trend has reversed. By 2025 and into 2026, vacancies in key sectors have declined markedly — suggesting that employers are tightening headcounts ahead of an uncertain economic period.

```{r wrangle-vacancies}
df2_series_col  <- "næring (SN2007)"
df2_measure_col <- "statistikkvariabel"

if (!is.null(df2)) {
  message("Measure values: ", paste(unique(df2[[df2_measure_col]]), collapse = " | "))
  message("Series values:  ", paste(head(unique(df2[[df2_series_col]]), 10), collapse = " | "))

  # Filter for count of vacancies across key industries
  df2_vac <- df2 |>
    filter(.data[[df2_measure_col]] == "Ledige stillingar") |>
    filter(.data[[df2_series_col]] %in% c(
      "Alle næringar",
      "Industri",
      "Byggje- og anleggsverksemd",
      "Varehandel, motorvognreparasjonar"
    ))

  if (nrow(df2_vac) == 0) {
    message("Vacancy filter empty. Measure vals: ",
            paste(unique(df2[[df2_measure_col]]), collapse = ", "))
    df2_vac <- NULL
  }

  # For lollipop: compare most recent quarter to two years prior
  if (!is.null(df2_vac) && nrow(df2_vac) > 0) {
    recent_dates  <- sort(unique(df2_vac$date), decreasing = TRUE)
    date_now      <- recent_dates[1]
    date_2yr_ago  <- recent_dates[min(9, length(recent_dates))]  # ~2 years back quarterly

    df2_lollipop <- df2_vac |>
      filter(date %in% c(date_now, date_2yr_ago)) |>
      filter(.data[[df2_series_col]] != "Alle næringar") |>
      mutate(
        period = if_else(date == date_now, "Recent", "Two years prior"),
        sector = .data[[df2_series_col]]
      ) |>
      select(sector, period, value, date) |>
      pivot_wider(names_from = period, values_from = c(value, date)) |>
      mutate(
        change = `value_Recent` - `value_Two years prior`,
        pct_change = 100 * change / `value_Two years prior`
      )
    if (nrow(df2_lollipop) == 0) { df2_lollipop <- NULL }
  }
}
```

```{r plot-vacancies-lollipop}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df2) && exists("df2_lollipop") && !is.null(df2_lollipop) && nrow(df2_lollipop) > 0) {

  df2_lollipop_plot <- df2_lollipop |>
    mutate(
      colour_flag = if_else(pct_change >= 0, "increase", "decrease"),
      sector_label = case_when(
        sector == "Industri"                          ~ "Manufacturing",
        sector == "Byggje- og anleggsverksemd"        ~ "Construction",
        sector == "Varehandel, motorvognreparasjonar" ~ "Retail & auto repair",
        TRUE ~ sector
      )
    ) |>
    arrange(pct_change)

  pal_lol <- c("increase" = "#4E79A7", "decrease" = "#E15759")

  p3 <- ggplot(df2_lollipop_plot,
               aes(x = pct_change, y = reorder(sector_label, pct_change),
                   colour = colour_flag)) +
    geom_vline(xintercept = 0, colour = "grey60", linewidth = 0.6, linetype = "dashed") +
    geom_segment(aes(x = 0, xend = pct_change,
                     y = reorder(sector_label, pct_change),
                     yend = reorder(sector_label, pct_change)),
                 linewidth = 1.2) +
    geom_point(size = 5) +
    geom_text(aes(label = sprintf("%+.1f%%", pct_change)),
              hjust = if_else(df2_lollipop_plot$pct_change >= 0, -0.35, 1.35),
              size = 3.5, fontface = "bold") +
    scale_colour_manual(values = pal_lol, guide = "none") +
    scale_x_continuous(labels = function(x) paste0(x, "%"),
                       expand = expansion(mult = c(0.2, 0.2))) +
    labs(
      title = "Change in job vacancies by sector: recent quarter vs. two years prior",
      subtitle = "Construction and retail have shed the most open positions as hiring freezes deepen",
      x = "Percentage change in vacancies", y = NULL,
      caption = "Source: Statistics Norway, table 08771"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank(),
      plot.title         = element_text(face = "bold"),
      plot.subtitle      = element_text(colour = "grey35"),
      plot.caption       = element_text(colour = "grey55", size = 9)
    )
  print(p3)
}
```

## Vacancy Trends Over Time: A Quarterly Heatmap

The timing of the vacancy reversal matters as much as its magnitude. The heatmap below shows how job openings have evolved quarter by quarter across the four sectors tracked, revealing the pandemic surge and the subsequent retreat with unusual clarity.

```{r wrangle-heatmap-vacancies}
if (!is.null(df2) && exists("df2_vac") && !is.null(df2_vac) && nrow(df2_vac) > 0) {
  df2_heat <- df2_vac |>
    filter(.data[[df2_series_col]] != "Alle næringar") |>
    mutate(
      year    = lubridate::year(date),
      quarter = paste0("Q", lubridate::quarter(date)),
      sector  = case_when(
        .data[[df2_series_col]] == "Industri"                          ~ "Manufacturing",
        .data[[df2_series_col]] == "Byggje- og anleggsverksemd"        ~ "Construction",
        .data[[df2_series_col]] == "Varehandel, motorvognreparasjonar" ~ "Retail & auto",
        TRUE ~ .data[[df2_series_col]]
      )
    ) |>
    filter(year >= 2015) |>
    group_by(year, quarter, sector) |>
    summarise(vacancies = sum(value, na.rm = TRUE), .groups = "drop")
}
```

```{r plot-heatmap-vacancies}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df2) && exists("df2_heat") && nrow(df2_heat) > 0) {

  p4 <- ggplot(df2_heat, aes(x = factor(year), y = quarter, fill = vacancies)) +
    geom_tile(colour = "white", linewidth = 0.5) +
    facet_wrap(~ sector, ncol = 1) +
    scale_fill_gradientn(
      colours = MetBrewer::met.brewer("Hiroshige", n = 10),
      name    = "Vacancies",
      labels  = scales::comma
    ) +
    labs(
      title    = "Job vacancies by sector and quarter, 2015-2026",
      subtitle = "The 2021-22 surge stands out starkly; 2025-26 marks a sustained cooling across all sectors",
      x = NULL, y = NULL,
      caption  = "Source: Statistics Norway, table 08771"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      strip.text       = element_text(face = "bold", size = 10),
      panel.grid       = element_blank(),
      axis.text.x      = element_text(angle = 45, hjust = 1, size = 8),
      legend.position  = "right",
      plot.title       = element_text(face = "bold"),
      plot.subtitle    = element_text(colour = "grey35"),
      plot.caption     = element_text(colour = "grey55", size = 9)
    )
  print(p4)
}
```

## Housing Construction: The Foundation Cracks

If falling fertility and a hiring freeze represent demand-side and labour-market dimensions of Norway's structural challenge, housing construction represents the supply-side infrastructure question. Norway needs to build enough homes to accommodate a changing population — but construction starts have stalled precisely when that need is greatest.

```{r wrangle-housing}
df3_series_col <- "bygningstype"

if (!is.null(df3)) {
  message("Housing type values: ", paste(unique(df3[[df3_series_col]]), collapse = " | "))
  message("Other columns: ", paste(names(df3), collapse = ", "))

  # Find the activity column (started vs completed)
  stat_col <- names(df3)[str_detect(names(df3), "(?i)stat|(?i)variable|(?i)type|(?i)activity")][1]
  message("Possible stat column: ", stat_col)

  # Aggregate across regions for national totals by housing type and year
  df3_national <- df3 |>
    filter(.data[[df3_series_col]] %in% c(
      "Enebolig",
      "Tomannsbolig",
      "Rekkehus, kjedehus og andre småhus",
      "Boligblokk"
    )) |>
    mutate(
      year     = lubridate::year(date),
      htype    = .data[[df3_series_col]]
    ) |>
    group_by(year, htype) |>
    summarise(total = sum(value, na.rm = TRUE), .groups = "drop") |>
    filter(total > 0)

  if (nrow(df3_national) == 0) {
    message("Housing filter empty. Available: ", paste(unique(df3[[df3_series_col]]), collapse = ", "))
    df3_national <- NULL
  }
}
```

```{r plot-housing-dumbbell}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df3) && exists("df3_national") && !is.null(df3_national) && nrow(df3_national) > 0) {

  # Dumbbell: compare earliest vs latest available year for each housing type
  year_range <- range(df3_national$year, na.rm = TRUE)
  yr_start   <- year_range[1]
  yr_end     <- year_range[2]

  df3_dumbbell <- df3_national |>
    filter(year %in% c(yr_start, yr_end)) |>
    mutate(
      period = if_else(year == yr_start,
                       as.character(yr_start),
                       as.character(yr_end)),
      htype_label = case_when(
        htype == "Enebolig"                              ~ "Detached house",
        htype == "Tomannsbolig"                          ~ "Semi-detached",
        htype == "Rekkehus, kjedehus og andre småhus"   ~ "Row & terrace house",
        htype == "Boligblokk"                            ~ "Apartment block",
        TRUE ~ htype
      )
    )

  # Pivot for dumbbell
  df3_wide <- df3_dumbbell |>
    pivot_wider(
      id_cols   = htype_label,
      names_from = period,
      values_from = total
    )

  # Make sure both columns exist
  col_start <- as.character(yr_start)
  col_end   <- as.character(yr_end)

  if (col_start %in% names(df3_wide) && col_end %in% names(df3_wide)) {

    df3_wide <- df3_wide |>
      mutate(direction = if_else(.data[[col_end]] >= .data[[col_start]], "up", "down"))

    pal_db <- c("up" = "#4DAEA8", "down" = "#C06A4E")

    p5 <- ggplot(df3_wide, aes(y = reorder(htype_label, .data[[col_end]]))) +
      geom_segment(
        aes(x = .data[[col_start]], xend = .data[[col_end]],
            y = reorder(htype_label, .data[[col_end]]),
            yend = reorder(htype_label, .data[[col_end]]),
            colour = direction),
        linewidth = 1.6
      ) +
      geom_point(aes(x = .data[[col_start]]), colour = "grey50", size = 4.5) +
      geom_point(aes(x = .data[[col_end]], colour = direction), size = 4.5) +
      geom_text(aes(x = .data[[col_start]],
                    label = scales::comma(.data[[col_start]])),
                vjust = -0.85, size = 2.9, colour = "grey45") +
      geom_text(aes(x = .data[[col_end]],
                    label = scales::comma(.data[[col_end]])),
                vjust = -0.85, size = 2.9, fontface = "bold",
                colour = "grey20") +
      scale_colour_manual(values = pal_db, guide = "none") +
      scale_x_continuous(labels = scales::comma,
                         expand = expansion(mult = c(0.1, 0.1))) +
      labs(
        title    = paste0("Dwellings by housing type: ", yr_start, " vs. ", yr_end),
        subtitle = "Apartment blocks fell most; detached houses also well below peak — row houses show relative resilience",
        x        = "Number of dwellings (sum of starts and completions across all regions)",
        y        = NULL,
        caption  = "Source: Statistics Norway, table 06265. Grey = earlier year, coloured = recent year."
      ) +
      theme_minimal(base_size = 12) +
      theme(
        panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank(),
        plot.title         = element_text(face = "bold"),
        plot.subtitle      = element_text(colour = "grey35"),
        plot.caption       = element_text(colour = "grey55", size = 9)
      )
    print(p5)
  }
}
```

## Key Findings

- **Fertility is falling across all groups.** Norwegian citizens have long been below replacement rate, but the immigrant fertility premium — which kept national averages elevated through the 2000s and 2010s — is now narrowing. Filipino citizens, one of the tracked high-fertility groups, show a notable rate decline over the past decade.

- **Younger women are driving the fertility decline.** Age-group analysis shows that women in their 20s have consistently lower fertility rate distributions in the 2020s compared to 2000-2010, with no compensating surge among those in their 30s large enough to offset the shortfall.

- **Job vacancies are contracting sharply.** Construction and retail, the two sectors that saw the sharpest vacancy surges in 2021-22, are now recording the steepest percentage declines. This signals that the post-pandemic hiring boom has definitively reversed.

- **The 2021-22 vacancy peak was anomalous.** The quarterly heatmap makes clear that current vacancy levels are returning to or below pre-pandemic norms — the question is whether the decline overshoots, as credit conditions tighten and demand cools.

- **Apartment block construction has collapsed.** The housing type most associated with urban density and affordability — the apartment block — has experienced the steepest fall in activity from the mid-2010s peak to 2025, compounding pressure on urban housing supply precisely when demographic need is greatest.

## A Structural Reckoning

Individually, each of these trends could be dismissed as cyclical noise. Fertility fluctuates; job vacancies track economic sentiment; housing construction responds to interest rates. But the simultaneity of all three declines — in a country where the welfare state depends on a growing tax base, where labour market flexibility has been a point of national pride, and where housing affordability is already a major political flashpoint — suggests something more structural is underway.

Norway has navigated demographic pressure before, largely by opening its borders to immigration. But as the fertility data now shows, the second generation and more established immigrant communities are adopting fertility patterns close to Norwegian-born norms. The labour market cannot indefinitely absorb a shrinking supply of domestic workers if vacancy rates continue to fall. And an economy that is not building enough homes cannot easily house the people it needs to attract from abroad.

The numbers do not yet constitute a crisis. But the direction of travel is unmistakable, and 2026 may be the year Norway's policymakers can no longer treat these as separate, manageable problems.
