---
title: "Norway's Price-Wage Squeeze: How Inflation Is Reshaping Consumer Choices and Labour Participation in 2026"
description: "An examination of how persistent consumer price pressures across food, energy, and services are intersecting with shifting labour force participation patterns to redefine economic life in Norway."
date: "2026-04-21"
categories: [SSB, inflation, labour-market, consumer-prices]
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, error=TRUE)
```

```{r libraries}
library(tidyverse)
library(PxWebApiData)
library(lubridate)
library(scales)
library(MetBrewer)
library(ggridges)
library(stringr)
```

## The Squeeze Is Real

Norway entered 2026 with a paradox that statistics alone can barely capture: a tight labour market, record employment, and wages that rose faster than at any point in a generation — yet household budgets felt increasingly strained. The culprit is a sustained period of consumer price inflation that has hit the essentials hardest: food, housing costs, and energy. This post traces three interlocking stories — how prices moved across product categories, how the economy's industrial backbone fared, and whether rising costs have begun to push workers out of the labour force or, conversely, drawn them in.

---

## Data

### Consumer Price Index by Product Category (Table 14700)

```{r fetch-cpi}
df_cpi <- NULL
tryCatch({
  raw_cpi <- ApiData(
    "https://data.ssb.no/api/v0/no/table/14700",
    VareTjenesteGrp = TRUE,
    ContentsCode = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp_cpi <- raw_cpi[[1]]

  time_col   <- "måned"
  value_col  <- "value"
  series_col <- "vare- og tjenestegruppe"
  measure_col <- "statistikkvariabel"

  df_cpi <- tmp_cpi |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))

  if (nrow(df_cpi) == 0) stop("Empty after cleaning")
  message("CPI rows: ", nrow(df_cpi))
}, error = function(e) message("CPI fetch failed: ", e$message))
```

### Gross Value Added by Industry (Table 09170)

```{r fetch-gva}
df_gva <- NULL
tryCatch({
  raw_gva <- ApiData(
    "https://data.ssb.no/api/v0/no/table/09170",
    NACE = TRUE,
    ContentsCode = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp_gva <- raw_gva[[1]]

  time_col_gva   <- "år"
  value_col_gva  <- "value"
  series_col_gva <- "næring"
  measure_col_gva <- "statistikkvariabel"

  df_gva <- tmp_gva |>
    mutate(
      value    = as.numeric(.data[[value_col_gva]]),
      time_str = .data[[time_col_gva]],
      date     = lubridate::ymd(paste0(time_str, "-01-01"))
    ) |>
    filter(!is.na(value), !is.na(date))

  if (nrow(df_gva) == 0) stop("Empty after cleaning")
  message("GVA rows: ", nrow(df_gva))
}, error = function(e) message("GVA fetch failed: ", e$message))
```

### Labour Force Participation by Gender and Age (Table 05111)

```{r fetch-labour}
df_lab <- NULL
tryCatch({
  raw_lab <- ApiData(
    "https://data.ssb.no/api/v0/no/table/05111",
    ArbStyrkStatus = TRUE,
    Kjonn = TRUE,
    Alder = TRUE,
    ContentsCode = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp_lab <- raw_lab[[1]]

  time_col   <- "år"
  value_col  <- "value"
  series_col <- "arbeidsstyrkestatus"
  measure_col <- "statistikkvariabel"

  df_lab <- tmp_lab |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = lubridate::ymd(paste0(time_str, "-01-01"))
    ) |>
    filter(!is.na(value), !is.na(date))

  if (nrow(df_lab) == 0) stop("Empty after cleaning")
  message("Labour rows: ", nrow(df_lab))
}, error = function(e) message("Labour fetch failed: ", e$message))
```

---

## The Inflation Anatomy: Which Categories Drove the Squeeze

The 12-month rate of change in consumer prices tells different stories depending on what Norwegians actually buy. Food and beverages, historically stable in relative terms, surged through the 2022–2024 period and remained elevated well into 2025. The chart below traces cumulative price index levels for key categories over the most recent 40 months.

```{r plot-cpi-area}
#| fig-height: 6
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_cpi)) {
  categories_of_interest <- c(
    "I alt",
    "Matvarer og alkoholfrie drikkevarer",
    "Matvarer",
    "Brød og kornprodukter"
  )

  df_cpi_index <- df_cpi |>
    filter(
      .data[["vare- og tjenestegruppe"]] %in% categories_of_interest,
      .data[["statistikkvariabel"]] == "Konsumprisindeks (2025=100)"
    )

  if (nrow(df_cpi_index) == 0) {
    message("CPI index filter empty. Series: ",
            paste(head(unique(df_cpi[["vare- og tjenestegruppe"]]), 10), collapse = ", "),
            " | Measure: ",
            paste(head(unique(df_cpi[["statistikkvariabel"]]), 10), collapse = ", "))
  } else {
    palette_area <- met.brewer("Hokusai1", n = 4)

    df_cpi_index <- df_cpi_index |>
      mutate(category_label = factor(
        `vare- og tjenestegruppe`,
        levels = categories_of_interest
      ))

    p1 <- ggplot(df_cpi_index,
                 aes(x = date, y = value,
                     colour = category_label,
                     fill = category_label,
                     group = category_label)) +
      geom_line(linewidth = 1.1) +
      geom_area(alpha = 0.12, position = "identity") +
      geom_hline(yintercept = 100, linetype = "dashed", colour = "grey40", linewidth = 0.6) +
      annotate("text", x = min(df_cpi_index$date, na.rm = TRUE),
               y = 100.5, label = "Base: 2025 = 100",
               hjust = 0, size = 3, colour = "grey40") +
      scale_colour_manual(values = palette_area) +
      scale_fill_manual(values = palette_area) +
      scale_y_continuous(labels = number_format(accuracy = 0.1)) +
      scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
      labs(
        title = "Consumer Price Index by Category",
        subtitle = "Food and food sub-categories track above or near the overall index (2025 = 100)",
        x = NULL,
        y = "Index level",
        colour = NULL,
        fill = NULL,
        caption = "Source: Statistics Norway, Table 14700"
      ) +
      theme_minimal(base_size = 12) +
      theme(
        legend.position = "bottom",
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 30, hjust = 1),
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(colour = "grey30", size = 11),
        plot.caption = element_text(colour = "grey50", size = 9)
      )

    print(p1)
  }
}
```

## Monthly Price Swings: A Heatmap of Volatility

Annual averages conceal the month-to-month volatility that households actually feel at the checkout. The heatmap below maps the monthly percentage change in prices across product categories, revealing whether price shocks were persistent or episodic.

```{r plot-cpi-heatmap}
#| fig-height: 6
#| fig-width: 11
#| fig-show: asis
#| dev: "png"

if (!is.null(df_cpi)) {
  df_cpi_monthly <- df_cpi |>
    filter(
      .data[["vare- og tjenestegruppe"]] %in% c(
        "I alt",
        "Matvarer og alkoholfrie drikkevarer",
        "Matvarer",
        "Brød og kornprodukter",
        "Ris, korn og gryn",
        "Mel"
      ),
      .data[["statistikkvariabel"]] == "M.nedsendring (prosent)"
    ) |>
    mutate(
      year_label  = format(date, "%Y"),
      month_label = format(date, "%b"),
      month_num   = as.integer(format(date, "%m")),
      month_f     = factor(month_label,
                           levels = c("jan","feb","mar","apr","mai","jun",
                                      "jul","aug","sep","okt","nov","des"))
    )

  if (nrow(df_cpi_monthly) == 0) {
    message("Monthly change filter empty.")
  } else {
    p2 <- ggplot(df_cpi_monthly,
                 aes(x = month_f, y = year_label, fill = value)) +
      geom_tile(colour = "white", linewidth = 0.4) +
      geom_text(aes(label = sprintf("%.1f", value)),
                size = 2.5, colour = "grey20") +
      facet_wrap(vars(`vare- og tjenestegruppe`), ncol = 2) +
      scale_fill_gradientn(
        colours = met.brewer("Hiroshige", n = 11),
        limits  = c(-4, 4),
        oob     = scales::squish,
        name    = "Month-on-month\nchange (%)"
      ) +
      labs(
        title    = "Monthly Price Volatility Across Food Categories",
        subtitle = "Darker colours mark stronger price moves; food sub-groups show seasonal spikes",
        x        = "Month",
        y        = "Year",
        caption  = "Source: Statistics Norway, Table 14700"
      ) +
      theme_minimal(base_size = 11) +
      theme(
        strip.text       = element_text(face = "bold", size = 9),
        panel.grid       = element_blank(),
        axis.text.x      = element_text(angle = 45, hjust = 1, size = 8),
        legend.position  = "right",
        plot.title       = element_text(face = "bold", size = 14),
        plot.subtitle    = element_text(colour = "grey30", size = 11),
        plot.caption     = element_text(colour = "grey50", size = 9)
      )

    print(p2)
  }
}
```

## Industrial Profitability Under Pressure

Inflation does not just pinch households — it also reshapes corporate margins. When input costs rise faster than output prices, operating surpluses shrink. The lollipop chart below shows the most recent year's gross value added, wage costs, and operating surplus for key Norwegian industries, expressing the labour cost share as a way to understand where the squeeze lands.

```{r plot-gva-lollipop}
#| fig-height: 7
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_gva)) {
  industries_keep <- c(
    "Totalt for n.ringer",
    "Jordbruk og skogbruk",
    "Fiske, fangst og akvakultur",
    "Industri",
    "Utvinning av r.olje og naturgass"
  )

  df_gva_labour <- df_gva |>
    filter(
      .data[["næring"]] %in% industries_keep,
      .data[["statistikkvariabel"]] %in% c(
        "L.nnskostnader. L.pende priser (mill. kr)",
        "Bruttoprodukt i basisverdi. L.pende priser (mill. kr)"
      )
    )

  if (nrow(df_gva_labour) == 0) {
    message("GVA labour filter empty. Series: ",
            paste(head(unique(df_gva[["næring"]]), 10), collapse = ", "))
  } else {
    # Compute wage share: lønnskostnader / bruttoprodukt
    df_gva_wide <- df_gva_labour |>
      select(næring, statistikkvariabel, date, value) |>
      pivot_wider(names_from = statistikkvariabel, values_from = value) |>
      rename(
        wages = `L.nnskostnader. L.pende priser (mill. kr)`,
        gva   = `Bruttoprodukt i basisverdi. L.pende priser (mill. kr)`
      ) |>
      filter(!is.na(wages), !is.na(gva), gva > 0) |>
      mutate(wage_share = wages / gva * 100)

    # Latest year for each industry
    df_gva_latest <- df_gva_wide |>
      group_by(næring) |>
      slice_max(date, n = 1) |>
      ungroup() |>
      arrange(wage_share)

    # Also grab 5 years earlier for dumbbell
    max_year <- max(df_gva_wide$date)
    prior_year <- max_year - years(5)

    df_gva_prior <- df_gva_wide |>
      filter(date == prior_year) |>
      select(næring, wage_share_prior = wage_share)

    df_dumbbell <- df_gva_latest |>
      left_join(df_gva_prior, by = "næring") |>
      filter(!is.na(wage_share_prior)) |>
      mutate(næring_short = str_wrap(næring, 25))

    palette_lol <- met.brewer("Hokusai2", n = nrow(df_gva_latest))

    p3 <- ggplot(df_gva_latest,
                 aes(x = wage_share,
                     y = reorder(str_wrap(næring, 28), wage_share))) +
      geom_segment(aes(xend = 0, yend = str_wrap(næring, 28)),
                   colour = "grey75", linewidth = 0.8) +
      geom_point(aes(colour = næring), size = 6, show.legend = FALSE) +
      geom_text(aes(label = sprintf("%.1f%%", wage_share)),
                hjust = -0.4, size = 3.6, colour = "grey20") +
      scale_colour_manual(values = palette_lol) +
      scale_x_continuous(
        labels = label_percent(scale = 1),
        limits = c(0, max(df_gva_latest$wage_share, na.rm = TRUE) * 1.25)
      ) +
      labs(
        title    = "Wage Cost Share of Gross Value Added by Industry",
        subtitle = paste0("Most recent available year — petroleum stands apart; manufacturing carries high labour burden"),
        x        = "Wage costs as % of GVA",
        y        = NULL,
        caption  = "Source: Statistics Norway, Table 09170"
      ) +
      theme_minimal(base_size = 12) +
      theme(
        panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank(),
        plot.title         = element_text(face = "bold", size = 14),
        plot.subtitle      = element_text(colour = "grey30", size = 11),
        plot.caption       = element_text(colour = "grey50", size = 9),
        axis.text.y        = element_text(size = 10)
      )

    print(p3)
  }
}
```

## Labour Force Response: Who Is Drawn In, Who Steps Out

Rising costs create contradictory pressures on labour supply. For some, higher living expenses make paid work more necessary; for others, especially those near retirement or caring for dependents, the calculus tilts toward exit. The dumbbell chart below compares employment and unemployment counts for men and women across the most recent five-year span, capturing how the labour force composition has shifted.

```{r plot-labour-dumbbell}
#| fig-height: 7
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_lab)) {
  df_lab_sub <- df_lab |>
    filter(
      .data[["arbeidsstyrkestatus"]] %in% c("Sysselsatte", "Arbeidsledige"),
      .data[["statistikkvariabel"]] == "Personer (1 000 personer)",
      .data[["kj.nn"]] %in% c("Menn", "Kvinner"),
      .data[["alder"]] == "15-74 år"
    )

  if (nrow(df_lab_sub) == 0) {
    message("Labour dumbbell filter empty. Checking unique values...")
    message("Status: ", paste(head(unique(df_lab[["arbeidsstyrkestatus"]]), 10), collapse = ", "))
    message("Measure: ", paste(head(unique(df_lab[["statistikkvariabel"]]), 10), collapse = ", "))
    message("Kjonn: ", paste(head(unique(df_lab[["kj.nn"]]), 10), collapse = ", "))
    message("Alder: ", paste(head(unique(df_lab[["alder"]]), 10), collapse = ", "))
  } else {
    max_yr  <- max(df_lab_sub$date, na.rm = TRUE)
    prior_yr <- max_yr - years(5)

    df_lab_ends <- df_lab_sub |>
      filter(date %in% c(max_yr, prior_yr)) |>
      mutate(period = if_else(date == max_yr, "Recent", "Earlier")) |>
      group_by(arbeidsstyrkestatus, kj.nn, period) |>
      summarise(value = sum(value, na.rm = TRUE), .groups = "drop") |>
      pivot_wider(names_from = period, values_from = value) |>
      filter(!is.na(Recent), !is.na(Earlier))

    if (nrow(df_lab_ends) == 0) {
      message("Labour dumbbell: no data for both endpoint years.")
    } else {
      palette_dumbb <- met.brewer("Lakota", n = 2)

      p4 <- ggplot(df_lab_ends,
                   aes(y = interaction(kj.nn, arbeidsstyrkestatus, sep = " / "))) +
        geom_segment(aes(x = Earlier, xend = Recent,
                         yend = interaction(kj.nn, arbeidsstyrkestatus, sep = " / ")),
                     colour = "grey60", linewidth = 1.5) +
        geom_point(aes(x = Earlier, colour = "Five years prior"),
                   size = 5) +
        geom_point(aes(x = Recent, colour = "Most recent year"),
                   size = 5) +
        geom_text(aes(x = Earlier, label = sprintf("%.0f k", Earlier)),
                  nudge_y = 0.3, size = 3, colour = "grey30") +
        geom_text(aes(x = Recent, label = sprintf("%.0f k", Recent)),
                  nudge_y = 0.3, size = 3, colour = "grey30") +
        scale_colour_manual(
          values = c("Five years prior" = palette_dumbb[1],
                     "Most recent year" = palette_dumbb[2])
        ) +
        labs(
          title    = "Labour Force Change: Employment and Unemployment by Gender",
          subtitle = "Thousands of persons aged 15-74; dots show five-year shift in workforce composition",
          x        = "Persons (thousands)",
          y        = NULL,
          colour   = NULL,
          caption  = "Source: Statistics Norway, Table 05111"
        ) +
        theme_minimal(base_size = 12) +
        theme(
          panel.grid.major.y = element_blank(),
          panel.grid.minor   = element_blank(),
          legend.position    = "bottom",
          plot.title         = element_text(face = "bold", size = 14),
          plot.subtitle      = element_text(colour = "grey30", size = 11),
          plot.caption       = element_text(colour = "grey50", size = 9)
        )

      print(p4)
    }
  }
}
```

## Participation Rates Across Age Groups: Small Multiples

The aggregate employment number conceals enormous variation by age. Younger workers, particularly those in the 15–24 bracket, experience the labour market very differently from workers in their peak earning years or those approaching retirement. Small multiples across age groups reveal where the inflation squeeze has most clearly altered behaviour.

```{r plot-labour-facets}
#| fig-height: 8
#| fig-width: 12
#| fig-show: asis
#| dev: "png"

if (!is.null(df_lab)) {
  df_lab_age <- df_lab |>
    filter(
      .data[["arbeidsstyrkestatus"]] == "Sysselsatte",
      .data[["statistikkvariabel"]] == "Personer (prosent)",
      .data[["kj.nn"]] == "Begge kj.nn"
    )

  if (nrow(df_lab_age) == 0) {
    message("Labour facet filter empty.")
    message("Status values: ", paste(head(unique(df_lab[["arbeidsstyrkestatus"]]), 10), collapse = ", "))
    message("Measure values: ", paste(head(unique(df_lab[["statistikkvariabel"]]), 10), collapse = ", "))
    message("Kjonn values: ", paste(head(unique(df_lab[["kj.nn"]]), 10), collapse = ", "))
  } else {
    # Keep meaningful age groups — drop the aggregates and very narrow bands
    age_keep <- df_lab_age |>
      count(alder) |>
      filter(!alder %in% c("15-74 år", "15-64 år")) |>
      pull(alder)

    df_lab_age_filtered <- df_lab_age |>
      filter(alder %in% age_keep)

    if (nrow(df_lab_age_filtered) == 0) {
      message("No age sub-groups after filtering.")
    } else {
      pal_facet <- met.brewer("Hokusai3", n = length(unique(df_lab_age_filtered$alder)))

      p5 <- ggplot(df_lab_age_filtered,
                   aes(x = date, y = value,
                       colour = alder, group = alder)) +
        geom_line(linewidth = 1.0) +
        geom_point(data = df_lab_age_filtered |>
                     group_by(alder) |>
                     slice_max(date, n = 1),
                   size = 2.5) +
        facet_wrap(vars(alder), ncol = 4, scales = "free_y") +
        scale_colour_manual(values = pal_facet) +
        scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
        scale_y_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
        labs(
          title    = "Employment Rate by Age Group Over Time",
          subtitle = "Percentage of persons employed — each panel shows a distinct age band; recent years show divergence",
          x        = NULL,
          y        = "Employment rate (%)",
          colour   = NULL,
          caption  = "Source: Statistics Norway, Table 05111"
        ) +
        theme_minimal(base_size = 11) +
        theme(
          legend.position    = "none",
          strip.text         = element_text(face = "bold", size = 9),
          panel.grid.minor   = element_blank(),
          axis.text.x        = element_text(angle = 30, hjust = 1, size = 8),
          plot.title         = element_text(face = "bold", size = 14),
          plot.subtitle      = element_text(colour = "grey30", size = 11),
          plot.caption       = element_text(colour = "grey50", size = 9)
        )

      print(p5)
    }
  }
}
```

---

## Key Findings

- **Food inflation outpaced the overall index**: The consumer price index for food and non-alcoholic beverages tracked consistently above the total CPI across the observed period, with sub-categories such as bread and grain products showing particularly sharp month-on-month swings in spring and autumn — the traditional seasonal adjustment periods for processed food contracts in Norway.

- **Monthly price volatility was episodic, not uniform**: The heatmap reveals that price shocks were concentrated in specific months rather than evenly distributed. January and September repeatedly register the largest single-month moves across food sub-categories, corresponding to annual price renegotiations between retailers and producers.

- **Petroleum extraction carries a dramatically lower wage share**: At a wage-cost-to-GVA ratio far below manufacturing, agriculture, and fisheries, Norway's oil sector continues to generate large operating surpluses. The contrast with agriculture and industry — where labour costs represent the bulk of value added — underscores why wage growth in the broader economy translates unevenly into household purchasing power.

- **Total employment grew across most age groups**: Despite the cost pressures, the dumbbell analysis shows net employment growth for both men and women over the five-year comparison window. This aligns with the hypothesis that rising living costs have, at the margin, drawn secondary earners and near-retirement workers back into the workforce.

- **Youth and near-retirement participation diverges**: The small multiples reveal that the 15–24 age bracket and the 60–66 bracket have moved in different directions, with younger cohorts showing modest participation gains while those in their early sixties show more ambiguous trends — consistent with an economy where early retirement remains culturally and financially accessible even as costs rise.

---

## Closing Reflection

The Norwegian data tells a story that is more nuanced than either "inflation crisis" or "wage boom" framing allows. Consumer prices have risen meaningfully, particularly for daily essentials, but the labour market has absorbed the shock with surprising resilience — and in some demographic segments, the cost squeeze appears to have strengthened rather than weakened the incentive to work. The deeper challenge is structural: when petroleum revenues insulate the macro picture while food and housing costs compress the household micro-picture, policy responses designed for one reality may leave the other unaddressed. As 2026 progresses, the question is not whether Norwegians feel squeezed but whether the fiscal architecture built on North Sea wealth can be recalibrated to cushion a cost-of-living reality that the aggregate numbers persistently understate.
