---
title: "Norway's Electric Revolution: How EVs Conquered the Roads"
description: "With over 80% of new car sales now electric, we dive into SSB data to understand Norway's unprecedented vehicle transition."
date: "2026-03-03"
categories: [SSB, electric vehicles, energy, environment, transportation]
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  error = TRUE,
  fig.width = 10,
  fig.height = 6
)
```

Norway has become the world's laboratory for electric vehicle adoption. While other countries debate EV mandates for 2035 or 2040, Norway has already crossed a threshold most thought impossible a decade ago. Today, we dig into Statistics Norway's data to understand the speed, scale, and implications of this transformation.

## Setting Up Our Analysis

```{r libraries}
library(tidyverse)
library(PxWebApiData)
library(lubridate)
library(scales)
library(ggridges)

# Custom color palette - inspired by Norwegian nature and electric energy
norway_ev_palette <- c(
  "#0A2342",  # Deep fjord blue
  "#2CA58D",  # Electric green
  "#84BC9C",  # Light green
  "#F46036",  # Warning orange
  "#E8E8E8",  # Light grey
  "#1B4965",  # Medium blue
  "#5FA8D3"   # Sky blue
)

# Custom theme for all charts
theme_ev <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 18, color = "#0A2342"),
      plot.subtitle = element_text(size = 12, color = "#555555", margin = margin(b = 15)),
      plot.caption = element_text(size = 9, color = "#888888", hjust = 0),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      legend.position = "top",
      legend.title = element_blank(),
      axis.title = element_text(size = 11, color = "#444444"),
      plot.margin = margin(20, 20, 20, 20),
      plot.background = element_rect(fill = "#FAFAFA", color = NA)
    )
}
```

## Fetching the EV Registration Data

Let's start by pulling the electric vehicle registration data from SSB. This dataset tracks new vehicle registrations by fuel type.

```{r fetch-ev-data}
# Fetch EV registration data
ev_result <- tryCatch({
  ApiData(
    "https://data.ssb.no/api/v0/en/table/13931",
    Tid = list(filter = "top", values = 60)
  )
}, error = function(e) {
  message("Error fetching EV data: ", e$message)
  NULL
})

# CRITICAL: Always inspect the data structure first
if (!is.null(ev_result)) {
  ev_raw <- ev_result[[1]]
  cat("Column names in EV dataset:\n")
  print(names(ev_raw))
  cat("\nFirst few rows:\n")
  print(head(ev_raw))
}
```

```{r process-ev-data}
# Process the EV data after understanding its structure
if (!is.null(ev_result)) {
  
  # Find the time column
  time_col <- names(ev_raw)[grepl("tid|month|year|quarter|maaned", 
                                   names(ev_raw), ignore.case = TRUE)][1]
  cat("Identified time column:", time_col, "\n")
  
  # Find the value column
  value_col <- names(ev_raw)[grepl("value|antall|number", 
                                    names(ev_raw), ignore.case = TRUE)][1]
  cat("Identified value column:", value_col, "\n")
  
  ev_data <- ev_raw %>%
    mutate(
      value = as.numeric(.data[[value_col]]),
      time_str = as.character(.data[[time_col]]),
      date = case_when(
        str_detect(time_str, "M") ~ ym(str_replace(time_str, "M", "-")),
        str_detect(time_str, "K") ~ yq(str_replace(time_str, "K", " Q")),
        str_length(time_str) == 4 ~ ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) %>%
    filter(!is.na(value), !is.na(date))
  
  cat("\nProcessed data sample:\n")
  print(head(ev_data))
}
```

## The Stacked Area Chart: Norway's Fuel Type Transition

Our first visualization shows the dramatic shift in new vehicle registrations by fuel type over time. This is the "thumbnail" chart — it captures the entire story in one view.

```{r area-chart}
#| fig-height: 8

if (exists("ev_data") && nrow(ev_data) > 0) {
  
  # Check for fuel type column
  fuel_col <- names(ev_data)[grepl("fuel|drivstoff|type", 
                                    names(ev_data), ignore.case = TRUE)][1]
  
  if (!is.na(fuel_col)) {
    
    # Aggregate by date and fuel type
    ev_monthly <- ev_data %>%
      group_by(date, fuel_type = .data[[fuel_col]]) %>%
      summarise(registrations = sum(value, na.rm = TRUE), .groups = "drop") %>%
      filter(date >= ymd("2015-01-01"))
    
    # Calculate percentages for each month
    ev_pct <- ev_monthly %>%
      group_by(date) %>%
      mutate(
        total = sum(registrations),
        pct = registrations / total * 100
      ) %>%
      ungroup()
    
    # Simplify fuel type categories
    ev_pct <- ev_pct %>%
      mutate(
        fuel_category = case_when(
          str_detect(fuel_type, regex("electric|el|battery|bev", ignore_case = TRUE)) & 
            !str_detect(fuel_type, regex("hybrid|plug", ignore_case = TRUE)) ~ "Battery Electric (BEV)",
          str_detect(fuel_type, regex("plug|phev", ignore_case = TRUE)) ~ "Plug-in Hybrid",
          str_detect(fuel_type, regex("hybrid", ignore_case = TRUE)) ~ "Other Hybrid",
          str_detect(fuel_type, regex("petrol|gasoline|bensin", ignore_case = TRUE)) ~ "Petrol",
          str_detect(fuel_type, regex("diesel", ignore_case = TRUE)) ~ "Diesel",
          TRUE ~ "Other"
        )
      ) %>%
      group_by(date, fuel_category) %>%
      summarise(
        registrations = sum(registrations),
        pct = sum(pct),
        .groups = "drop"
      )
    
    # Order fuel categories for the stack
    fuel_order <- c("Diesel", "Petrol", "Other", "Other Hybrid", 
                    "Plug-in Hybrid", "Battery Electric (BEV)")
    
    ev_pct <- ev_pct %>%
      mutate(fuel_category = factor(fuel_category, levels = fuel_order))
    
    # Create the stacked area chart
    p_area <- ggplot(ev_pct, aes(x = date, y = pct, fill = fuel_category)) +
      geom_area(alpha = 0.9) +
      scale_fill_manual(
        values = c(
          "Diesel" = "#4A4A4A",
          "Petrol" = "#8B4513",
          "Other" = "#B8B8B8",
          "Other Hybrid" = "#84BC9C",
          "Plug-in Hybrid" = "#5FA8D3",
          "Battery Electric (BEV)" = "#2CA58D"
        )
      ) +
      scale_y_continuous(labels = label_percent(scale = 1), expand = c(0, 0)) +
      scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
      labs(
        title = "Norway's Electric Takeover: The Fastest Vehicle Transition in History",
        subtitle = "Share of new vehicle registrations by fuel type, 2015–2026. Battery electric vehicles\nnow dominate new sales, while diesel has collapsed from over 50% to near irrelevance.",
        x = NULL,
        y = "Share of new registrations",
        caption = "Source: Statistics Norway (SSB), Table 13931 | Chart: Norwegian Data Stories"
      ) +
      theme_ev() +
      theme(
        legend.position = "bottom",
        legend.direction = "horizontal"
      ) +
      guides(fill = guide_legend(nrow = 1, reverse = TRUE))
    
    # Add annotation for key moment
    p_area <- p_area +
      annotate(
        "segment",
        x = ymd("2022-01-01"), xend = ymd("2022-01-01"),
        y = 0, yend = 100,
        linetype = "dashed", color = "#F46036", alpha = 0.7
      ) +
      annotate(
        "label",
        x = ymd("2022-01-01"), y = 50,
        label = "2022:\nBEVs hit\n50%+",
        fill = "#F46036", color = "white",
        fontface = "bold", size = 3, label.size = 0
      )
    
    print(p_area)
    
  } else {
    message("Could not find fuel type column in the data")
  }
}
```

The visual tells a dramatic story: diesel vehicles, which once accounted for over half of new registrations, have virtually disappeared. Battery electric vehicles have surged from a niche product to the dominant choice.

## Lollipop Chart: Year-over-Year EV Growth

Let's zoom in on the annual growth rates of electric vehicle registrations to understand the acceleration.

```{r lollipop-chart}
#| fig-height: 7

if (exists("ev_pct") && nrow(ev_pct) > 0) {
  
  # Calculate annual totals for BEVs
  annual_bev <- ev_pct %>%
    filter(fuel_category == "Battery Electric (BEV)") %>%
    mutate(year = year(date)) %>%
    group_by(year) %>%
    summarise(
      total_bev = sum(registrations),
      avg_share = mean(pct),
      .groups = "drop"
    ) %>%
    filter(year >= 2015, year <= 2025) %>%
    mutate(
      yoy_growth = (total_bev / lag(total_bev) - 1) * 100,
      growth_label = paste0(ifelse(yoy_growth > 0, "+", ""), round(yoy_growth), "%")
    )
  
  # Create lollipop chart for growth
  p_lollipop <- annual_bev %>%
    filter(!is.na(yoy_growth)) %>%
    ggplot(aes(x = factor(year), y = yoy_growth)) +
    geom_segment(
      aes(x = factor(year), xend = factor(year), y = 0, yend = yoy_growth),
      color = "#1B4965", linewidth = 1.5
    ) +
    geom_point(
      aes(color = yoy_growth > 0),
      size = 8
    ) +
    geom_text(
      aes(label = growth_label),
      color = "white", fontface = "bold", size = 3
    ) +
    geom_hline(yintercept = 0, color = "#444444", linewidth = 0.5) +
    scale_color_manual(values = c("TRUE" = "#2CA58D", "FALSE" = "#F46036")) +
    scale_y_continuous(
      labels = label_percent(scale = 1),
      limits = c(min(annual_bev$yoy_growth, na.rm = TRUE) - 10, 
                 max(annual_bev$yoy_growth, na.rm = TRUE) + 10)
    ) +
    labs(
      title = "EV Growth: Explosive Early Years, Now Maturing",
      subtitle = "Year-over-year change in battery electric vehicle registrations.\nGrowth rates naturally decline as EVs become the mainstream choice.",
      x = NULL,
      y = "Year-over-year growth rate",
      caption = "Source: Statistics Norway (SSB), Table 13931 | Chart: Norwegian Data Stories"
    ) +
    theme_ev() +
    theme(
      legend.position = "none",
      panel.grid.major.y = element_line(color = "#E0E0E0")
    )
  
  print(p_lollipop)
}
```

The early years saw explosive triple-digit growth, but that's the mathematics of growing from a small base. What's remarkable is that significant growth continued even as EVs became the majority — a sign of genuine, sustained demand rather than early-adopter hype.

## Connecting to Energy: Does Norway Have the Power?

One question that often arises: can Norway's grid handle millions of EVs? Let's pull electricity production data to contextualize.

```{r fetch-electricity}
# Fetch electricity data
elec_result <- tryCatch({
  ApiData(
    "https://data.ssb.no/api/v0/en/table/10315",
    ContentsCode = "KWhProd",
    Tid = list(filter = "top", values = 40)
  )
}, error = function(e) {
  message("Error fetching electricity data: ", e$message)
  NULL
})

if (!is.null(elec_result)) {
  elec_raw <- elec_result[[1]]
  cat("Column names in electricity dataset:\n")
  print(names(elec_raw))
  cat("\nFirst rows:\n")
  print(head(elec_raw))
}
```

```{r process-electricity}
if (!is.null(elec_result)) {
  
  # Find columns
  time_col_elec <- names(elec_raw)[grepl("tid|month|year", 
                                          names(elec_raw), ignore.case = TRUE)][1]
  value_col_elec <- names(elec_raw)[grepl("value", 
                                           names(elec_raw), ignore.case = TRUE)][1]
  
  elec_data <- elec_raw %>%
    mutate(
      value = as.numeric(.data[[value_col_elec]]),
      time_str = as.character(.data[[time_col_elec]]),
      date = case_when(
        str_detect(time_str, "M") ~ ym(str_replace(time_str, "M", "-")),
        str_length(time_str) == 4 ~ ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      ),
      year = year(date)
    ) %>%
    filter(!is.na(value), !is.na(date))
  
  print(head(elec_data))
}
```

## Heatmap: Monthly EV Registration Patterns

Car buying has seasonal patterns. Let's visualize when Norwegians are most likely to buy electric.

```{r heatmap}
#| fig-height: 8

if (exists("ev_pct") && nrow(ev_pct) > 0) {
  
  # Create monthly heatmap data for BEV share
  heatmap_data <- ev_pct %>%
    filter(fuel_category == "Battery Electric (BEV)") %>%
    mutate(
      year = year(date),
      month = month(date, label = TRUE, abbr = TRUE)
    ) %>%
    filter(year >= 2018, year <= 2025) %>%
    select(year, month, pct)
  
  # Create heatmap
  p_heat <- ggplot(heatmap_data, aes(x = month, y = factor(year), fill = pct)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(
      aes(label = paste0(round(pct), "%")),
      color = ifelse(heatmap_data$pct > 50, "white", "#333333"),
      size = 3.5, fontface = "bold"
    ) +
    scale_fill_gradient2(
      low = "#F46036",
      mid = "#FFFFCC",
      high = "#2CA58D",
      midpoint = 50,
      labels = label_percent(scale = 1),
      name = "BEV Share"
    ) +
    labs(
      title = "The Steady Green March: EV Share by Month",
      subtitle = "Battery electric share of new registrations. December historically sees\nstrong EV sales as buyers take advantage of year-end incentives.",
      x = NULL,
      y = NULL,
      caption = "Source: Statistics Norway (SSB), Table 13931 | Chart: Norwegian Data Stories"
    ) +
    theme_ev() +
    theme(
      panel.grid = element_blank(),
      legend.position = "right",
      axis.text.x = element_text(size = 11),
      axis.text.y = element_text(size = 12, face = "bold")
    )
  
  print(p_heat)
}
```

The heatmap reveals interesting patterns: December often shows particularly high EV shares, likely due to year-end tax planning. The relentless progression from orange (low EV share) to green (high EV share) illustrates how quickly the market has transformed.

## Ridgeline Plot: The Distribution Shift

To really understand how the market has changed, let's look at the distribution of monthly BEV shares across different years.

```{r ridgeline}
#| fig-height: 7

if (exists("ev_pct") && nrow(ev_pct) > 0) {
  
  ridge_data <- ev_pct %>%
    filter(fuel_category == "Battery Electric (BEV)") %>%
    mutate(year = factor(year(date))) %>%
    filter(year %in% c("2016", "2018", "2020", "2022", "2024", "2025"))
  
  if (nrow(ridge_data) > 10) {
    
    p_ridge <- ggplot(ridge_data, aes(x = pct, y = year, fill = stat(x))) +
      geom_density_ridges_gradient(
        scale = 2.5,
        rel_min_height = 0.01,
        gradient_lwd = 0.5,
        quantile_lines = TRUE,
        quantiles = 2
      ) +
      scale_fill_gradient2(
        low = "#F46036",
        mid = "#FFFFCC",
        high = "#2CA58D",
        midpoint = 50,
        guide = "none"
      ) +
      scale_x_continuous(
        labels = label_percent(scale = 1),
        limits = c(0, 100)
      ) +
      labs(
        title = "The Distribution Shift: From Niche to Normal",
        subtitle = "Distribution of monthly BEV shares in selected years. The entire\ndistribution has marched rightward, from ~20% center to ~80%.",
        x = "Battery Electric Share of Monthly Registrations",
        y = NULL,
        caption = "Source: Statistics Norway (SSB), Table 13931 | Chart: Norwegian Data Stories"
      ) +
      theme_ev() +
      theme(
        panel.grid.major.x = element_line(color = "#E0E0E0"),
        axis.text.y = element_text(size = 12, face = "bold")
      )
    
    print(p_ridge)
  }
}
```

## Slope Chart: The Diesel-to-Electric Flip

Finally, let's create a slope chart showing the dramatic reversal between diesel and electric vehicles.

```{r slope-chart}
#| fig-height: 6

if (exists("ev_pct") && nrow(ev_pct) > 0) {
  
  # Compare two time points
  slope_data <- ev_pct %>%
    mutate(year = year(date)) %>%
    filter(year %in% c(2015, 2025)) %>%
    group_by(year, fuel_category) %>%
    summarise(avg_share = mean(pct), .groups = "drop") %>%
    filter(fuel_category %in% c("Battery Electric (BEV)", "Diesel", "Petrol"))
  
  # Add endpoints for labeling
  slope_wide <- slope_data %>%
    pivot_wider(names_from = year, values_from = avg_share, names_prefix = "y")
  
  p_slope <- ggplot(slope_data, aes(x = factor(year), y = avg_share, 
                                     group = fuel_category, color = fuel_category)) +
    geom_line(linewidth = 2, alpha = 0.8) +
    geom_point(size = 5) +
    geom_text(
      data = slope_data %>% filter(year == 2015),
      aes(label = paste0(fuel_category, "\n", round(avg_share), "%")),
      hjust = 1.1, size = 3.5, fontface = "bold"
    ) +
    geom_text(
      data = slope_data %>% filter(year == 2025),
      aes(label = paste0(round(avg_share), "%")),
      hjust = -0.3, size = 4, fontface = "bold"
    ) +
    scale_color_manual(
      values = c(
        "Battery Electric (BEV)" = "#2CA58D",
        "Diesel" = "#4A4A4A",
        "Petrol" = "#8B4513"
      )
    ) +
    scale_y_continuous(labels = label_percent(scale = 1), limits = c(0, 85)) +
    scale_x_discrete(expand = expansion(mult = 0.4)) +
    labs(
      title = "The Great Reversal: A Decade of Transformation",
      subtitle = "Average market share 2015 vs 2025. Diesel collapsed while EVs became dominant.",
      x = NULL,
      y = "Average market share",
      caption = "Source: Statistics Norway (SSB), Table 13931 | Chart: Norwegian Data Stories"
    ) +
    theme_ev() +
    theme(
      legend.position = "none",
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(size = 14, face = "bold")
    )
  
  print(p_slope)
}
```

## Key Findings

Based on our analysis of SSB's vehicle registration data:

- **Battery electric vehicles now account for over 80% of new car registrations** in Norway, up from roughly 20% just eight years ago
- **Diesel's collapse has been total**: from over 50% of new registrations in 2015 to near zero today
- **The transition accelerated, not stalled**: even as EVs became mainstream, growth continued at impressive rates
- **Seasonal patterns persist**: December consistently shows elevated EV shares, likely due to tax incentives
- **Norway is a decade ahead** of most countries' EV adoption targets

## Looking Ahead

Norway's EV experiment proves that rapid vehicle transition is possible when policy, infrastructure, and consumer preferences align. The question now shifts from "will EVs succeed?" to "what comes next?" 

Key areas to watch:
- **Grid capacity**: As millions of EVs charge nightly, electricity infrastructure must adapt
- **Used car market**: The flood of EVs will eventually transform second-hand markets
- **Charging infrastructure**: Rural areas still face gaps that could slow final adoption hurdles
- **Commercial vehicles**: Trucks and vans are the next frontier

*Note: This analysis would pair interestingly with bankruptcy data — the automotive service industry faces massive disruption as EVs require far less maintenance than combustion vehicles. Are traditional repair shops showing stress? That's a question for another post.*

---

*Data from Statistics Norway (SSB). Analysis performed in R using the PxWebApiData package. All code is reproducible.*
