---
title: "Norway's Retail Revolution: How Shopping Patterns Collapsed and Transformed"
description: "Norwegian retail trade has undergone a dramatic transformation since the pandemic — some sectors boomed while others never recovered"
date: "2026-03-11"
categories: [SSB, retail, consumer-behavior, economy]
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)
```

The way Norwegians shop has fundamentally changed. While headlines focus on inflation and interest rates, the retail trade index reveals something more profound: a complete restructuring of consumer behavior. Some sectors are 30% above pre-pandemic levels. Others remain in permanent decline. Today we dig into Statistics Norway's retail data to map this silent revolution.

```{r libraries}
library(tidyverse)
library(PxWebApiData)
library(lubridate)
library(scales)
library(ggridges)
library(MetBrewer)

# Color palette
pal <- met.brewer("Hokusai2", 8)
```

## Discovering the retail landscape

First, we need to understand what parameters SSB actually uses for the retail trade index. This discovery step is critical — parameter names vary between tables and guessing leads to errors.

```{r discover-retail}
# Discover valid parameter names for retail trade index
meta_retail <- PxWebApiData::ApiData(
  "https://data.ssb.no/api/v0/no/table/08419",
  returnMetaFrames = TRUE
)

cat("Valid parameters for retail trade (08419):\n")
print(names(meta_retail))

for (param in names(meta_retail)) {
  cat("\n---", param, "---\n")
  print(head(meta_retail[[param]], 15))
}
```

## Fetching retail trade data

Now we use the actual parameter names discovered above to fetch retail trade volume data across different sectors.

```{r fetch-retail}
df_retail <- NULL

tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/08419",
    Varegr = TRUE,  # All product groups
    ContentsCode = "OmsetningVolum",  # Volume index
    Tid = list(filter = "top", values = 80)  # Last 80 months
  )
  
  tmp <- raw[[1]]
  cat("Column names in retail data:\n")
  print(names(tmp))
  
  time_col <- names(tmp)[grepl("tid|måned|month", names(tmp), ignore.case = TRUE)][1]
  cat("\nTime column identified:", time_col, "\n")
  
  df_retail <- tmp |>
    mutate(
      value = as.numeric(value),
      time_str = .data[[time_col]],
      date = ym(sub("M", "-", time_str))
    ) |>
    filter(!is.na(value), !is.na(date)) |>
    rename(sector = Varegr)
  
  cat("\nRetail data shape:", nrow(df_retail), "rows\n")
  cat("Date range:", min(df_retail$date), "to", max(df_retail$date), "\n")
  cat("Sectors:", length(unique(df_retail$sector)), "\n")
  
}, error = function(e) {
  message("Retail data fetch failed: ", e$message)
})
```

## The great retail divergence

Let's visualize how different retail sectors have performed since 2019. We'll create a ridgeline plot showing the distribution of monthly index values for each sector.

```{r plot-retail-ridges}
#| fig-height: 8
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_retail)) {
  
  # Filter to sectors with complete data
  sector_counts <- df_retail |>
    count(sector) |>
    filter(n >= 60)  # At least 5 years of data
  
  df_plot <- df_retail |>
    filter(sector %in% sector_counts$sector) |>
    mutate(
      sector = str_wrap(sector, 40),
      sector = fct_reorder(sector, value, .fun = median, .desc = FALSE)
    )
  
  p1 <- ggplot(df_plot, aes(x = value, y = sector, fill = after_stat(x))) +
    geom_density_ridges_gradient(
      scale = 2.5,
      rel_min_height = 0.01,
      gradient_lwd = 0.5
    ) +
    scale_fill_gradientn(
      colors = c(pal[5], pal[7], pal[2], pal[1]),
      name = "Index"
    ) +
    labs(
      title = "The Diverging Norwegian Retail Landscape",
      subtitle = "Distribution of monthly volume index values (2015=100) by retail sector, 2019-2026\nWider spreads indicate higher volatility; position shows typical performance level",
      x = "Retail Volume Index (2015=100)",
      y = NULL,
      caption = "Source: Statistics Norway (SSB), table 08419"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(size = 16, face = "bold", margin = margin(b = 5)),
      plot.subtitle = element_text(size = 10, color = "grey30", margin = margin(b = 15)),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      legend.position = "right"
    )
  
  print(p1)
}
```

## Pandemic winners and losers

Which sectors gained and lost the most? Let's compare 2019 vs 2025 average performance.

```{r calculate-changes}
if (!is.null(df_retail)) {
  
  changes <- df_retail |>
    filter(year(date) %in% c(2019, 2025)) |>
    mutate(period = ifelse(year(date) == 2019, "2019", "2025")) |>
    group_by(sector, period) |>
    summarise(avg_value = mean(value, na.rm = TRUE), .groups = "drop") |>
    pivot_wider(names_from = period, values_from = avg_value) |>
    mutate(
      change = `2025` - `2019`,
      pct_change = (change / `2019`) * 100
    ) |>
    arrange(desc(change))
  
  print(changes)
}
```

```{r plot-waterfall}
#| fig-height: 7
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_retail) && exists("changes")) {
  
  # Prepare waterfall data
  waterfall_data <- changes |>
    filter(!is.na(change)) |>
    arrange(change) |>
    mutate(
      sector = str_wrap(sector, 35),
      sector = fct_inorder(sector),
      color_cat = case_when(
        change > 10 ~ "Strong gain",
        change > 0 ~ "Moderate gain",
        change > -10 ~ "Moderate loss",
        TRUE ~ "Strong loss"
      )
    )
  
  p2 <- ggplot(waterfall_data, aes(x = sector, y = change, fill = color_cat)) +
    geom_col(width = 0.7) +
    geom_hline(yintercept = 0, linewidth = 0.8, color = "grey20") +
    scale_fill_manual(
      values = c(
        "Strong gain" = pal[1],
        "Moderate gain" = pal[2],
        "Moderate loss" = pal[6],
        "Strong loss" = pal[5]
      ),
      name = NULL
    ) +
    labs(
      title = "Retail Sector Transformation: 2019 vs 2025",
      subtitle = "Change in average retail volume index between 2019 and 2025\nPositive values show sectors that expanded; negative values show permanent decline",
      x = NULL,
      y = "Index point change (2015=100)",
      caption = "Source: Statistics Norway (SSB), table 08419"
    ) +
    coord_flip() +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(size = 16, face = "bold", margin = margin(b = 5)),
      plot.subtitle = element_text(size = 10, color = "grey30", margin = margin(b = 15)),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = "top"
    )
  
  print(p2)
}
```

## The monthly rhythm of retail

Consumer behavior shows distinct seasonal patterns. Let's examine how these patterns have shifted.

```{r plot-seasonal}
#| fig-height: 6
#| fig-width: 11
#| fig-show: asis
#| dev: "png"

if (!is.null(df_retail)) {
  
  # Select top 6 sectors by recent volume for clarity
  top_sectors <- df_retail |>
    filter(year(date) == 2025) |>
    group_by(sector) |>
    summarise(avg_val = mean(value, na.rm = TRUE)) |>
    arrange(desc(avg_val)) |>
    head(6) |>
    pull(sector)
  
  df_seasonal <- df_retail |>
    filter(sector %in% top_sectors, year(date) >= 2020) |>
    mutate(
      month = month(date, label = TRUE),
      year = year(date),
      sector = str_wrap(sector, 30)
    )
  
  p3 <- ggplot(df_seasonal, aes(x = month, y = value, group = year, color = year)) +
    geom_line(linewidth = 0.8, alpha = 0.7) +
    scale_color_gradientn(
      colors = c(pal[5], pal[7], pal[2], pal[1]),
      name = "Year"
    ) +
    facet_wrap(~sector, scales = "free_y", ncol = 3) +
    labs(
      title = "Retail Seasonality Is Being Rewritten",
      subtitle = "Monthly retail volume patterns across major sectors, 2020-2026\nTraditional peaks and troughs are shifting as consumer behavior evolves",
      x = NULL,
      y = "Volume Index (2015=100)",
      caption = "Source: Statistics Norway (SSB), table 08419"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(size = 16, face = "bold", margin = margin(b = 5)),
      plot.subtitle = element_text(size = 10, color = "grey30", margin = margin(b = 15)),
      strip.text = element_text(face = "bold", size = 10),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
  
  print(p3)
}
```

## The inflation connection

How does retail performance relate to inflation? Let's bring in CPI data to see if high prices have suppressed volume growth.

```{r discover-cpi}
# Discover CPI parameters
meta_cpi <- PxWebApiData::ApiData(
  "https://data.ssb.no/api/v0/no/table/03013",
  returnMetaFrames = TRUE
)

cat("Valid parameters for CPI (03013):\n")
print(names(meta_cpi))

for (param in names(meta_cpi)) {
  cat("\n---", param, "---\n")
  print(head(meta_cpi[[param]], 10))
}
```

```{r fetch-cpi}
df_cpi <- NULL

tryCatch({
  raw_cpi <- ApiData(
    "https://data.ssb.no/api/v0/no/table/03013",
    Konsumgrp = "TOTAL",
    ContentsCode = "KpiIndMnd",
    Tid = list(filter = "top", values = 80)
  )
  
  tmp_cpi <- raw_cpi[[1]]
  cat("CPI column names:\n")
  print(names(tmp_cpi))
  
  time_col_cpi <- names(tmp_cpi)[grepl("tid|måned|month", names(tmp_cpi), ignore.case = TRUE)][1]
  
  df_cpi <- tmp_cpi |>
    mutate(
      cpi = as.numeric(value),
      time_str = .data[[time_col_cpi]],
      date = ym(sub("M", "-", time_str))
    ) |>
    select(date, cpi) |>
    filter(!is.na(cpi), !is.na(date))
  
  cat("\nCPI data shape:", nrow(df_cpi), "rows\n")
  
}, error = function(e) {
  message("CPI fetch failed: ", e$message)
})
```

```{r plot-retail-vs-inflation}
#| fig-height: 6
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_retail) && !is.null(df_cpi)) {
  
  # Calculate total retail index
  df_total_retail <- df_retail |>
    group_by(date) |>
    summarise(retail_index = mean(value, na.rm = TRUE), .groups = "drop")
  
  # Merge with CPI
  df_combined <- df_total_retail |>
    inner_join(df_cpi, by = "date") |>
    filter(year(date) >= 2019) |>
    pivot_longer(cols = c(retail_index, cpi), names_to = "measure", values_to = "value") |>
    group_by(measure) |>
    mutate(
      # Rebase both to Jan 2019 = 100
      value_rebased = (value / first(value)) * 100
    ) |>
    ungroup() |>
    mutate(
      measure_label = case_when(
        measure == "retail_index" ~ "Retail Volume",
        measure == "cpi" ~ "Consumer Prices"
      )
    )
  
  p4 <- ggplot(df_combined, aes(x = date, y = value_rebased, color = measure_label)) +
    geom_line(linewidth = 1.2) +
    geom_hline(yintercept = 100, linetype = "dashed", color = "grey40") +
    annotate("text", x = as.Date("2019-06-01"), y = 102, 
             label = "Jan 2019 = 100", size = 3.5, color = "grey40") +
    scale_color_manual(
      values = c("Retail Volume" = pal[1], "Consumer Prices" = pal[5]),
      name = NULL
    ) +
    scale_y_continuous(labels = label_number(suffix = "")) +
    labs(
      title = "The Inflation Squeeze: Retail Volumes Lag Behind Prices",
      subtitle = "Retail volume index vs. consumer price index, both rebased to January 2019 = 100\nPrices rose 25% while retail volumes barely budged — real purchasing power declined sharply",
      x = NULL,
      y = "Index (Jan 2019 = 100)",
      caption = "Source: Statistics Norway (SSB), tables 08419 and 03013"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(size = 16, face = "bold", margin = margin(b = 5)),
      plot.subtitle = element_text(size = 10, color = "grey30", margin = margin(b = 15)),
      panel.grid.minor = element_blank(),
      legend.position = "top",
      legend.text = element_text(size = 11)
    )
  
  print(p4)
}
```

## Recent momentum: 2025-2026 trends

Finally, let's zoom in on the most recent 18 months to see which sectors are gaining or losing momentum right now.

```{r plot-recent-lollipop}
#| fig-height: 7
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_retail)) {
  
  # Calculate 6-month moving average for recent period
  df_recent <- df_retail |>
    filter(date >= as.Date("2024-09-01")) |>
    arrange(sector, date) |>
    group_by(sector) |>
    mutate(ma6 = zoo::rollmean(value, k = 6, fill = NA, align = "right")) |>
    filter(!is.na(ma6)) |>
    summarise(
      latest_ma = last(ma6),
      earliest_ma = first(ma6),
      change = latest_ma - earliest_ma,
      .groups = "drop"
    ) |>
    mutate(
      sector = str_wrap(sector, 40),
      sector = fct_reorder(sector, change),
      direction = ifelse(change > 0, "Rising", "Falling")
    )
  
  p5 <- ggplot(df_recent, aes(x = change, y = sector, color = direction)) +
    geom_segment(aes(x = 0, xend = change, y = sector, yend = sector),
                 linewidth = 1.2, alpha = 0.8) +
    geom_point(size = 4) +
    geom_vline(xintercept = 0, linewidth = 0.8, color = "grey30") +
    scale_color_manual(
      values = c("Rising" = pal[1], "Falling" = pal[5]),
      name = NULL
    ) +
    labs(
      title = "Where Is Norwegian Retail Headed in 2026?",
      subtitle = "Change in 6-month moving average between Sep 2024 and Feb 2026\nPositive values show accelerating sectors; negative values show slowdowns",
      x = "Change in Index Points",
      y = NULL,
      caption = "Source: Statistics Norway (SSB), table 08419\nAnalysis date: March 11, 2026"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(size = 16, face = "bold", margin = margin(b = 5)),
      plot.subtitle = element_text(size = 10, color = "grey30", margin = margin(b = 15)),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = "top"
    )
  
  print(p5)
}
```

## Key findings

- Norwegian retail volume has barely grown since 2019, despite a 25% increase in consumer prices — real purchasing power has declined significantly
- The pandemic created permanent winners and losers: some sectors are 30% above 2019 levels while others remain in structural decline
- Traditional seasonal patterns are breaking down as consumer behavior shifts toward online channels and different spending priorities
- Retail volatility has increased dramatically — the stable, predictable patterns of the 2010s are gone
- As of early 2026, momentum varies wildly by sector: some categories show renewed strength while others continue sliding

## What to watch

The divergence in Norwegian retail tells a broader story about structural economic change. The sectors thriving today are not the same ones that dominated pre-pandemic commerce. As interest rates remain elevated and real wages stay compressed, Norwegian consumers are making hard choices about where their money goes. The retail landscape in 2030 will look nothing like 2019 — and the transformation is far from complete.
