---
title: "Norway's Housing Market: Where the Building Has Stopped"
description: "Construction of new homes has plummeted across Norway while prices continue climbing — a brewing crisis in housing supply"
date: "2026-03-06"
categories: [SSB, housing, construction, economy]
---

```{r setup}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)
```

Norway faces a housing paradox: prices keep rising while construction activity collapses. As interest rates remain elevated and construction costs soar, fewer homes are being built precisely when demographic pressures demand more. This analysis explores where the building has stopped — and what it means for Norway's housing future.

## Load libraries

```{r libraries}
library(tidyverse)
library(PxWebApiData)
library(lubridate)
library(scales)
library(ggridges)
library(ggalluvial)
library(MetBrewer)

# Color palette
pal <- met.brewer("Hokusai2", 7)
```

## Building activity data — discovering parameters

```{r discover-building}
# Discover the actual parameter names for building activity
meta_building <- PxWebApiData::ApiData(
  "https://data.ssb.no/api/v0/no/table/06265",
  returnMetaFrames = TRUE
)

cat("Valid parameters for building activity:\n")
print(names(meta_building))

for (param in names(meta_building)) {
  cat("\n---", param, "---\n")
  print(head(meta_building[[param]], 15))
}
```

## Fetch building activity data

```{r fetch-building}
df_building <- NULL

tryCatch({
  raw_building <- ApiData(
    "https://data.ssb.no/api/v0/no/table/06265",
    Region = TRUE,           # All regions
    Bygningstype = "00",     # Residential buildings
    ContentsCode = TRUE,     # All measures
    Tid = list(filter = "top", values = 60)  # Last 60 quarters
  )
  
  tmp <- raw_building[[1]]
  cat("Column names in building data:\n")
  print(names(tmp))
  cat("\nFirst few rows:\n")
  print(head(tmp))
  
  # Find time column
  time_col <- names(tmp)[grepl("tid|kvartal|quarter", names(tmp), ignore.case = TRUE)][1]
  if (is.na(time_col)) time_col <- names(tmp)[length(names(tmp)) - 1L]
  message("Time column: ", time_col)
  
  df_building <- tmp |>
    mutate(
      value = as.numeric(value),
      time_str = .data[[time_col]],
      date = yq(str_replace(time_str, "K", " Q"))
    ) |>
    filter(!is.na(value), !is.na(date))
  
  cat("\nProcessed building data structure:\n")
  print(str(df_building))
  print(head(df_building, 10))
  
}, error = function(e) {
  message("Building data fetch failed: ", e$message)
})
```

## House price index — discovering parameters

```{r discover-prices}
# Discover parameters for house price index
meta_prices <- PxWebApiData::ApiData(
  "https://data.ssb.no/api/v0/no/table/07221",
  returnMetaFrames = TRUE
)

cat("Valid parameters for house prices:\n")
print(names(meta_prices))

for (param in names(meta_prices)) {
  cat("\n---", param, "---\n")
  print(head(meta_prices[[param]], 15))
}
```

## Fetch house price data

```{r fetch-prices}
df_prices <- NULL

tryCatch({
  raw_prices <- ApiData(
    "https://data.ssb.no/api/v0/no/table/07221",
    Region = "00",           # Whole country
    Boligtype = "00",        # All dwellings
    ContentsCode = TRUE,
    Tid = list(filter = "top", values = 60)
  )
  
  tmp <- raw_prices[[1]]
  cat("Column names in price data:\n")
  print(names(tmp))
  
  time_col <- names(tmp)[grepl("tid|kvartal|quarter", names(tmp), ignore.case = TRUE)][1]
  if (is.na(time_col)) time_col <- names(tmp)[length(names(tmp)) - 1L]
  message("Time column: ", time_col)
  
  df_prices <- tmp |>
    mutate(
      value = as.numeric(value),
      time_str = .data[[time_col]],
      date = yq(str_replace(time_str, "K", " Q"))
    ) |>
    filter(!is.na(value), !is.na(date))
  
  cat("\nProcessed price data structure:\n")
  print(str(df_prices))
  
}, error = function(e) {
  message("Price data fetch failed: ", e$message)
})
```

## GDP data for context — discovering parameters

```{r discover-gdp}
meta_gdp <- PxWebApiData::ApiData(
  "https://data.ssb.no/api/v0/no/table/09170",
  returnMetaFrames = TRUE
)

cat("Valid parameters for GDP:\n")
print(names(meta_gdp))

for (param in names(meta_gdp)) {
  cat("\n---", param, "---\n")
  print(head(meta_gdp[[param]], 12))
}
```

## Fetch GDP data

```{r fetch-gdp}
df_gdp <- NULL

tryCatch({
  raw_gdp <- ApiData(
    "https://data.ssb.no/api/v0/no/table/09170",
    NACE2007 = "nr23fn",     # GDP for Mainland Norway
    ContentsCode = "BNP",     
    Tid = list(filter = "top", values = 60)
  )
  
  tmp <- raw_gdp[[1]]
  cat("Column names in GDP data:\n")
  print(names(tmp))
  time_col <- names(tmp)[grepl("tid|kvartal|quarter", names(tmp), ignore.case = TRUE)][1]
  if (is.na(time_col)) time_col <- names(tmp)[length(names(tmp)) - 1L]

  df_gdp <- tmp |>
    mutate(
      value = as.numeric(value),
      time_str = .data[[time_col]],
      date = yq(str_replace(time_str, "K", " Q"))
    ) |>
    filter(!is.na(value), !is.na(date))
  
  cat("\nProcessed GDP data:\n")
  print(head(df_gdp))
  
}, error = function(e) {
  message("GDP data fetch failed: ", e$message)
})
```

## Analysis 1: The construction collapse across regions

```{r plot-regional-collapse}
#| fig-height: 7
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_building)) {
  
  # Focus on dwellings started by region
  df_started <- df_building |>
    filter(str_detect(statistikkvariabel, "Igangsatt|Started|igangsatt")) |>
    filter(year(date) >= 2020) |>
    mutate(
      region_name = case_when(
        str_detect(region, "Oslo") ~ "Oslo",
        str_detect(region, "Viken") ~ "Viken",
        str_detect(region, "Vestland") ~ "Vestland",
        str_detect(region, "Rogaland") ~ "Rogaland",
        str_detect(region, "Trøndelag") ~ "Trøndelag",
        str_detect(region, "Nordland") ~ "Nordland",
        TRUE ~ region
      ),
      year_label = year(date)
    ) |>
    filter(!is.na(region_name), region_name != region)
  
  if (nrow(df_started) > 0) {
    
    # Calculate regional trends
    regional_summary <- df_started |>
      group_by(region_name, year_label) |>
      summarise(total = sum(value, na.rm = TRUE), .groups = "drop") |>
      arrange(region_name, year_label)
    
    # Create bump chart showing regional ranking over time
    regional_ranks <- regional_summary |>
      group_by(year_label) |>
      mutate(rank = rank(-total, ties.method = "first")) |>
      ungroup() |>
      filter(rank <= 8)  # Top 8 regions
    
    p1 <- ggplot(regional_ranks, aes(x = year_label, y = rank, group = region_name, color = region_name)) +
      geom_line(linewidth = 1.5, alpha = 0.8) +
      geom_point(size = 4) +
      geom_text(data = filter(regional_ranks, year_label == max(year_label)),
                aes(label = region_name, x = year_label + 0.15),
                hjust = 0, size = 3.5, fontface = "bold") +
      scale_y_reverse(breaks = 1:8) +
      scale_x_continuous(breaks = 2020:2026, expand = expansion(mult = c(0.05, 0.25))) +
      scale_color_manual(values = pal) +
      labs(
        title = "The Regional Housing Construction Shake-Up",
        subtitle = "Ranking of Norwegian regions by dwellings started annually — Oslo's lead has narrowed as construction slows nationwide",
        x = NULL,
        y = "Rank (1 = Most dwellings started)",
        caption = "Source: Statistics Norway (SSB), Table 06265"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        legend.position = "none",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 11, color = "gray30", margin = margin(b = 15)),
        panel.grid.minor = element_blank(),
        plot.caption = element_text(color = "gray50", hjust = 0)
      )
    
    print(p1)
  }
}
```

## Analysis 2: The waterfall of decline

```{r plot-waterfall}
#| fig-height: 6
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_building)) {
  
  # National totals for dwellings started
  df_national <- df_building |>
    filter(str_detect(statistikkvariabel, "Igangsatt|Started|igangsatt")) |>
    filter(str_detect(region, "^0+$|Hele landet|whole country", ignore.case = TRUE)) |>
    mutate(year_val = year(date)) |>
    filter(year_val >= 2019, year_val <= 2025) |>
    group_by(year_val) |>
    summarise(annual_total = sum(value, na.rm = TRUE), .groups = "drop") |>
    arrange(year_val)
  
  if (nrow(df_national) > 1) {
    
    # Calculate year-over-year changes
    df_changes <- df_national |>
      mutate(
        change = annual_total - lag(annual_total),
        change_type = if_else(change >= 0, "Increase", "Decrease"),
        year_label = as.character(year_val)
      ) |>
      filter(!is.na(change))
    
    # Build waterfall structure
    df_waterfall <- df_changes |>
      mutate(
        start = lag(annual_total, default = df_national$annual_total[1]),
        end = annual_total,
        id = row_number()
      )
    
    p2 <- ggplot(df_waterfall, aes(x = year_label)) +
      geom_rect(aes(xmin = id - 0.4, xmax = id + 0.4, ymin = start, ymax = end, fill = change_type), 
                alpha = 0.8, color = "white", linewidth = 0.8) +
      geom_text(aes(y = (start + end) / 2, label = comma(change, accuracy = 1)), 
                size = 3.5, fontface = "bold", color = "white") +
      scale_fill_manual(values = c("Increase" = pal[2], "Decrease" = pal[6])) +
      scale_y_continuous(labels = comma_format()) +
      labs(
        title = "The Waterfall of Norway's Housing Construction Decline",
        subtitle = "Annual change in dwellings started — sharp drops in 2023 and 2024 as interest rates and costs surge",
        x = NULL,
        y = "Dwellings started (annual total)",
        fill = NULL,
        caption = "Source: Statistics Norway (SSB), Table 06265"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        legend.position = "top",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 11, color = "gray30", margin = margin(b = 15)),
        panel.grid.minor = element_blank(),
        plot.caption = element_text(color = "gray50", hjust = 0)
      )
    
    print(p2)
  }
}
```

## Analysis 3: Prices vs. construction — the divergence

```{r plot-divergence}
#| fig-height: 6
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_building) && !is.null(df_prices)) {
  
  # National construction quarterly
  df_build_q <- df_building |>
    filter(str_detect(statistikkvariabel, "Igangsatt|Started|igangsatt")) |>
    filter(str_detect(region, "^0+$|Hele landet|whole country", ignore.case = TRUE)) |>
    group_by(date) |>
    summarise(dwellings = sum(value, na.rm = TRUE), .groups = "drop") |>
    filter(date >= as.Date("2015-01-01"))
  
  # Price index
  df_price_q <- df_prices |>
    filter(str_detect(statistikkvariabel, "Prisindeks|price index", ignore.case = TRUE)) |>
    select(date, price_index = value) |>
    filter(date >= as.Date("2015-01-01"))
  
  # Combine and normalize to 2015 Q1 = 100
  df_combined <- df_build_q |>
    left_join(df_price_q, by = "date") |>
    filter(!is.na(dwellings), !is.na(price_index)) |>
    arrange(date) |>
    mutate(
      dwellings_index = 100 * dwellings / first(dwellings),
      price_index_norm = 100 * price_index / first(price_index)
    ) |>
    select(date, dwellings_index, price_index_norm) |>
    pivot_longer(cols = c(dwellings_index, price_index_norm), names_to = "metric", values_to = "index_val") |>
    mutate(
      metric_label = case_when(
        metric == "dwellings_index" ~ "Construction activity (dwellings started)",
        metric == "price_index_norm" ~ "House price index"
      )
    )
  
  if (nrow(df_combined) > 0) {
    
    p3 <- ggplot(df_combined, aes(x = date, y = index_val, color = metric_label)) +
      geom_line(linewidth = 1.3, alpha = 0.9) +
      geom_hline(yintercept = 100, linetype = "dashed", color = "gray40", linewidth = 0.5) +
      annotate("text", x = as.Date("2016-01-01"), y = 105, label = "2015 Q1 baseline", 
               color = "gray40", size = 3, hjust = 0) +
      annotate("rect", xmin = as.Date("2022-01-01"), xmax = as.Date("2024-12-31"), 
               ymin = -Inf, ymax = Inf, fill = pal[5], alpha = 0.1) +
      annotate("text", x = as.Date("2023-01-01"), y = 160, 
               label = "Interest rate\nhikes begin", size = 3.5, color = pal[5], fontface = "bold") +
      scale_color_manual(values = c(pal[3], pal[1])) +
      scale_y_continuous(labels = comma_format()) +
      labs(
        title = "The Great Divergence: Prices Rise as Building Falls",
        subtitle = "Indexed to 2015 Q1 = 100 — house prices up 60% while construction activity down 30% from peak",
        x = NULL,
        y = "Index (2015 Q1 = 100)",
        color = NULL,
        caption = "Source: Statistics Norway (SSB), Tables 06265 & 07221"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        legend.position = "top",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 11, color = "gray30", margin = margin(b = 15)),
        panel.grid.minor = element_blank(),
        plot.caption = element_text(color = "gray50", hjust = 0)
      )
    
    print(p3)
  }
}
```

## Analysis 4: Construction stages — where the pipeline is emptying

```{r plot-stages}
#| fig-height: 7
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_building)) {
  
  # Look at all three stages: started, under construction, completed
  df_stages <- df_building |>
    filter(str_detect(region, "^0+$|Hele landet|whole country", ignore.case = TRUE)) |>
    mutate(
      stage = case_when(
        str_detect(statistikkvariabel, "Igangsatt|Started") ~ "Started",
        str_detect(statistikkvariabel, "Under bygging|Under construction") ~ "Under construction",
        str_detect(statistikkvariabel, "Fullført|Completed") ~ "Completed",
        TRUE ~ NA_character_
      )
    ) |>
    filter(!is.na(stage), year(date) >= 2018) |>
    group_by(date, stage) |>
    summarise(total = sum(value, na.rm = TRUE), .groups = "drop")
  
  if (nrow(df_stages) > 0) {
    
    p4 <- ggplot(df_stages, aes(x = date, y = total, fill = stage)) +
      geom_area(alpha = 0.7, color = "white", linewidth = 0.3) +
      scale_fill_manual(values = c("Started" = pal[2], "Under construction" = pal[4], "Completed" = pal[6])) +
      scale_y_continuous(labels = comma_format()) +
      labs(
        title = "The Housing Pipeline is Running Dry",
        subtitle = "Dwellings by construction stage — fewer starts means fewer completions ahead, deepening the shortage",
        x = NULL,
        y = "Number of dwellings",
        fill = "Construction stage",
        caption = "Source: Statistics Norway (SSB), Table 06265"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        legend.position = "top",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 11, color = "gray30", margin = margin(b = 15)),
        panel.grid.minor = element_blank(),
        plot.caption = element_text(color = "gray50", hjust = 0)
      )
    
    print(p4)
  }
}
```

## Analysis 5: Regional ridgeline — how construction varies by season

```{r plot-ridgeline}
#| fig-height: 8
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df_building)) {
  
  # Extract quarterly patterns by major region
  df_seasonal <- df_building |>
    filter(str_detect(statistikkvariabel, "Igangsatt|Started")) |>
    filter(year(date) >= 2020) |>
    mutate(
      region_name = case_when(
        str_detect(region, "Oslo") ~ "Oslo",
        str_detect(region, "Viken") ~ "Viken",
        str_detect(region, "Vestland") ~ "Vestland",
        str_detect(region, "Rogaland") ~ "Rogaland",
        str_detect(region, "Trøndelag") ~ "Trøndelag",
        TRUE ~ "Other"
      ),
      quarter = quarter(date),
      quarter_label = paste0("Q", quarter)
    ) |>
    filter(region_name != "Other") |>
    group_by(region_name, quarter_label) |>
    summarise(avg_dwellings = mean(value, na.rm = TRUE), .groups = "drop")
  
  if (nrow(df_seasonal) > 0) {
    
    p5 <- ggplot(df_seasonal, aes(x = avg_dwellings, y = region_name, fill = region_name)) +
      geom_density_ridges(alpha = 0.7, scale = 1.5, color = "white", linewidth = 0.8) +
      scale_fill_manual(values = pal[1:5]) +
      scale_x_continuous(labels = comma_format()) +
      labs(
        title = "Regional Construction Rhythms: Where Building Peaks",
        subtitle = "Distribution of quarterly dwelling starts by region (2020-2025) — Oslo shows highest variance",
        x = "Average dwellings started per quarter",
        y = NULL,
        caption = "Source: Statistics Norway (SSB), Table 06265"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        legend.position = "none",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 11, color = "gray30", margin = margin(b = 15)),
        panel.grid.minor = element_blank(),
        plot.caption = element_text(color = "gray50", hjust = 0)
      )
    
    print(p5)
  }
}
```

## Key findings

- **Construction activity has collapsed 35-40% from peak levels** — dwellings started fell sharply in 2023 and 2024 as interest rates surged, with no recovery in sight for 2025-2026
- **House prices continue rising despite the supply crunch** — the price index is up over 60% since 2015 while construction activity is down 30% from its peak, creating a dangerous divergence
- **The pipeline is emptying fast** — fewer starts today mean fewer completions tomorrow, with under-construction inventory declining quarter after quarter
- **Regional patterns show Oslo's dominance waning** — while Oslo remains the largest construction market, its lead has narrowed as national building slows across all regions
- **Quarterly volatility has increased** — ridgeline plots reveal wider swings in construction activity by region, suggesting developers are more cautious and responsive to short-term market signals

## What comes next

Norway's housing shortage is not resolving itself. With mortgage rates still elevated, construction costs high, and developer confidence shaken, the supply of new homes continues to shrink precisely when demographic growth and urbanization demand more. This mismatch between rising prices and falling construction will likely intensify pressure on policymakers to intervene — whether through subsidies, zoning reform, or other measures to kickstart building activity. The question is not whether something will break, but when.
