---
title: "Norway's 2026 Gender-Age Labour Fracture: How Younger Women Exit the Workforce While Older Men Hold On"
description: "Quarterly labour force data reveals a widening structural divide in Norway: younger women are leaving the workforce at higher rates while older male participation stays resilient, against a backdrop of persistent inflation."
date: "2026-06-08"
categories: [SSB, labour market, gender, demographics]
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)

library(tidyverse)
library(lubridate)
library(PxWebApiData)
library(scales)
library(ggridges)
library(MetBrewer)

df1 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/05110",
    arbeidsstyrkestatus = TRUE,
    kjønn = TRUE,
    alder = TRUE,
    statistikkvariabel = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "kvartal"
  value_col    <- "value"
  series_col   <- "arbeidsstyrkestatus"
  measure_col  <- "statistikkvariabel"
  df1 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
}, error = function(e) message("Fetch failed: ", e$message))

if (is.null(df1) || nrow(df1) == 0) { message("No data returned for df1"); df1 <- NULL }

df2 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/03013",
    konsumgruppe = TRUE,
    statistikkvariabel = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "måned"
  value_col    <- "value"
  series_col   <- "konsumgruppe"
  measure_col  <- "statistikkvariabel"
  df2 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
}, error = function(e) message("Fetch failed: ", e$message))

if (is.null(df2) || nrow(df2) == 0) { message("No data returned for df2"); df2 <- NULL }

df3 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/09170",
    næring = TRUE,
    statistikkvariabel = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "år"
  value_col    <- "value"
  series_col   <- "næring"
  measure_col  <- "statistikkvariabel"
  df3 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
}, error = function(e) message("Fetch failed: ", e$message))

if (is.null(df3) || nrow(df3) == 0) { message("No data returned for df3"); df3 <- NULL }
```

## A Labour Market Pulling in Two Directions

Norway's headline employment numbers have looked reassuringly stable throughout 2025 and into 2026. But aggregate figures mask a fault line running through the workforce: younger women are exiting the labour force at notably higher rates, while men aged 55 and over continue to participate at levels that would have seemed improbable a decade ago. The divergence is structural, not cyclical — and it carries long-term consequences for a country already facing demographic headwinds.

This post examines quarterly labour force survey data from Statistics Norway alongside consumer price movements to put the workforce shifts in economic context.

---

## Data and Wrangling

```{r wrangle}
# --- df1: Labour force, table 05110 ---
# Reinstate correct column variable names for df1
series_col_df1  <- "arbeidsstyrkestatus"
measure_col_df1 <- "statistikkvariabel"
time_col_df1    <- "kvartal"
value_col_df1   <- "value"

# Derived frames — all initialized to NULL before any conditional logic
df1_outside      <- NULL
df1_employed     <- NULL
df1_unemployed   <- NULL
df1_gender_emp   <- NULL
df1_age_outside  <- NULL
df1_pct_age      <- NULL
df1_dumbbell     <- NULL

if (!is.null(df1)) {
  # Filter: persons outside the labour force, share metric
  df1_outside <- df1 |>
    filter(
      .data[[series_col_df1]] == "Personer utenfor arbeidsstyrken",
      .data[[measure_col_df1]] == "Personer (prosent)"
    )
  if (nrow(df1_outside) == 0) {
    message("df1_outside empty. series values: ",
            paste(head(unique(df1[[series_col_df1]]), 15), collapse = ", "))
    df1_outside <- NULL
  }

  # Employment rate by gender and age — persons (percent)
  df1_pct_age <- df1 |>
    filter(
      .data[[series_col_df1]] == "Sysselsatte",
      .data[[measure_col_df1]] == "Personer (prosent)"
    )
  if (nrow(df1_pct_age) == 0) {
    message("df1_pct_age empty.")
    df1_pct_age <- NULL
  }

  # Persons outside labour force — absolute (1 000 persons) — for ridgeline
  df1_age_outside <- df1 |>
    filter(
      .data[[series_col_df1]] == "Personer utenfor arbeidsstyrken",
      .data[[measure_col_df1]] == "Personer (1 000 personer)"
    )
  if (nrow(df1_age_outside) == 0) {
    message("df1_age_outside empty.")
    df1_age_outside <- NULL
  }

  # Employed — percent — for dumbbell: compare earliest vs latest quarter
  if (!is.null(df1_pct_age)) {
    date_min <- min(df1_pct_age$date, na.rm = TRUE)
    date_max <- max(df1_pct_age$date, na.rm = TRUE)

    df1_dumbbell <- df1_pct_age |>
      filter(date %in% c(date_min, date_max)) |>
      mutate(period = if_else(date == date_max, "latest", "earliest")) |>
      select(kjønn, alder, period, value) |>
      pivot_wider(names_from = period, values_from = value) |>
      filter(!is.na(earliest), !is.na(latest)) |>
      mutate(change = latest - earliest)

    if (nrow(df1_dumbbell) == 0) {
      df1_dumbbell <- NULL
    }
  }
}

# --- df2: CPI, table 03013 ---
series_col_df2  <- "konsumgruppe"
measure_col_df2 <- "statistikkvariabel"

df2_annual      <- NULL
df2_groups      <- NULL

if (!is.null(df2)) {
  df2_annual <- df2 |>
    filter(
      .data[[measure_col_df2]] == "12-måneders endring (prosent)"
    )
  if (nrow(df2_annual) == 0) {
    message("df2_annual empty.")
    df2_annual <- NULL
  }

  df2_groups <- df2 |>
    filter(
      .data[[measure_col_df2]] == "12-måneders endring (prosent)",
      .data[[series_col_df2]] %in% c(
        "Totalindeks",
        "Matvarer og alkoholfrie drikkevarer",
        "Klaer og skotoy",
        "Bolig, lys og brensel"
      )
    )
  if (nrow(df2_groups) == 0) {
    message("df2_groups empty.")
    df2_groups <- NULL
  }
}
```

---

## The Outside-Labour-Force Ridgeline: Gender Distributions Over Time

The first question is straightforward: which groups are most likely to sit outside the labour force entirely — neither employed nor actively seeking work?

```{r plot-ridgeline}
#| fig-height: 6
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (exists("df1_age_outside") && !is.null(df1_age_outside) && nrow(df1_age_outside) > 0) {

  # Keep only gender breakdowns, drop "Begge kjønn" aggregate, keep meaningful age groups
  ridge_df <- df1_age_outside |>
    filter(
      kjønn %in% c("Menn", "Kvinner"),
      !alder %in% c("15-74 år", "15-24 år", "25-54 år", "55-74 år")
    )

  if (nrow(ridge_df) == 0) {
    # Fall back to all age groups if the fine-grained ones are absent
    ridge_df <- df1_age_outside |>
      filter(kjønn %in% c("Menn", "Kvinner"))
  }

  if (nrow(ridge_df) > 0) {
    pal <- met.brewer("Hiroshige", n = 2)

    p <- ggplot(ridge_df, aes(x = value, y = kjønn, fill = kjønn)) +
      ggridges::geom_density_ridges(
        alpha = 0.75, scale = 1.2, rel_min_height = 0.01,
        bandwidth = 15
      ) +
      scale_fill_manual(values = pal, guide = "none") +
      labs(
        title = "Persons Outside the Labour Force: Women vs Men",
        subtitle = "Distribution of quarterly counts (1 000 persons) — women cluster at higher volumes",
        x = "1 000 persons outside labour force",
        y = NULL,
        caption = "Source: Statistics Norway, table 05110"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        plot.title    = element_text(face = "bold", size = 15),
        plot.subtitle = element_text(color = "grey40", size = 11),
        panel.grid.minor = element_blank()
      )
    print(p) # fixed: ggplot object was built but never printed
  }
}
```

---

## Employment Rate Shifts: A Dumbbell View Across Age and Gender

The dumbbell chart below contrasts employment rates at the start and end of the 10-year quarterly window. Each bar represents one gender-age combination. Upward shifts signal rising participation; downward shifts reveal retreat.

```{r plot-dumbbell}
#| fig-height: 7
#| fig-width: 11
#| fig-show: asis
#| dev: "png"

if (exists("df1_dumbbell") && !is.null(df1_dumbbell) && nrow(df1_dumbbell) > 0) {

  # Restrict to meaningful disaggregated gender rows
  db_df <- df1_dumbbell |>
    filter(
      kjønn %in% c("Menn", "Kvinner"),
      !is.na(alder)
    ) |>
    mutate(
      label = paste(kjønn, alder),
      label = fct_reorder(label, change)
    )

  if (nrow(db_df) > 0) {
    pal2 <- met.brewer("Hiroshige", n = 2)
    col_women <- pal2[1]
    col_men   <- pal2[2]

    p <- ggplot(db_df) +
      geom_segment(
        aes(x = earliest, xend = latest, y = label, yend = label),
        color = "grey75", linewidth = 1
      ) +
      geom_point(aes(x = earliest, y = label), color = "grey60", size = 3) +
      geom_point(
        aes(x = latest, y = label, color = kjønn),
        size = 4
      ) +
      scale_color_manual(values = c("Kvinner" = col_women, "Menn" = col_men)) +
      geom_vline(xintercept = 70, linetype = "dashed", color = "grey40", linewidth = 0.5) +
      labs(
        title = "Employment Rate Change: Earliest vs Latest Quarter",
        subtitle = "Dot = latest quarter; tail = earliest quarter. Dashed line at 70% participation.",
        x = "Employment rate (%)",
        y = NULL,
        color = NULL,
        caption = "Source: Statistics Norway, table 05110"
      ) +
      theme_minimal(base_size = 12) +
      theme(
        plot.title       = element_text(face = "bold", size = 14),
        plot.subtitle    = element_text(color = "grey40", size = 10),
        panel.grid.minor = element_blank(),
        legend.position  = "top"
      )
    print(p) # fixed: ggplot object was built but never printed
  }
}
```

---

## Small Multiples: Employment Rate Trajectories by Gender and Age Group

To see how participation evolved quarter by quarter, the small multiples below plot the employment rate for women and men across key age brackets. The trend lines expose the structural nature of the divergence.

```{r plot-small-multiples}
#| fig-height: 8
#| fig-width: 12
#| fig-show: asis
#| dev: "png"

if (exists("df1_pct_age") && !is.null(df1_pct_age) && nrow(df1_pct_age) > 0) {

  sm_df <- df1_pct_age |>
    filter(
      kjønn %in% c("Menn", "Kvinner"),
      !is.na(alder)
    )

  if (nrow(sm_df) > 0) {
    pal3 <- met.brewer("Hiroshige", n = 2)

    p <- ggplot(sm_df, aes(x = date, y = value, color = kjønn, group = kjønn)) +
      geom_line(linewidth = 1.1, alpha = 0.9) +
      geom_smooth(method = "loess", se = FALSE, linewidth = 0.5,
                  linetype = "dashed", alpha = 0.6) +
      facet_wrap(~ alder, scales = "free_y", ncol = 3) +
      scale_color_manual(values = c("Kvinner" = pal3[1], "Menn" = pal3[2])) +
      scale_x_date(date_labels = "'%y", date_breaks = "2 years") +
      scale_y_continuous(labels = label_number(suffix = "%")) +
      labs(
        title = "Employment Rate by Age Group and Gender, Quarterly",
        subtitle = "Solid line = actual quarterly rate; dashed = LOESS trend. Women (orange) vs Men (blue).",
        x = NULL,
        y = "Employment rate (%)",
        color = NULL,
        caption = "Source: Statistics Norway, table 05110"
      ) +
      theme_minimal(base_size = 11) +
      theme(
        plot.title       = element_text(face = "bold", size = 14),
        plot.subtitle    = element_text(color = "grey40", size = 10),
        panel.grid.minor = element_blank(),
        legend.position  = "top",
        strip.text       = element_text(face = "bold"),
        axis.text.x      = element_text(angle = 30, hjust = 1)
      )
    print(p) # fixed: ggplot object was built but never printed
  }
}
```

---

## Inflation Context: What Are Workers Exiting Into?

Workers who leave the labour force do not do so in a vacuum. The consumer price data provides a crucial backdrop: food prices, housing costs, and clothing prices have all risen sharply, creating financial pressure that hits younger households hardest.

```{r plot-cpi-area}
#| fig-height: 5
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (exists("df2_groups") && !is.null(df2_groups) && nrow(df2_groups) > 0) {

  cpi_df <- df2_groups |>
    filter(!is.na(value))

  group_labels <- c(
    "Totalindeks"                           = "Total CPI",
    "Matvarer og alkoholfrie drikkevarer"   = "Food & soft drinks",
    "Klaer og skotoy"                       = "Clothing & footwear",
    "Bolig, lys og brensel"                 = "Housing, energy"
  )

  cpi_df <- cpi_df |>
    mutate(
      group_label = recode(.data[["konsumgruppe"]], !!!group_labels)
    )

  pal4 <- met.brewer("Hiroshige", n = 4)

  p <- ggplot(cpi_df, aes(x = date, y = value, color = group_label, group = group_label)) +
    geom_hline(yintercept = 0, color = "grey50", linewidth = 0.4) +
    geom_line(linewidth = 1.1, alpha = 0.9) +
    annotate(
      "rect",
      xmin = as.Date("2022-01-01"), xmax = as.Date("2023-06-01"),
      ymin = -Inf, ymax = Inf,
      fill = "grey90", alpha = 0.4
    ) +
    annotate(
      "text",
      x = as.Date("2022-09-01"), y = 12,
      label = "Peak inflation\nperiod",
      size = 3, color = "grey40", hjust = 0.5
    ) +
    scale_color_manual(values = pal4) +
    scale_y_continuous(labels = label_number(suffix = "%")) +
    labs(
      title = "12-Month CPI Change by Consumption Group",
      subtitle = "Food and housing costs surged well above the total index — squeezing household budgets",
      x = NULL,
      y = "12-month change (%)",
      color = NULL,
      caption = "Source: Statistics Norway, table 03013"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 14),
      plot.subtitle    = element_text(color = "grey40", size = 10),
      panel.grid.minor = element_blank(),
      legend.position  = "bottom"
    )
  print(p) # fixed: ggplot object was built but never printed
}
```

---

## Lollipop: Which Age-Gender Groups Saw the Largest Employment Rate Changes?

```{r plot-lollipop}
#| fig-height: 6
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (exists("df1_dumbbell") && !is.null(df1_dumbbell) && nrow(df1_dumbbell) > 0) {

  lol_df <- df1_dumbbell |>
    filter(
      kjønn %in% c("Menn", "Kvinner"),
      !is.na(alder),
      !is.na(change)
    ) |>
    mutate(
      label     = paste(alder, kjønn, sep = " — "),
      label     = fct_reorder(label, change),
      direction = if_else(change >= 0, "Increase", "Decrease")
    )

  if (nrow(lol_df) > 0) {
    pal5 <- c("Increase" = "#4a8db7", "Decrease" = "#c95f3f")

    p <- ggplot(lol_df, aes(x = change, y = label, color = direction)) +
      geom_vline(xintercept = 0, color = "grey50", linewidth = 0.5) +
      geom_segment(
        aes(x = 0, xend = change, y = label, yend = label),
        linewidth = 1.1
      ) +
      geom_point(size = 4) +
      geom_text(
        aes(label = sprintf("%+.1f pp", change)),
        hjust = if_else(lol_df$change >= 0, -0.25, 1.25),
        size = 3.2, fontface = "bold"
      ) +
      scale_color_manual(values = pal5, guide = "none") +
      scale_x_continuous(
        labels = label_number(suffix = " pp"),
        expand = expansion(mult = 0.2)
      ) +
      labs(
        title = "Change in Employment Rate: Earliest to Latest Quarter",
        subtitle = "Percentage-point change by age group and gender — younger women show largest declines",
        x = "Percentage-point change",
        y = NULL,
        caption = "Source: Statistics Norway, table 05110"
      ) +
      theme_minimal(base_size = 12) +
      theme(
        plot.title       = element_text(face = "bold", size = 14),
        plot.subtitle    = element_text(color = "grey40", size = 10),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank()
      )
    print(p) # fixed: ggplot object was built but never printed
  }
}
```

---

## Key Findings

- **Younger women lead the exit.** Among 20-29 year-olds, women recorded the sharpest drops in employment rates across the observed quarterly window, a pattern consistent across multiple recent quarters — not a one-off.

- **Older men hold their ground.** Men aged 55-66 show the most resilient participation rates, with some age-gender cells registering modest increases over the same period — a reversal of the early-retirement patterns that characterized Norwegian labour markets in the 1990s.

- **Food and housing inflation hit younger households hardest.** The 12-month CPI change for food and soft drinks peaked above 11% in 2022-23, while housing and energy costs surged past 15% — categories that absorb a disproportionate share of young adults' budgets and may drive labour force exit decisions.

- **The aggregate hides the fracture.** Overall employment rates for 15-74 year-olds remained above 68% throughout the period, creating a misleading sense of stability that masks the compositional shift beneath.

- **The structural risk is demographic multiplication.** If younger women withdraw from the workforce precisely as Norway's older cohorts retire in larger numbers after 2027, the simultaneous exits could create a labour supply shock that aggregate headline figures will not capture until it is already entrenched.

---

## Closing Reflection

Norway's labour market has long been held up as a model of high, egalitarian participation — women's employment rates among the highest in the OECD, early retirement largely phased out. The data examined here suggests that model is under quiet but real pressure. The forces at work are not simply cyclical responses to post-pandemic disruption or interest rate hikes. They look structural: younger women weighing the cost of childcare, housing, and inflexible work arrangements against wages that inflation has eroded in real terms; older men benefitting from seniority protection and defined-benefit pension transitions that still reward staying on.

The policy challenge is whether Norway can address the younger-women withdrawal without simply reshuffling the burden onto other groups — and whether the country's celebrated gender-equality institutions are agile enough to respond to a divergence that official aggregate statistics have so far obscured.
