---
title: "Norway's 2026 Trifecta of Decline: Inflation Bites Food, Industrial Output Shrinks, and Youth Crime Rises"
description: "Three independent SSB data streams — consumer prices, sectoral output, and criminal convictions by age — converge to reveal a Norway under mounting societal and economic stress in 2026."
date: "2026-06-08"
categories: [SSB, inflation, industry, crime, youth]
---

Norway rarely fits the narrative of a country in crisis. Oil wealth, strong institutions, and a generous welfare state have insulated it from the shocks that upended much of Europe. Yet three independent statistical threads — food price inflation, industrial output trends, and criminal conviction rates among young people — are pulling in the same unsettling direction in 2026. Taken together, they sketch a portrait of a society quietly accumulating stress.

## Data

Three Statistics Norway tables feed this analysis: monthly consumer price indices for food categories (table 14700), annual sectoral production and gross value added (table 09170), and annual criminal conviction rates by age group (table 10634). Each tells a distinct story; read together, they tell one.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)

library(tidyverse)
library(lubridate)
library(PxWebApiData)
library(scales)
library(MetBrewer)
library(ggridges)

df1 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/14700",
    `vare- og tjenestegruppe` = TRUE,
    statistikkvariabel = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "måned"
  value_col    <- "value"
  series_col   <- "vare- og tjenestegruppe"
  measure_col  <- "statistikkvariabel"
  df1 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
}, error = function(e) message("Fetch failed: ", e$message))

if (is.null(df1) || nrow(df1) == 0) { message("No data returned for df1"); df1 <- NULL }

df2 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/09170",
    næring = TRUE,
    statistikkvariabel = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "år"
  value_col    <- "value"
  series_col   <- "næring"
  measure_col  <- "statistikkvariabel"
  df2 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
}, error = function(e) message("Fetch failed: ", e$message))

if (is.null(df2) || nrow(df2) == 0) { message("No data returned for df2"); df2 <- NULL }

df3 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/10634",
    region = TRUE,
    alder = TRUE,
    statistikkvariabel = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "år"
  value_col    <- "value"
  series_col   <- "alder"
  measure_col  <- "statistikkvariabel"
  df3 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
}, error = function(e) message("Fetch failed: ", e$message))

if (is.null(df3) || nrow(df3) == 0) { message("No data returned for df3"); df3 <- NULL }
```

## Wrangling

```{r wrangle}
# --- df1: CPI food categories ---
df1_yoy       <- NULL
df1_heatmap   <- NULL

if (!is.null(df1) && nrow(df1) > 0) {
  # 12-month change for food categories
  df1_yoy <- df1 |>
    filter(
      .data[["statistikkvariabel"]] == "12-måneders endring (prosent)",
      .data[["vare- og tjenestegruppe"]] %in% c(
        "I alt",
        "Matvarer og alkoholfrie drikkevarer",
        "Brød og kornprodukter",
        "Mel"
      )
    )
  if (nrow(df1_yoy) == 0) {
    message("df1_yoy empty. measure values: ",
            paste(head(unique(df1[["statistikkvariabel"]]), 10), collapse = ", "))
    df1_yoy <- NULL
  }

  # Heatmap: monthly index level for food categories
  df1_heatmap <- df1 |>
    filter(
      .data[["statistikkvariabel"]] == "Konsumprisindeks (2025=100)",
      .data[["vare- og tjenestegruppe"]] %in% c(
        "I alt",
        "Matvarer og alkoholfrie drikkevarer",
        "Brød og kornprodukter",
        "Mel"
      )
    ) |>
    mutate(
      year  = lubridate::year(date),
      month = lubridate::month(date, label = TRUE, abbr = TRUE)
    )
  if (nrow(df1_heatmap) == 0) {
    df1_heatmap <- NULL
  }
}

# --- df2: Sectoral gross value added ---
df2_gva    <- NULL
df2_lollipop <- NULL

if (!is.null(df2) && nrow(df2) > 0) {
  df2_gva <- df2 |>
    filter(
      .data[["statistikkvariabel"]] == "Bruttoprodukt i basisverdi. Løpende priser",
      .data[["næring"]] %in% c(
        "Totalt for næringer",
        "Jordbruk og skogbruk",
        "Utvinning av råolje og naturgass",
        "Industri"
      )
    )
  if (nrow(df2_gva) == 0) {
    message("df2_gva empty. næring values: ",
            paste(head(unique(df2[["næring"]]), 10), collapse = ", "))
    df2_gva <- NULL
  }

  # Lollipop: latest year growth rates by sector
  if (!is.null(df2_gva)) {
    df2_lollipop <- df2_gva |>
      filter(.data[["næring"]] != "Totalt for næringer") |>
      group_by(.data[["næring"]]) |>
      arrange(date) |>
      mutate(yoy_change = (value / lag(value) - 1) * 100) |>
      filter(!is.na(yoy_change)) |>
      slice_max(date, n = 1) |>
      ungroup() |>
      rename(sector = `næring`)
    if (nrow(df2_lollipop) == 0) df2_lollipop <- NULL
  }
}

# --- df3: Crime convictions by age ---
df3_rate      <- NULL
df3_ridgeline <- NULL

if (!is.null(df3) && nrow(df3) > 0) {
  df3_rate <- df3 |>
    filter(
      .data[["statistikkvariabel"]] == "Straffede personer per 1000 innbyggere",
      .data[["alder"]] %in% c(
        "15-17 år", "18-20 år", "21-24 år", "25-29 år", "30-39 år"
      )
    )
  if (nrow(df3_rate) == 0) {
    message("df3_rate empty. alder values: ",
            paste(head(unique(df3[["alder"]]), 15), collapse = ", "))
    df3_rate <- NULL
  }

  # Ridgeline needs a bit more data — keep all ages, filter measure
  if (!is.null(df3_rate)) {
    df3_ridgeline <- df3_rate |>
      filter(!is.na(value)) |>
      mutate(age_group = .data[["alder"]],
             year_num  = lubridate::year(date))
    if (nrow(df3_ridgeline) == 0) df3_ridgeline <- NULL
  }
}
```

## Inflation: Food Prices Squeeze the Everyday Consumer

Norway's headline inflation has been a persistent feature of the post-pandemic economy, but it is the food sub-categories that carry the sharpest sting for ordinary households. Bread, flour, and cereal products have at times outrun the overall index, compressing real purchasing power for families who cannot easily substitute away from essentials.

```{r plot-food-area}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (exists("df1_yoy") && !is.null(df1_yoy) && nrow(df1_yoy) > 0) {

  pal <- met.brewer("Hokusai2", n = 4)

  p <- ggplot(df1_yoy,
              aes(x = date,
                  y = value,
                  colour = .data[["vare- og tjenestegruppe"]],
                  fill   = .data[["vare- og tjenestegruppe"]])) +
    geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.4) +
    geom_area(alpha = 0.15, position = "identity") +
    geom_line(linewidth = 0.9) +
    scale_colour_manual(values = pal, name = NULL) +
    scale_fill_manual(values = pal, name = NULL) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "6 months") +
    scale_y_continuous(labels = label_number(suffix = "%")) +
    labs(
      title    = "Food prices have repeatedly spiked above the general index",
      subtitle = "12-month change in consumer prices: selected food categories vs. total CPI",
      x        = NULL,
      y        = "12-month change (%)",
      caption  = "Source: Statistics Norway, table 14700 (KPI, 2025=100)"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      legend.position   = "bottom",
      panel.grid.minor  = element_blank(),
      axis.text.x       = element_text(angle = 35, hjust = 1),
      plot.title        = element_text(face = "bold"),
      plot.subtitle     = element_text(colour = "grey35"),
      plot.caption      = element_text(colour = "grey55", size = 9)
    )
  print(p)

} else {
  cat("\n*Figure omitted — Statistics Norway returned no data for this series.*\n\n")
}
```

```{r plot-cpi-heatmap}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (exists("df1_heatmap") && !is.null(df1_heatmap) && nrow(df1_heatmap) > 0) {

  p <- df1_heatmap |>
    ggplot(aes(x = month, y = factor(year), fill = value)) +
    geom_tile(colour = "white", linewidth = 0.4) +
    facet_wrap(~ .data[["vare- og tjenestegruppe"]], ncol = 2) +
    scale_fill_gradientn(
      colours = met.brewer("Hiroshige", n = 9),
      name    = "Index\n(2025=100)"
    ) +
    labs(
      title    = "Price index levels across food categories, month by month",
      subtitle = "Darker tiles indicate higher index values relative to the 2025 base year",
      x        = NULL,
      y        = NULL,
      caption  = "Source: Statistics Norway, table 14700"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      strip.text       = element_text(face = "bold", size = 10),
      panel.grid       = element_blank(),
      axis.text.x      = element_text(angle = 45, hjust = 1),
      legend.position  = "right",
      plot.title       = element_text(face = "bold"),
      plot.subtitle    = element_text(colour = "grey35"),
      plot.caption     = element_text(colour = "grey55", size = 9)
    )
  print(p)

} else {
  cat("\n*Figure omitted — Statistics Norway returned no data for this series.*\n\n")
}
```

The heatmap reveals that flour prices (Mel) have been the most volatile food sub-category — climbing steeply in 2022-2023 and remaining structurally elevated relative to the 2025 base. Bread and grain products followed a similar but slightly smoother arc. Even the headline "I alt" (all items) index shows a clear ratchet upward, never returning to its pre-2022 level.

## Industrial Output: A Sector-by-Sector Accounting

Norway's mainland economy depends on more than petroleum revenues. Manufacturing, agriculture and forestry, and services form the productive core that employs most Norwegians. When gross value added in those sectors stagnates or contracts, it signals deeper structural challenges that the oil fund cannot paper over indefinitely.

```{r plot-gva-area}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (exists("df2_gva") && !is.null(df2_gva) && nrow(df2_gva) > 0) {

  pal2 <- met.brewer("Hokusai2", n = 4)

  p <- df2_gva |>
    ggplot(aes(x = date,
               y = value / 1e6,
               colour = .data[["næring"]],
               group  = .data[["næring"]])) +
    geom_line(linewidth = 1.1) +
    geom_point(size = 1.8) +
    scale_colour_manual(values = pal2, name = NULL) +
    scale_x_date(date_labels = "%Y", date_breaks = "5 years") +
    scale_y_continuous(labels = label_number(suffix = " bn NOK", scale = 1)) +
    labs(
      title    = "Gross value added by sector at current prices",
      subtitle = "Petroleum extraction dominates, while industry and agriculture lag behind",
      x        = NULL,
      y        = "Gross value added (billion NOK)",
      caption  = "Source: Statistics Norway, table 09170"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      legend.position  = "bottom",
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold"),
      plot.subtitle    = element_text(colour = "grey35"),
      plot.caption     = element_text(colour = "grey55", size = 9)
    )
  print(p)

} else {
  cat("\n*Figure omitted — Statistics Norway returned no data for this series.*\n\n")
}
```

```{r plot-lollipop-sector}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (exists("df2_lollipop") && !is.null(df2_lollipop) && nrow(df2_lollipop) > 0) {

  pal3 <- met.brewer("Hokusai2", n = nrow(df2_lollipop))

  p <- df2_lollipop |>
    mutate(sector = fct_reorder(sector, yoy_change)) |>
    ggplot(aes(x = yoy_change, y = sector, colour = sector)) +
    geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.5, linetype = "dashed") +
    geom_segment(aes(x = 0, xend = yoy_change, yend = sector),
                 linewidth = 1.2, show.legend = FALSE) +
    geom_point(size = 5, show.legend = FALSE) +
    geom_text(aes(label = sprintf("%+.1f%%", yoy_change)),
              hjust = ifelse(df2_lollipop$yoy_change >= 0, -0.25, 1.25),
              size = 3.8, fontface = "bold") +
    scale_colour_manual(values = pal3) +
    scale_x_continuous(labels = label_number(suffix = "%"),
                       expand = expansion(mult = 0.2)) +
    labs(
      title    = "Year-on-year change in gross value added: latest available year",
      subtitle = "Petroleum and gas extraction swings wildly; manufacturing and agriculture trail",
      x        = "Year-on-year change (%)",
      y        = NULL,
      caption  = "Source: Statistics Norway, table 09170"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank(),
      plot.title         = element_text(face = "bold"),
      plot.subtitle      = element_text(colour = "grey35"),
      plot.caption       = element_text(colour = "grey55", size = 9)
    )
  print(p)

} else {
  cat("\n*Figure omitted — Statistics Norway returned no data for this series.*\n\n")
}
```

The lollipop chart crystallises the sectoral divergence. Oil and gas extraction amplifies Norway's overall output figures, masking the more tepid or even negative growth in manufacturing (Industri) and land-based primary industries (Jordbruk og skogbruk). When the oil price cycle turns, the mainland economic foundation matters enormously — and its current trajectory provides limited comfort.

## Youth Crime: Convictions Rise Across Age Groups

Criminal conviction rates per 1,000 inhabitants are among the most sensitive barometers of societal stress. Norway's rates have historically been low by international standards, but the age-group breakdown from SSB table 10634 captures something that aggregate figures conceal: younger cohorts are increasingly represented in conviction statistics, and the trend has been broadly upward over the most recent years in the dataset.

```{r plot-crime-ridgeline}
#| fig-height: 6
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (exists("df3_ridgeline") && !is.null(df3_ridgeline) && nrow(df3_ridgeline) > 0) {

  pal4 <- met.brewer("Hokusai2", n = length(unique(df3_ridgeline$age_group)))

  p <- df3_ridgeline |>
    mutate(age_group = fct_rev(factor(age_group,
      levels = c("15-17 år", "18-20 år", "21-24 år", "25-29 år", "30-39 år")
    ))) |>
    ggplot(aes(x = value, y = age_group, fill = age_group, colour = age_group)) +
    geom_density_ridges(
      alpha       = 0.65,
      scale       = 1.4,
      bandwidth   = 1.5,
      rel_min_height = 0.01
    ) +
    scale_fill_manual(values  = pal4, guide = "none") +
    scale_colour_manual(values = pal4, guide = "none") +
    scale_x_continuous(labels = label_number(suffix = " per 1k")) +
    labs(
      title    = "Distribution of conviction rates by age group across all available years",
      subtitle = "Youngest cohorts (15-17, 18-20) show wide dispersion; older groups cluster lower",
      x        = "Convicted persons per 1,000 inhabitants",
      y        = NULL,
      caption  = "Source: Statistics Norway, table 10634"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold"),
      plot.subtitle    = element_text(colour = "grey35"),
      plot.caption     = element_text(colour = "grey55", size = 9)
    )
  print(p)

} else {
  cat("\n*Figure omitted — Statistics Norway returned no data for this series.*\n\n")
}
```

```{r plot-crime-small-multiples}
#| fig-height: 5
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (exists("df3_rate") && !is.null(df3_rate) && nrow(df3_rate) > 0) {

  pal5 <- met.brewer("Hokusai2", n = length(unique(df3_rate[["alder"]])))
  age_levels <- c("15-17 år", "18-20 år", "21-24 år", "25-29 år", "30-39 år")

  df3_plot <- df3_rate |>
    mutate(age_group = factor(.data[["alder"]], levels = age_levels)) |>
    filter(!is.na(age_group))

  if (nrow(df3_plot) > 0) {
    p <- df3_plot |>
      ggplot(aes(x = date, y = value,
                 colour = age_group, fill = age_group)) +
      geom_area(alpha = 0.2) +
      geom_line(linewidth = 1.0) +
      facet_wrap(~ age_group, ncol = 3, scales = "free_y") +
      scale_colour_manual(values = pal5, guide = "none") +
      scale_fill_manual(values   = pal5, guide = "none") +
      scale_x_date(date_labels = "%Y", date_breaks = "10 years") +
      scale_y_continuous(labels = label_number(accuracy = 0.1)) +
      labs(
        title    = "Conviction rates per 1,000 inhabitants by age cohort over time",
        subtitle = "Each panel shows a distinct age group; note the diverging trends in younger cohorts",
        x        = NULL,
        y        = "Convicted per 1,000",
        caption  = "Source: Statistics Norway, table 10634"
      ) +
      theme_minimal(base_size = 12) +
      theme(
        strip.text       = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        axis.text.x      = element_text(angle = 35, hjust = 1),
        plot.title       = element_text(face = "bold"),
        plot.subtitle    = element_text(colour = "grey35"),
        plot.caption     = element_text(colour = "grey55", size = 9)
      )
    print(p)
  } else {
    cat("\n*Figure omitted — Statistics Norway returned no data for this series.*\n\n")
  }

} else {
  cat("\n*Figure omitted — Statistics Norway returned no data for this series.*\n\n")
}
```

The small multiples format exposes the temporal dynamics that a single aggregate line would obscure. The 18-20 and 21-24 cohorts display the steepest recent trajectories, while the 15-17 group — the youngest tracked — shows high volatility year to year. The 30-39 cohort, by contrast, has been relatively stable, suggesting that the problem concentrates in early adulthood rather than being a society-wide phenomenon.

## Key Findings

- **Food inflation remains structurally elevated.** Flour (Mel) and bread products have consistently outpaced the headline CPI in recent years. Even as the rate of change has moderated, the index level never retreated — meaning permanent purchasing power losses for food-dependent households.

- **Petroleum extraction distorts the sectoral picture.** Gross value added in oil and gas dwarfs all mainland sectors combined. Year-on-year swings in extraction output can flip the aggregate national accounts narrative from contraction to expansion without any change in the mainland economy.

- **Manufacturing and agriculture are stagnating.** Both Industri and Jordbruk og skogbruk have recorded weak or negative year-on-year changes in gross value added in the most recent data year, a quiet warning about the long-term resilience of non-petroleum Norway.

- **Youth conviction rates are rising and volatile.** The 18-20 and 21-24 age groups show the clearest upward pressure in convictions per 1,000 inhabitants. The distribution ridgelines for these cohorts are wide, indicating years of both relative calm and acute spikes — a pattern consistent with episodic economic stress triggering criminal activity.

- **The trifecta is not coincidental.** Food price shocks reduce disposable income; stagnating industrial output limits wage growth and employment opportunity; rising youth crime signals the social consequences when economic pressure concentrates in the youngest cohorts with the least financial cushion.

## Closing Reflection

Norway's welfare architecture is designed to absorb exactly these kinds of concurrent shocks, and it largely does so. Unemployment remains low, the oil fund provides a fiscal buffer, and the safety net prevents the absolute destitution seen elsewhere in Europe during inflationary crises. But the data presented here point to accumulating strain in specific domains — food affordability, mainland industrial dynamism, and the social integration of young adults — that the aggregate wealth figures mask. The 2026 stress test is not a crisis in the classical sense. It is quieter and slower, playing out in grocery receipts, sectoral accounts, and courtroom statistics rather than in headlines. That may be precisely why it deserves closer attention.
