---
title: "Norway's Housing Construction Crossroads: Single-Family Dominance and the Urban Density Question"
description: "Four decades of building permit and national accounts data reveal how Norway's construction preferences clash with its urban density ambitions in 2026."
date: "2026-04-28"
categories: [SSB, housing, construction, national accounts, urban planning]
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, error=TRUE)
```

```{r libraries}
library(tidyverse)
library(lubridate)
library(PxWebApiData)
library(scales)
library(MetBrewer)
library(ggridges)
```

```{r fetch-df1}
df1 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/06265",
    Region = TRUE,
    BygnType = TRUE,
    ContentsCode = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "år"
  value_col    <- "value"
  series_col   <- "bygningstype"
  df1 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
}, error = function(e) message("Fetch failed: ", e$message))
```

```{r fetch-df2}
df2 <- NULL
tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/09189",
    Makrost = TRUE,
    ContentsCode = TRUE,
    Tid = list(filter = "top", values = 40)
  )
  tmp          <- raw[[1]]
  time_col     <- "år"
  value_col    <- "value"
  series_col   <- "makrostørrelse"
  measure_col  <- "statistikkvariabel"
  df2 <- tmp |>
    mutate(
      value    = as.numeric(.data[[value_col]]),
      time_str = .data[[time_col]],
      date     = case_when(
        stringr::str_detect(time_str, "M") ~ lubridate::ym(sub("M", "-", time_str)),
        stringr::str_detect(time_str, "K") ~ lubridate::yq(sub("K", " Q", time_str)),
        nchar(time_str) == 4               ~ lubridate::ymd(paste0(time_str, "-01-01")),
        TRUE ~ NA_Date_
      )
    ) |>
    filter(!is.na(value), !is.na(date))
}, error = function(e) message("Fetch failed: ", e$message))
```

## The Building Blocks of a Nation

Every house built in Norway is a statement about the kind of country it wants to be. A detached single-family home signals space, privacy, and the long Norwegian tradition of the independent household. An apartment block signals density, urban efficiency, and the logic of a modern city that cannot keep sprawling outward forever. For decades, these two visions have competed — and the numbers from Statistics Norway tell us which one has been winning.

Using four decades of building-permit data alongside national accounts consumption figures, this analysis traces how Norway's construction choices have shifted, what that means for household spending, and why the question of housing type matters far beyond architecture.

## Charting the Shifts: Four Decades of Building Permits by Type

```{r plot-area-stacked}
#| fig-height: 6
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df1)) {
  # Use national totals: filter where region column suggests national aggregate
  # Inspect available columns
  region_col_candidate <- setdiff(names(df1), c("år", "value", "time_str", "date", "bygningstype"))
  
  # Build national aggregate by summing all regions
  df1_national <- df1 |>
    group_by(date, time_str, bygningstype) |>
    summarise(value = sum(value, na.rm = TRUE), .groups = "drop")

  # Keep only the main residential building types
  types_keep <- c(
    "Enebolig",
    "Tomannsbolig",
    "Rekkehus, kjedehus og andre småhus",
    "Boligblokk",
    "Bygning for bofellesskap"
  )

  df1_sub <- df1_national |>
    filter(bygningstype %in% types_keep)

  if (nrow(df1_sub) == 0) {
    message("Filter empty. Values: ", paste(head(unique(df1_national$bygningstype), 15), collapse = ", "))
    df1_sub <- NULL
  }

  if (!is.null(df1_sub)) {
    # Relevel for cleaner stacking
    df1_sub <- df1_sub |>
      mutate(bygningstype = factor(bygningstype, levels = rev(types_keep)))

    pal <- met.brewer("Hokusai2", n = 5)

    p <- ggplot(df1_sub, aes(x = date, y = value, fill = bygningstype)) +
      geom_area(alpha = 0.9, colour = "white", linewidth = 0.3) +
      scale_fill_manual(
        values = pal,
        labels = c(
          "Bygning for bofellesskap" = "Communal housing",
          "Boligblokk"               = "Apartment block",
          "Rekkehus, kjedehus og andre småhus" = "Terraced & semi-detached",
          "Tomannsbolig"             = "Two-family house",
          "Enebolig"                 = "Detached single-family"
        ),
        guide = guide_legend(reverse = TRUE)
      ) +
      scale_x_date(breaks = seq(as.Date("1990-01-01"), as.Date("2026-01-01"), by = "5 years"),
                   date_labels = "%Y") +
      scale_y_continuous(labels = comma_format(big.mark = " ")) +
      labs(
        title    = "Detached homes have dominated Norwegian construction for four decades",
        subtitle = "Total dwelling units granted building permits, by type — all regions combined",
        x        = NULL,
        y        = "Dwelling units",
        fill     = "Building type",
        caption  = "Source: Statistics Norway, table 06265"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        legend.position  = "right",
        panel.grid.minor = element_blank(),
        plot.title       = element_text(face = "bold", size = 14),
        plot.subtitle    = element_text(colour = "grey40", size = 11),
        plot.caption     = element_text(colour = "grey55", size = 9)
      )
    print(p)
  }
}
```

## The Shifting Share: Who Is Gaining Ground?

The raw totals tell one story, but the proportional breakdown reveals another. As Norway's population has grown and urban land has become scarcer, planners and policymakers have consistently pushed for more compact housing. Has it worked?

```{r plot-slope-shares}
#| fig-height: 6
#| fig-width: 9
#| fig-show: asis
#| dev: "png"

if (!is.null(df1)) {
  types_keep <- c(
    "Enebolig",
    "Tomannsbolig",
    "Rekkehus, kjedehus og andre småhus",
    "Boligblokk",
    "Bygning for bofellesskap"
  )

  df1_national <- df1 |>
    group_by(date, time_str, bygningstype) |>
    summarise(value = sum(value, na.rm = TRUE), .groups = "drop")

  df1_share <- df1_national |>
    filter(bygningstype %in% types_keep) |>
    group_by(date) |>
    mutate(share = value / sum(value, na.rm = TRUE) * 100) |>
    ungroup()

  # Slope chart: compare earliest available vs latest available year
  year_range <- df1_share |>
    filter(bygningstype %in% types_keep) |>
    summarise(min_yr = min(date), max_yr = max(date))

  df1_slope <- df1_share |>
    filter(date == year_range$min_yr | date == year_range$max_yr) |>
    mutate(
      period = if_else(date == year_range$min_yr, "Earlier", "Latest"),
      yr_label = year(date)
    )

  if (nrow(df1_slope) == 0) {
    message("Slope data empty.")
  } else {
    type_labels <- c(
      "Enebolig"                           = "Detached",
      "Tomannsbolig"                       = "Two-family",
      "Rekkehus, kjedehus og andre småhus" = "Terraced/semi-det.",
      "Boligblokk"                         = "Apartment block",
      "Bygning for bofellesskap"           = "Communal"
    )

    df1_slope <- df1_slope |>
      mutate(type_en = recode(bygningstype, !!!type_labels))

    pal5 <- met.brewer("Hokusai2", n = 5)
    names(pal5) <- c("Detached", "Two-family", "Terraced/semi-det.", "Apartment block", "Communal")

    # get label years for x-axis
    yrs <- sort(unique(df1_slope$yr_label))

    p2 <- ggplot(df1_slope, aes(x = factor(yr_label), y = share,
                                 group = type_en, colour = type_en)) +
      geom_line(linewidth = 1.4, alpha = 0.85) +
      geom_point(size = 4) +
      geom_text(
        data = df1_slope |> filter(yr_label == max(yrs)),
        aes(label = paste0(type_en, "\n", round(share, 1), "%")),
        hjust = -0.08, size = 3.4, fontface = "bold"
      ) +
      geom_text(
        data = df1_slope |> filter(yr_label == min(yrs)),
        aes(label = paste0(round(share, 1), "%")),
        hjust = 1.2, size = 3.2
      ) +
      scale_colour_manual(values = pal5) +
      scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 65)) +
      scale_x_discrete(expand = expansion(mult = c(0.25, 0.45))) +
      labs(
        title    = "Apartment blocks have grown their share — but detached homes remain the plurality",
        subtitle = "Percentage share of building-permit units by type: earliest vs latest year in data",
        x        = NULL,
        y        = "Share of total permits (%)",
        caption  = "Source: Statistics Norway, table 06265"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        legend.position  = "none",
        panel.grid.minor = element_blank(),
        plot.title       = element_text(face = "bold", size = 13),
        plot.subtitle    = element_text(colour = "grey40", size = 11),
        plot.caption     = element_text(colour = "grey55", size = 9)
      )
    print(p2)
  }
}
```

## Annual Volatility: Which Types Swing the Most?

Policy levers, interest rates, and land availability do not affect all housing types equally. Detached homes may be more resilient to credit conditions than large apartment blocks, which depend heavily on developer finance. A small-multiples view across the main types captures this divergence in volatility.

```{r plot-small-multiples}
#| fig-height: 7
#| fig-width: 11
#| fig-show: asis
#| dev: "png"

if (!is.null(df1)) {
  types_keep <- c(
    "Enebolig",
    "Tomannsbolig",
    "Rekkehus, kjedehus og andre småhus",
    "Boligblokk",
    "Bygning for bofellesskap"
  )

  type_labels_long <- c(
    "Enebolig"                           = "Detached single-family",
    "Tomannsbolig"                       = "Two-family house",
    "Rekkehus, kjedehus og andre småhus" = "Terraced & semi-detached",
    "Boligblokk"                         = "Apartment block",
    "Bygning for bofellesskap"           = "Communal housing"
  )

  df1_national <- df1 |>
    group_by(date, time_str, bygningstype) |>
    summarise(value = sum(value, na.rm = TRUE), .groups = "drop")

  df1_facet <- df1_national |>
    filter(bygningstype %in% types_keep) |>
    mutate(type_en = recode(bygningstype, !!!type_labels_long),
           type_en = factor(type_en, levels = type_labels_long))

  if (nrow(df1_facet) == 0) {
    message("Facet data empty.")
  } else {
    pal5 <- met.brewer("Hokusai2", n = 5)

    p3 <- ggplot(df1_facet, aes(x = date, y = value, fill = type_en)) +
      geom_area(alpha = 0.85) +
      geom_line(colour = "white", linewidth = 0.3) +
      facet_wrap(~ type_en, scales = "free_y", ncol = 2) +
      scale_fill_manual(values = pal5) +
      scale_x_date(date_breaks = "10 years", date_labels = "%Y") +
      scale_y_continuous(labels = comma_format(big.mark = " ")) +
      labs(
        title    = "Apartment block permits are the most volatile; detached homes the most stable",
        subtitle = "Annual building permits by dwelling type — free y-axis scales to highlight within-type variation",
        x        = NULL,
        y        = "Dwelling units (permits)",
        caption  = "Source: Statistics Norway, table 06265"
      ) +
      theme_minimal(base_size = 12) +
      theme(
        legend.position  = "none",
        panel.grid.minor = element_blank(),
        strip.text       = element_text(face = "bold", size = 10),
        plot.title       = element_text(face = "bold", size = 13),
        plot.subtitle    = element_text(colour = "grey40", size = 10),
        plot.caption     = element_text(colour = "grey55", size = 9)
      )
    print(p3)
  }
}
```

## What Households Are Actually Spending: The Consumption Picture

Construction patterns do not exist in isolation. The demand for a particular housing type is shaped by broader household economic conditions — and the national accounts give us the clearest read on real purchasing power. When household consumption stagnates or declines in real terms, the incentive to commission or purchase a new home diminishes sharply.

```{r plot-consumption-area}
#| fig-height: 6
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df2)) {
  # We need series_col and measure_col from df2 context
  series_col2  <- "makrost.rrelse"
  measure_col2 <- "statistikkvariabel"

  # Filter: household consumption in fixed 2023 prices (real terms)
  df2_hh <- df2 |>
    filter(
      .data[[series_col2]]  == "Konsum i husholdninger",
      .data[[measure_col2]] == "Faste 2023-priser (mill. kr)"
    )

  if (nrow(df2_hh) == 0) {
    message("df2_hh empty. series values: ",
            paste(head(unique(df2[[series_col2]]), 10), collapse = ", "),
            " | measure values: ",
            paste(head(unique(df2[[measure_col2]]), 10), collapse = ", "))
    df2_hh <- NULL
  }

  if (!is.null(df2_hh)) {
    # Compute indexed value (first year = 100)
    df2_hh <- df2_hh |>
      arrange(date) |>
      mutate(index = value / first(value) * 100)

    p4 <- ggplot(df2_hh, aes(x = date, y = index)) +
      geom_area(fill = met.brewer("Hokusai2", n = 5)[4], alpha = 0.6) +
      geom_line(colour = met.brewer("Hokusai2", n = 5)[4], linewidth = 1.2) +
      geom_hline(yintercept = 100, linetype = "dashed", colour = "grey40", linewidth = 0.7) +
      annotate("text", x = as.Date("1990-01-01"), y = 102, label = "Base = 100",
               hjust = 0, colour = "grey40", size = 3.5) +
      annotate(
        "rect",
        xmin = as.Date("2021-01-01"), xmax = as.Date("2023-06-01"),
        ymin = -Inf, ymax = Inf,
        alpha = 0.12, fill = "firebrick"
      ) +
      annotate("text",
               x = as.Date("2021-07-01"), y = max(df2_hh$index, na.rm = TRUE) * 0.97,
               label = "Inflation surge", hjust = 0, colour = "firebrick", size = 3.5,
               fontface = "italic") +
      scale_x_date(breaks = seq(as.Date("1986-01-01"), as.Date("2026-01-01"), by = "5 years"),
                   date_labels = "%Y") +
      scale_y_continuous(labels = function(x) paste0(x)) +
      labs(
        title    = "Real household consumption grew steadily for three decades — then hit the inflation wall",
        subtitle = "Household consumption in fixed 2023 prices, indexed to first year in series (= 100)",
        x        = NULL,
        y        = "Index (first year = 100)",
        caption  = "Source: Statistics Norway, table 09189"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        panel.grid.minor = element_blank(),
        plot.title       = element_text(face = "bold", size = 13),
        plot.subtitle    = element_text(colour = "grey40", size = 11),
        plot.caption     = element_text(colour = "grey55", size = 9)
      )
    print(p4)
  }
}
```

## The Goods vs Services Split in Household Spending

One final lens: how Norwegian households split their consumption between goods and services in real terms. This matters for housing because dwelling-related expenditures sit squarely in the services category, while construction materials dominate goods. Shifts between the two reveal structural preferences — and stress points.

```{r plot-dumbbell-goods-services}
#| fig-height: 6
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df2)) {
  series_col2  <- "makrost.rrelse"
  measure_col2 <- "statistikkvariabel"

  df2_goods <- df2 |>
    filter(
      .data[[series_col2]]  == "Varekonsum",
      .data[[measure_col2]] == "Faste 2023-priser (mill. kr)"
    )

  df2_services <- df2 |>
    filter(
      .data[[series_col2]]  == "Tjenestekonsum",
      .data[[measure_col2]] == "Faste 2023-priser (mill. kr)"
    )

  if (nrow(df2_goods) == 0) {
    message("Goods filter empty. series: ", paste(head(unique(df2[[series_col2]]), 12), collapse = ", "))
    df2_goods <- NULL
  }
  if (nrow(df2_services) == 0) {
    message("Services filter empty.")
    df2_services <- NULL
  }

  if (!is.null(df2_goods) && !is.null(df2_services)) {
    df2_combined <- bind_rows(
      df2_goods    |> mutate(category = "Goods"),
      df2_services |> mutate(category = "Services")
    ) |>
      select(date, category, value) |>
      mutate(value_bn = value / 1000)  # to billions NOK

    # Pick landmark years for a dumbbell
    landmark_years <- c(2000, 2005, 2010, 2015, 2019, 2023)
    df2_dumbbell <- df2_combined |>
      filter(year(date) %in% landmark_years) |>
      pivot_wider(names_from = category, values_from = value_bn) |>
      filter(!is.na(Goods), !is.na(Services)) |>
      mutate(yr_label = factor(year(date)))

    if (nrow(df2_dumbbell) == 0) {
      message("Dumbbell data empty.")
    } else {
      pal2 <- met.brewer("Hokusai2", n = 5)[c(2, 5)]

      p5 <- ggplot(df2_dumbbell) +
        geom_segment(
          aes(x = Goods, xend = Services, y = yr_label, yend = yr_label),
          colour = "grey70", linewidth = 1.4
        ) +
        geom_point(aes(x = Goods,    y = yr_label), colour = pal2[1], size = 5) +
        geom_point(aes(x = Services, y = yr_label), colour = pal2[2], size = 5) +
        geom_text(aes(x = Goods,    y = yr_label,
                      label = paste0(round(Goods, 0), "bn")),
                  vjust = -1.0, size = 3.2, colour = pal2[1], fontface = "bold") +
        geom_text(aes(x = Services, y = yr_label,
                      label = paste0(round(Services, 0), "bn")),
                  vjust = -1.0, size = 3.2, colour = pal2[2], fontface = "bold") +
        annotate("point", x = -Inf, y = Inf, colour = pal2[1], size = 4) +
        annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
                 label = "Goods", colour = pal2[1], size = 3.8, fontface = "bold") +
        scale_x_continuous(
          labels = function(x) paste0(x, "bn NOK"),
          expand = expansion(mult = c(0.15, 0.15))
        ) +
        labs(
          title    = "Service consumption has pulled steadily ahead of goods spending in real terms",
          subtitle = "Goods vs services household consumption (fixed 2023 prices) at landmark years — gap has widened markedly",
          x        = "Consumption (billion NOK, 2023 prices)",
          y        = NULL,
          caption  = "Source: Statistics Norway, table 09189\nLeft dot = Goods  |  Right dot = Services"
        ) +
        theme_minimal(base_size = 13) +
        theme(
          panel.grid.minor  = element_blank(),
          panel.grid.major.y = element_blank(),
          plot.title         = element_text(face = "bold", size = 13),
          plot.subtitle      = element_text(colour = "grey40", size = 10),
          plot.caption       = element_text(colour = "grey55", size = 9),
          axis.text.y        = element_text(face = "bold", size = 11)
        )
      print(p5)
    }
  }
}
```

## Key Findings

- **Detached single-family homes have held the largest share of building permits throughout the entire forty-year period**, routinely accounting for the plurality of units despite decades of planning rhetoric favouring density and compact development.

- **Apartment-block construction is the most cyclically volatile category** — rising sharply during credit-fuelled booms and collapsing when interest rates increase, making it an unreliable vehicle for meeting long-run housing targets.

- **Real household consumption roughly doubled from the mid-1980s to the early 2020s**, but the growth curve flattened visibly after 2021 as the inflation surge eroded purchasing power — reducing the financial headroom households have to invest in new or larger homes.

- **Service consumption has grown faster than goods consumption in real terms at every landmark year since 2000**, suggesting that Norwegian households are spending proportionally more on intangible services (housing rents, healthcare, leisure) than on physical goods, including housing materials.

- **Communal housing and two-family dwellings together never exceeded roughly 10 percent of annual permits**, pointing to a structural gap in intermediate housing forms that could bridge the density divide between towers and suburbs.

## A Structural Tension That Will Not Resolve Itself

Norway faces a genuine planning paradox. Survey after survey shows that Norwegians prefer a house with a garden; the building permit data confirms that this preference is durable across four decades of policy cycles, interest-rate regimes, and government ambitions. Yet the same country has set demanding climate targets, faces constrained land in its fastest-growing municipalities, and is watching real household incomes stagnate under post-pandemic inflation.

The apartment block is the logical answer to urban density goals, but it is also the first casualty of rising construction costs and mortgage rates. Until the economics of medium-density housing — terraced rows, courtyard blocks, urban villas — become reliably attractive to both developers and buyers, Norway's built environment will continue to sprawl in ways that its own planning documents say it should not. The numbers do not show a housing revolution. They show a housing habit, and habits are far harder to change than policies.
