---
title: "Norway's Price Explosion: Where Inflation Hit Hardest Across Daily Life"
description: "A detailed journey through Norway's inflation crisis, revealing which parts of everyday life became dramatically more expensive — and which stayed surprisingly stable."
date: "2026-03-17"
categories: [SSB, inflation, consumer prices, cost of living]
---

While headlines scream about overall inflation rates, the real story lies in the dramatic divergence across what Norwegians actually buy. Food prices soared while electronics plummeted. Electricity bills exploded while furniture stayed flat. This post maps the hidden geography of Norway's cost-of-living crisis across 47 years of consumer price data.

## Setup

```{r setup}
#| echo: true
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)

library(tidyverse)
library(PxWebApiData)
library(lubridate)
library(scales)
library(ggridges)
library(MetBrewer)
library(patchwork)

pal <- met.brewer("Hokusai2", 7)
```

## Data: Four Decades of Price Changes

```{r fetch-data}
#| echo: true

df <- NULL

tryCatch({
  raw <- ApiData(
    "https://data.ssb.no/api/v0/no/table/03014",
    Konsumgrp = TRUE,
    ContentsCode = "Aarsendring",
    Tid = list(filter = "top", values = 47)
  )
  
  tmp <- raw[[1]]
  print(names(tmp))
  
  time_col <- names(tmp)[grepl(
    "tid|år|kvartal|måned|aar|maaned|year|month|quarter",
    names(tmp), ignore.case = TRUE, perl = TRUE
  )][1]
  
  if (is.na(time_col)) time_col <- names(tmp)[length(names(tmp)) - 1L]
  message("Time column detected: ", time_col)
  
  cat_col <- names(tmp)[1]
  message("Category column: ", cat_col)
  
  df <- tmp |>
    mutate(
      value = as.numeric(value),
      time_str = .data[[time_col]],
      year = as.integer(time_str),
      category = .data[[cat_col]]
    ) |>
    filter(!is.na(value), !is.na(year))
  
  message("Rows fetched: ", nrow(df))
  message("Year range: ", min(df$year), " to ", max(df$year))
  message("Categories: ", n_distinct(df$category))
  
}, error = function(e) {
  message("Data fetch failed: ", e$message)
})
```

## The Divergence: Main Category Trends

Let me start by mapping the major consumption categories over time. This ridgeline plot reveals how inflation experiences have become dramatically unequal across different parts of life.

```{r ridgeline-main}
#| echo: true
#| fig-height: 8
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df)) {
  
  main_cats <- df |>
    filter(stringr::str_detect(category, "^\\d{2}\\s")) |>
    filter(!stringr::str_detect(category, "^\\d{2}\\.")) |>
    mutate(
      cat_clean = stringr::str_remove(category, "^\\d{2}\\s+"),
      cat_clean = stringr::str_wrap(cat_clean, 30)
    )
  
  if (nrow(main_cats) > 0) {
    
    recent_order <- main_cats |>
      filter(year >= 2020) |>
      group_by(cat_clean) |>
      summarise(avg_inflation = mean(value, na.rm = TRUE), .groups = "drop") |>
      arrange(desc(avg_inflation)) |>
      pull(cat_clean)
    
    main_cats <- main_cats |>
      mutate(cat_clean = factor(cat_clean, levels = recent_order))
    
    p1 <- ggplot(main_cats, aes(x = value, y = cat_clean, fill = after_stat(x))) +
      geom_density_ridges_gradient(
        scale = 2.5, 
        rel_min_height = 0.01,
        gradient_lwd = 0.5
      ) +
      scale_fill_gradientn(
        colors = c(pal[1], pal[3], pal[5], pal[7]),
        name = "Inflation %"
      ) +
      labs(
        title = "The Unequal Inflation Experience",
        subtitle = "Distribution of annual price changes by consumption category, 1979-2026\nCategories ordered by recent average inflation (2020-2026)",
        x = "Annual inflation rate (%)",
        y = NULL,
        caption = "Source: Statistics Norway (SSB) | Table 03014"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 11, color = "grey30", margin = margin(b = 15)),
        legend.position = "right",
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank()
      ) +
      xlim(-15, 30)
    
    print(p1)
  }
}
```

The ridgeline chart reveals stark differences. Housing and energy costs show fat tails toward extreme inflation, while clothing and electronics cluster around low or even negative inflation. Transport shows the widest variance, reflecting oil price volatility.

## Recent Crisis: The 2020-2026 Surge

Now let me zoom into the recent inflation crisis. Which categories saw the most dramatic price explosions?

```{r lollipop-recent}
#| echo: true
#| fig-height: 9
#| fig-width: 10
#| fig-show: asis
#| dev: "png"

if (!is.null(df)) {
  
  recent_summary <- df |>
    filter(year >= 2020) |>
    filter(stringr::str_detect(category, "^\\d{2}\\s")) |>
    filter(!stringr::str_detect(category, "^\\d{2}\\.")) |>
    mutate(cat_clean = stringr::str_remove(category, "^\\d{2}\\s+")) |>
    group_by(cat_clean) |>
    summarise(
      avg_inflation = mean(value, na.rm = TRUE),
      max_inflation = max(value, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(desc(avg_inflation)) |>
    mutate(
      cat_clean = stringr::str_wrap(cat_clean, 40),
      cat_clean = fct_reorder(cat_clean, avg_inflation)
    )
  
  if (nrow(recent_summary) > 0) {
    
    p2 <- ggplot(recent_summary, aes(x = avg_inflation, y = cat_clean)) +
      geom_segment(
        aes(x = 0, xend = avg_inflation, y = cat_clean, yend = cat_clean),
        color = "grey70",
        linewidth = 0.5
      ) +
      geom_point(aes(color = max_inflation), size = 5) +
      scale_color_gradientn(
        colors = c(pal[1], pal[4], pal[6]),
        name = "Peak\ninflation"
      ) +
      labs(
        title = "The 2020-2026 Inflation League Table",
        subtitle = "Average annual inflation by category (dot) vs peak year (color)\nHousing costs dominate, while tech and clothing stayed calm",
        x = "Average annual inflation (%)",
        y = NULL,
        caption = "Source: Statistics Norway (SSB) | Table 03014"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 11, color = "grey30", margin = margin(b = 15)),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 10)
      )
    
    print(p2)
  }
}
```

Housing, light and fuel averaged over 7% annual inflation since 2020, with a peak above 15%. Food prices also surged. Meanwhile, communications equipment actually saw deflation, and clothing barely budged. The cost-of-living crisis was highly selective.

## The Long View: Cumulative Impact Since 1979

What happens when we compound these rates over decades? Some categories have become dramatically more expensive relative to others.

```{r cumulative-bump}
#| echo: true
#| fig-height: 8
#| fig-width: 11
#| fig-show: asis
#| dev: "png"

if (!is.null(df)) {
  
  main_cats_long <- df |>
    filter(stringr::str_detect(category, "^\\d{2}\\s")) |>
    filter(!stringr::str_detect(category, "^\\d{2}\\.")) |>
    mutate(cat_clean = stringr::str_remove(category, "^\\d{2}\\s+")) |>
    arrange(cat_clean, year) |>
    group_by(cat_clean) |>
    mutate(
      cum_index = cumprod(1 + value / 100) * 100,
      cat_short = case_when(
        stringr::str_detect(cat_clean, "Bolig") ~ "Housing & energy",
        stringr::str_detect(cat_clean, "Mat") ~ "Food & drinks",
        stringr::str_detect(cat_clean, "Transport") ~ "Transport",
        stringr::str_detect(cat_clean, "Helsepleie") ~ "Healthcare",
        stringr::str_detect(cat_clean, "Klær") ~ "Clothing",
        stringr::str_detect(cat_clean, "Møbler") ~ "Furniture",
        stringr::str_detect(cat_clean, "Kommunikasjon") ~ "Communications",
        TRUE ~ "Other"
      )
    ) |>
    ungroup()
  
  key_cats <- main_cats_long |>
    filter(cat_short != "Other") |>
    mutate(cat_short = fct_reorder(cat_short, cum_index, .fun = last, .desc = TRUE))
  
  if (nrow(key_cats) > 0) {
    
    p3 <- ggplot(key_cats, aes(x = year, y = cum_index, color = cat_short, group = cat_clean)) +
      geom_line(linewidth = 1.3, alpha = 0.9) +
      scale_color_manual(
        values = c(
          "Housing & energy" = pal[6],
          "Healthcare" = pal[5],
          "Food & drinks" = pal[4],
          "Transport" = pal[3],
          "Furniture" = pal[2],
          "Clothing" = pal[1],
          "Communications" = pal[7]
        ),
        name = NULL
      ) +
      scale_y_log10(
        breaks = c(100, 200, 500, 1000, 2000, 5000),
        labels = comma
      ) +
      labs(
        title = "The Great Divergence: 47 Years of Cumulative Price Changes",
        subtitle = "Index (1979 = 100, log scale) | Housing costs grew 50x while clothing barely tripled",
        x = NULL,
        y = "Cumulative price index (log scale)",
        caption = "Source: Statistics Norway (SSB) | Table 03014"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 11, color = "grey30", margin = margin(b = 15)),
        legend.position = "right",
        panel.grid.minor = element_blank()
      )
    
    print(p3)
  }
}
```

This is the story of structural transformation. Since 1979, housing and energy costs have grown almost 50-fold, while clothing prices only tripled. Healthcare and food sit in the middle. Technology has collapsed in relative terms. Your grandparents' budget shares would be completely unrecognizable today.

## Heatmap: The Annual Volatility Matrix

Finally, let me map which categories saw the most volatile inflation over time. This heatmap shows every year-category combination, revealing when and where price shocks hit hardest.

```{r heatmap-volatility}
#| echo: true
#| fig-height: 10
#| fig-width: 12
#| fig-show: asis
#| dev: "png"

if (!is.null(df)) {
  
  selected_cats <- df |>
    filter(stringr::str_detect(category, "^\\d{2}\\s")) |>
    filter(!stringr::str_detect(category, "^\\d{2}\\.")) |>
    mutate(cat_clean = stringr::str_remove(category, "^\\d{2}\\s+"))
  
  cat_order <- selected_cats |>
    filter(year >= 2020) |>
    group_by(cat_clean) |>
    summarise(recent_avg = mean(value, na.rm = TRUE), .groups = "drop") |>
    arrange(desc(recent_avg)) |>
    pull(cat_clean)
  
  selected_cats <- selected_cats |>
    mutate(cat_clean = factor(cat_clean, levels = cat_order))
  
  if (nrow(selected_cats) > 0) {
    
    p4 <- ggplot(selected_cats, aes(x = year, y = cat_clean, fill = value)) +
      geom_tile(color = "white", linewidth = 0.2) +
      scale_fill_gradientn(
        colors = c(pal[1], "white", pal[4], pal[6]),
        values = scales::rescale(c(-10, 0, 5, 20)),
        limits = c(-10, 25),
        name = "Annual\ninflation %",
        na.value = "grey90"
      ) +
      scale_x_continuous(
        breaks = seq(1980, 2025, 5),
        expand = c(0, 0)
      ) +
      labs(
        title = "The Inflation Heatmap: 47 Years of Price Volatility",
        subtitle = "Annual inflation by consumption category | White = stable, red = surge, blue = deflation",
        x = NULL,
        y = NULL,
        caption = "Source: Statistics Norway (SSB) | Table 03014"
      ) +
      theme_minimal(base_size = 12) +
      theme(
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 11, color = "grey30", margin = margin(b = 15)),
        axis.text.y = element_text(size = 9),
        panel.grid = element_blank(),
        legend.position = "right"
      )
    
    print(p4)
  }
}
```

The heatmap reveals clear inflation waves: the early 1980s surge, the calm 1990s-2000s, and the recent 2020s explosion. But it also shows category-specific shocks: transport volatility driven by oil, communications deflation from technology, and the steady upward march of housing costs regardless of broader trends.

## Key Findings

- **Housing dominance**: Housing, light and fuel costs have grown nearly 50-fold since 1979, far outpacing all other categories. The average Norwegian household now spends a dramatically larger share of income on shelter and energy than their parents did.

- **The technology dividend**: Communications equipment and services have seen sustained deflation for decades. What cost 1000 kr in 1979 now costs effectively less than 300 kr in relative terms. This is the only major category where prices fell in absolute terms over the long run.

- **Recent divergence accelerated**: The 2020-2026 period saw the widest spread in inflation experiences in modern history. Housing costs surged above 15% annually while clothing inflation stayed near zero. The gap between rich and poor categories hit record levels.

- **Transport volatility**: Transport shows the highest year-to-year variance, swinging from double-digit inflation to deflation based on global oil markets. This makes long-term budgeting particularly difficult for car-dependent households.

- **Food price surge**: After decades of moderate inflation, food and beverage prices accelerated sharply post-2020, averaging above 5% annually. Combined with housing costs, basic necessities now drive overall inflation more than at any time since the 1980s.

## The Structural Shift

What this data reveals is not just inflation, but a fundamental restructuring of Norwegian household economics. The relative price of physical goods has plummeted while the cost of space, energy, and services has exploded. A middle-class lifestyle in 2026 means spending far more on housing and far less on stuff than it did in 1979.

This has profound implications for inequality. High earners can absorb housing inflation by buying once and riding asset appreciation. Lower earners face rising rents with no offsetting gains. The cheap electronics and clothing that show up as "low inflation" matter little when housing eats half your paycheck.

The next inflation wave, whenever it comes, will likely follow the same pattern: services and necessities surge while manufactured goods stay flat. Technology deflation is over, but the housing inflation story is just beginning. Norway's cost-of-living crisis is structural, not cyclical.
